---
title: "CF-Control distance analysis, 2"
author: "Rebecca L. Knoll"
date: " last edit `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: show
    number_sections: yes
    smart: no
    toc: yes
    df_print: paged
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries and run functions script}
library(microViz)

pacman::p_load(tidyverse, phyloseq, magrittr, janitor, microbiome, knitr, lubridate, naniar, readxl, ggplot2, ggpubr)

library(dendextend)
library(vegan)

ps_full <- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/ps_IMProveCF_patientsAndControls_Run1-23_18102023.rds")

source("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/IMMProveCF/functions_full.R")

visit_sum_palette <- c("black", "#C3BC3FFF", "#6388B4FF", "#BB7693FF", "#55AD89FF", "#8176AA")

dom_palette <- c(Streptococcus="#69b3a2", Staphylococcus = "#8175AA", Fusobacterium="#6FB899FF", Prevotella_7="#31A1B3FF", Rothia="#027B8EFF", Pseudomonas="#EE6AA7", Eikenella="#94D0C0FF", Haemophilus="#CCB22BFF", Achromobacter="#9F8F12FF", Gemella= "#97CFD0" , Moraxella= "#6FB899", `missing sample` = "#CFCFCF", `Burkholderia-Caballeronia-Paraburkholderia` = "#E5C494")
```

```{r, prepare data}

ps_full_relab <- transform_sample_counts(ps_full, function(x) x/sum(x)*100)

#subset materials
ps_full_throat <- subset_samples(ps_full_relab, material== "Throat")
ps_full_stool <- subset_samples(ps_full_relab, material== "Stool")

ps_full_throat_glom <- tax_glom(ps_full_throat, taxrank = "Genus")
ps_full_stool_glom <- tax_glom(ps_full_stool, taxrank = "Genus")
```

# Analysis on most abundant taxa per patient in throat
```{r}
#Get top taxa per patient
#find.top.taxa2 is sourced from functions.R
top.throat<- find.top.taxa2(ps_full_throat_glom, "Genus",1)
top.throat$Species<- NULL

rslt <- top.throat[, "taxa"]
dd <- matrix(unlist(rslt), nrow=1)
colnames(dd) <- rownames(top.throat)
top.throat <- t(dd)

top.throat_df <- data.frame(x1 = row.names(top.throat), top.throat)%>%
  mutate(dominantGenus = top.throat)
top.throat_df$top.throat<- NULL

# add clinical data to p

##Add dominant Genus to ps_full_throat_glom sample data
ps_full_throat_glom_j <- microViz::ps_join(ps_full_throat_glom, top.throat_df, by = "x1")

##Add dominant Genus to ps_full_throat sample data
ps_full_throat_j <- microViz::ps_join(ps_full_throat, top.throat_df, by = "x1")

# Control's dominant taxa
summary(as_factor(sample_data(ps_full_throat_j)$dominantGenus[sample_data(ps_full_throat_j)$project=="IMMPIMMP"]))

# Patients's dominant taxa
summary(as_factor(sample_data(ps_full_throat_j)$dominantGenus[sample_data(ps_full_throat_j)$project!="IMMPIMMP"]))

 ### plot principal component analysis by dominant Genus
PCA_all_visit_throat_dominantGenus <- plot_ordination(ps_full_throat_j, ordinate(ps_full_throat_j, "MDS"), color = "dominantGenus", shape="visit_sum")

xlabels <- c("0","3","6-12","15-18","21-24","Control")

PCA_all_visit_throat_dominantGenus+
  geom_line(aes(group=id.x), color="grey")+
  geom_point(size = 4)+
  ggtitle("Variation by dominant Genus")+
  scale_color_manual(values = dom_palette)+
  scale_shape_manual(values = c(15, 16, 17, 18, 12, 23),labels = xlabels)+
  theme_bw()+
  theme(text=element_text(size=24), legend.position = "right", legend.text = element_text(size=16), legend.title = element_text(size=18))+
  guides(color = guide_legend(title = "dominant genus"))+
   guides(shape = guide_legend(title = "months from treatment start"))

### plot principal component analysis by visit_sum
PCA_all_visit_throat_dominantGenus <- plot_ordination(ps_full_throat_j, ordinate(ps_full_throat_j, "MDS"), color = "visit_sum", shape="project")

project_label <- c("Control","CF")

xlabels <- c("0", "3", "6-12", "15-18", "21-24", "Control")

p1 <- PCA_all_visit_throat_dominantGenus +
  geom_point(size = 4, alpha = 0.7) +
  scale_color_manual(values = visit_sum_palette, labels = xlabels) +
  scale_shape_manual(values = c(15, 16), labels = project_label) +
  theme_classic() +
  theme(text = element_text(size = 18), legend.position = "none", legend.text = element_text(size = 16), legend.title = element_blank()) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 2))+
  stat_ellipse()
p1

p1_leg <- PCA_all_visit_throat_dominantGenus +
  geom_point(size = 4, alpha = 0.7) +
  scale_color_manual(values = visit_sum_palette, labels = xlabels) +
  scale_shape_manual(values = c(15, 16), labels = project_label) +
  theme_classic() +
  theme(text = element_text(size = 16), legend.position = "bottom", legend.text = element_text(size = 16)) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 6),shape = guide_legend(title = "", title.position = "left", ncol = 2))+
  stat_ellipse()
p1_leg

#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/beta_throat_controls.png", width = 10, height = 6)

```
# create density plots for axis 1 and 2
```{r, fig.width=4, fig.height=4}
# Perform PCoA
ord_obj <- ordinate(ps_full_throat_j, "MDS")

# Extract coordinates from PCoA
pcoa_coords <- ord_obj$vectors

# Create a data frame for ggplot
df <- data.frame(x1 = pcoa_coords[, 1], x2 = pcoa_coords[, 2], visit_sum = sample_data(ps_full_throat_j)$visit_sum,age_y = sample_data(ps_full_throat_j)$age_y,sex= sample_data(ps_full_throat_j)$sex,id.x= sample_data(ps_full_throat_j)$id.x)

# Create a named vector for custom labels
custom_labels <- c("1" = "0", 
                   "2" = "3", 
                   "3-5" = "6-12", 
                   "6-7" = "15-18", 
                   "8-10" = "21-24", 
                   "Control" = "Control")
# Create density plots on the x and y axes
plot_x_th <- ggplot(df, aes(x = x1, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  #clean_theme()+
  scale_x_continuous(breaks = c(-0.3, 0, 0.3), labels = c("-0.3", "0", "0.3"))+
  scale_fill_manual(values = visit_sum_palette, labels = xlabels)+
  theme(legend.position = "none", text = element_text(size = 18), legend.title = element_blank())

# Define the common y-axis limits
y_axis_limits <- c(0, 2.5)  # Adjust these values as needed

# Density plot with fixed y-axis range
axis1_den <- ggplot(df, aes(x = x1, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.8) +
  theme_classic() +
  scale_x_continuous(breaks = c(-0.3, 0, 0.3), labels = c("-0.3", "0", "0.3")) +
    scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  theme(
    legend.position = "none",
    text = element_text(size = 18),
    legend.title = element_blank(),
    strip.text.y = element_text(angle = 0)  # Rotate the facet labels
  ) +
  facet_grid(rows = vars(visit_sum), labeller =  as_labeller(custom_labels)) +
  ylim(y_axis_limits)+
  labs( x = "Axis 1")
axis1_den

# Boxplot comparing x between the 5 groups
axis1_bp <- ggplot(df, aes(x = as.factor(visit_sum), y = x1, fill = as.factor(visit_sum))) +
  geom_boxplot(alpha=0.8) +
  theme_classic() +
  scale_x_discrete(labels = xlabels) +
  scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  theme(
    legend.position = "none",
    text = element_text(size = 18),
    legend.title = element_blank()
  ) +
  labs(x = "Months from tt start", y = "Axis 1")
axis1_bp

plot_y_th <- ggplot(df, aes(x = x2, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  #clean_theme()+
  xlim(c(-0.6, 0.4))+
  scale_fill_manual(values = visit_sum_palette, labels = xlabels)+
  theme(legend.position = "none",text = element_text(size = 18), legend.title = element_blank())+
  coord_flip()

# Define the common y-axis limits
y_axis_limits <- c(0, 3.5)  # Adjust these values as needed

# Density plot with fixed y-axis range
axis2_den <- ggplot(df, aes(x = x2, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.8) +
  theme_classic() +
  scale_x_continuous(breaks = c(-0.6, 0, 0.4), labels = c("-0.6", "0", "0.4")) +
  scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  theme(
    legend.position = "none",
    text = element_text(size = 18),
    legend.title = element_blank(),
    strip.text.y = element_text(angle = 0)  # Rotate the facet labels
  ) +
  facet_grid(rows = vars(visit_sum), labeller =  as_labeller(custom_labels)) +
  ylim(y_axis_limits)+
  labs(x = "Axis 2")
axis2_den

# Boxplot comparing x between the 5 groups
axis2_bp <- ggplot(df, aes(x = as.factor(visit_sum), y = x2, fill = as.factor(visit_sum))) +
  geom_boxplot(alpha=0.8) +
  theme_classic() +
  scale_x_discrete(labels = xlabels) +
  scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  theme(
    legend.position = "none",
    text = element_text(size = 18),
    legend.title = element_blank(),
  ) +
  labs(x = "Months from tt start", y = "Axis 2")
axis2_bp
# Extract the legend. Returns a gtable
leg <- get_legend(p1_leg)

# Convert to a ggplot and print
legend <- as_ggplot(leg)

# Arranging the plot

axis <- ggarrange(axis1_den, axis2_den, axis1_bp, axis2_bp, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(1, 1),
          common.legend = F)
axis

throat_density <- ggarrange(plot_x_th, NULL, p1, plot_y_th, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(2, 1), heights = c(1, 2),
          common.legend = F)
throat_density
```

```{r}
# Load required libraries
library(tidyverse)
library(lmerTest)
library(ggpubr)

# Define the custom labels
xlabels <- c("0", "3", "6-12", "15-18", "21-24", "Control")

# Convert visit_sum to factor with specified levels
df<- df%>%
  mutate(visit_sum = factor(visit_sum, levels = c("1", "2", "3-5", "6-7", "8-10", "Control")))

# Define the function for linear model and boxplot
lm_corrected_pvalues_boxplot <- function(x, ylabel) {
  
  print(ylabel) # x is the diversity parameter I am interested in
  
  # Fit the linear mixed-effects model
  lm <- summary(lmerTest::lmer(x ~ visit_sum + age_y + sex + (1|id.x), data = df))
  
  # Extract coefficients and compute FDR
  coefs <- data.frame(coef(lm))
  fdr <- p.adjust(coefs$Pr...t.., method = "fdr", n = nrow(coefs))
  
  # Prepare the statistics table
  lm_stats <- bind_cols(coefs, fdr) %>%
    mutate(p = Pr...t..) %>%
    mutate(fdr = ...6) %>%
    select(-c(Pr...t.., ...6)) %>%
    rownames_to_column() %>%
    mutate(Months_after_ETI_start = rowname) %>%
    mutate(Months_after_ETI_start = case_when(
      Months_after_ETI_start == "(Intercept)" ~ "Baseline (Intercept)",
      Months_after_ETI_start == "visit_sum2" ~ "3",
      Months_after_ETI_start == "visit_sum3-5" ~ "6-12",
      Months_after_ETI_start == "visit_sum6-7" ~ "15-18",
      Months_after_ETI_start == "visit_sum8-10" ~ "21-24",
      Months_after_ETI_start == "visit_sumControl" ~ "Control",
      Months_after_ETI_start == "age_y" ~ "age_y",
      Months_after_ETI_start == "sex2" ~ "sex"
    )) %>%
    mutate(fdr_star = case_when(
      fdr <= 0.001 ~ "***",
      fdr <= 0.01 ~ "**",
      fdr <= 0.05 ~ "*",
      fdr <= 0.1 ~ ".",
      fdr > 0.1 ~ "ns"
    )) %>%
    mutate(p_star = case_when(
      p <= 0.001 ~ "***",
      p <= 0.01 ~ "**",
      p <= 0.05 ~ "*",
      p <= 0.1 ~ ".",
      p > 0.1 ~ "ns"
    )) %>%
    select(-rowname) %>%
    select(Months_after_ETI_start, Estimate, Std..Error, df, t.value, p, fdr, p_star, fdr_star) %>%
    mutate(p = round(p, 5)) %>%
    mutate(fdr = round(fdr, 5))
  
  lm_stats_print <- lm_stats %>% 
    mutate(alpha_diversity = ylabel) %>% 
    select(alpha_diversity, everything())

  # Adjust table for significance asterisks on the plot
  group1 <- c("(Intercept)", "1", "1", "1", "1", "1", "1", "1")
  
  lm_stats <- lm_stats %>%
    bind_cols(group1) %>%
    mutate(group1 = ...10) %>%
    mutate(group2 = case_when(
      Months_after_ETI_start == "Baseline (Intercept)" ~ "(Intercept)",
      Months_after_ETI_start == "3" ~ "2",
      Months_after_ETI_start == "6-12" ~ "3-5",
      Months_after_ETI_start == "15-18" ~ "6-7",
      Months_after_ETI_start == "21-24" ~ "8-10",
      Months_after_ETI_start == "Control" ~ "Control"
    )) %>%
    filter(group2 != "(Intercept)")

  # Create the boxplot
  boxplot <- ggplot(df, aes(visit_sum, x)) +
    geom_boxplot(aes(fill = visit_sum), outlier.shape = NA, alpha = 0.65) +
    geom_jitter(width = 0.1) +
    scale_fill_manual(values = visit_sum_palette) +
    theme_classic() +
    ylab(ylabel) +
    xlab("Months from ETI treatment start") +
    scale_x_discrete(labels = xlabels) +
    theme(
      text = element_text(size = 16), 
      legend.position = "none"
    ) +
    guides(fill = 'none') +
    stat_pvalue_manual(lm_stats, label = "fdr_star", y.position = (max(x, na.rm = TRUE) - 0.07 * max(x, na.rm = TRUE)), step.increase = 0.05)
  
  return(list(boxplot = boxplot, lm_stats = lm_stats_print))
}

# Run function on  data
axis1_throat <- lm_corrected_pvalues_boxplot(df$x1, "Axis 1")
axis1_throat_lm_stats <- axis1_throat$lm_stats # Access lm_stats
p3 <- axis1_throat$boxplot # Access boxplot

# Print the boxplot
print(p3)

axis2_throat <- lm_corrected_pvalues_boxplot(df$x2, "Axis 2")
axis2_throat_lm_stats <- axis2_throat$lm_stats # Access lm_stats
p4 <- axis2_throat$boxplot # Access boxplot

# Print the boxplot
print(p4)

ggarrange(p3,p4)

# run statistics on controls vs all CF samples,
# in order to make this feasible in a linear model I have to revert the order of visit_sum
#
df$visit_sum_reorder <- factor(df$visit_sum, levels = c("Control", "1", "2", "3-5", "6-7", "8-10"))

lm_corrected_pvalues_boxplot_reorder <- function(x, ylabel) {
  
  print(ylabel) # x is the diversity parameter I am interested in
  lm <- summary(lmerTest::lmer(x~ visit_sum_reorder + age_y + sex + (1|id.x), data=df))
  
  coefs <- data.frame(coef(lm))
  fdr <- p.adjust(coefs$Pr...t.., method = "fdr", n = nrow(coefs))
  
  lm_stats <- bind_cols(coefs, fdr)%>%
    mutate(p = Pr...t..)%>%
    mutate(fdr = ...6)%>%
    select(-c(Pr...t.., ...6))%>%
    rownames_to_column()%>%
    mutate(Months_after_ETI_start=rowname)%>%
    mutate(Months_after_ETI_start= case_when(Months_after_ETI_start=="(Intercept)"~"Control (Intercept)", Months_after_ETI_start=="visit_sum_reorder2"~"3", Months_after_ETI_start=="visit_sum_reorder3-5"~"6-12",Months_after_ETI_start=="visit_sum_reorder6-7"~"15-18",Months_after_ETI_start=="visit_sum_reorder8-10"~ "21-24", Months_after_ETI_start=="visit_sum_reorder1"~"Baseline", Months_after_ETI_start=="age_y"~ "age_y", Months_after_ETI_start=="sex2"~ "sex"))%>%
    mutate(fdr_star = case_when(fdr<=0.001 ~ "***", fdr<=0.01 ~ "**", fdr<=0.05 ~ "*",fdr<=0.1 ~ ".",fdr>=0.1 ~ "ns"))%>%
    mutate(p_star = case_when(p<=0.001 ~ "***", p<=0.01 ~ "**", p<=0.05 ~ "*",p<=0.1 ~ ".", p>=0.1 ~ "ns"))%>%
    select(-rowname)%>%
    select(Months_after_ETI_start, Estimate, Std..Error, df, t.value, p, fdr, p_star, fdr_star)%>%
    mutate(p=round(p,5))%>%
    mutate(fdr=round(fdr, 5))
  
  lm_stats_print <- lm_stats %>% 
    mutate(alpha_diversity = ylabel) %>% 
    select(alpha_diversity, everything())

# to make the significant asterixes printable on the plot I have to adjust the table a little
group1 <- c("(Intercept)","Control","Control","Control","Control","Control","Control","Control")

lm_stats <-  lm_stats%>%
    bind_cols(group1)%>%
    mutate(group1=...10)%>%
    mutate(group2= case_when(Months_after_ETI_start=="Control (Intercept)"~"(Intercept)",
                             Months_after_ETI_start=="3"~"2",
                             Months_after_ETI_start=="Baseline"~"1", Months_after_ETI_start=="6-12" ~ "3-5", Months_after_ETI_start=="15-18"~"6-7",Months_after_ETI_start=="21-24"~"8-10"))%>%
    filter(group2!="(Intercept)")

  boxplot <- df%>%
    ggplot(aes(visit_sum, x))+
    geom_boxplot(aes(fill=visit_sum), outlier.shape = NA, alpha=0.65) +
    geom_jitter(width = 0.1)+
    scale_fill_manual(values=visit_sum_palette)+
    theme_classic()+
    ylab(ylabel)+
    xlab("Months from ETI treatment start")+
    scale_x_discrete(labels= xlabels)+
    theme(text=element_text(size = 16), legend.position = "none")+
    guides(fill='none') +
    stat_pvalue_manual(lm_stats, label = "fdr_star",y.position = (min(x)-0.07*min(x)), step.increase = 0.05)
  
    return(list(boxplot = boxplot, lm_stats = lm_stats_print, lm_stats_sig=lm_stats))
}

#Axis 1
axis1.2_throat <- lm_corrected_pvalues_boxplot_reorder(df$x1, "Axis 1")
axis1_throat_lm_stats_reorder <- axis1.2_throat$lm_stats # Access lm_stats
axis1_throat_reorder <- axis1.2_throat$boxplot # Access boxplot
axis1_throat_reorder

axi1_wC_thr <- axis1_throat$boxplot +
  geom_signif(
    y_position = c(-0.6, -0.55, -0.5, -0.45),  
    xmin = c(2, 3, 4, 5), 
    xmax = c(6, 6, 6, 6),
    annotation = c(".", "*", "ns", "*"), 
    tip_length = -0.03, 
    vjust = 1,
    size = 0.35
  )
axi1_wC_thr

#Axis 2
axis2.2_thr <- lm_corrected_pvalues_boxplot_reorder(df$x2, "Axis 2")
axis2_thr_lm_stats_reorder <- axis2.2_thr$lm_stats # Access lm_stats
axis2_thr_reorder <- axis2.2_thr$boxplot # Access boxplot
axis2_thr_reorder
axi2_wC_thr <- axis2_throat$boxplot +
  geom_signif(
    y_position = c(-0.6, -0.55, -0.5, -0.45), 
    xmin = c(2, 3, 4, 5), 
    xmax = c(6, 6, 6, 6),
    annotation = c("ns", "ns", "ns", "ns"), 
    tip_length = -0.03, 
    vjust = 1,
    size = 0.35
  )
axi2_wC_thr

# Arranging the plot

axis_throat <- ggarrange(axis1_den, axis2_den, axi1_wC_thr, axi2_wC_thr, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(1, 1),
          common.legend = F)
axis_throat
```

# PERMANOVA in throat
```{r calculate bray curtis distance for all throat samples and run PERMANOVA}

# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_throat_j,
                             method="bray", weighted=F)
#extract metadata
metadata<- as(sample_data(ps_full_throat_j),"data.frame")

### unstratified

adonis2(formula = BC_dist ~ visit_sum, data = metadata, permutations = 999, by = "margin")

adonis2(formula = BC_dist ~ visit_sum + sex + age_y, data = metadata, permutations = 999, by = "margin")

adonis2(formula = BC_dist ~ visit_sum + sex + age_y +id.x, data = metadata, permutations = 999, by = "margin") # sex and age are not significant when including id.x as a explantory variable

adonis2(formula = BC_dist ~ visit_sum +id.x, data = metadata, permutations = 999, by = "margin") # sex and age are not significant when including id.x as a explantory variable

### stratified PERMANOVA per id
adonis2(formula = BC_dist ~ visit_sum , data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

adonis2(formula = BC_dist ~ visit_sum + project , data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

### stratified PERMANOVA per project (health vs CF)
adonis2(formula = BC_dist ~ visit_sum + id.x, data = metadata, permutations = 999, by = "margin", strata = metadata$project)

### unstratified PERMANOVA health vs CF
adonis2(formula = BC_dist ~ project, data = metadata, permutations = 999, by = "margin")
adonis2(formula = BC_dist ~ project + id.x, data = metadata, permutations = 999, by = "margin")
adonis2(formula = BC_dist ~ project + visit_sum, data = metadata, permutations = 999, by = "margin")
### stratified PERMANOVA health vs CF
adonis2(formula = BC_dist ~ project + visit_sum + sex + age_y, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

adonis2(formula = BC_dist ~ project + id.x, data = metadata, permutations = 999, by = "margin", strata = metadata$visit_sum)

adonis2(formula = BC_dist ~ visit_sum + id.x, data = metadata, permutations = 999, by = "margin", strata = metadata$project)
```
# PERMANOVA per time point comparison

```{r}

# check baseline vs follow-up samples
metadata<- metadata %>% 
  mutate(ETI_yn = as_factor(case_when(visit_sum=="1"~"no", visit_sum=="Control"~"Control", T~"yes")))

summary(metadata$ETI_yn)

### plot principal component analysis by visit_sum
PCA_all_visit_throat_dominantGenus <- plot_ordination(ps_full_throat_j, ordinate(ps_full_throat_j, "MDS"), color = "visit_sum", shape="project")

project_label <- c("Control","CF")

labels <- c("Baseline", "ETI", "ETI", "ETI", "ETI", "Control")

bfu_palette <- c("black", "#55AD89FF","#55AD89FF","#55AD89FF","#55AD89FF","#8176AA" )

PCA_all_visit_throat_dominantGenus +
  geom_point(size = 4, alpha = 0.7) +
  scale_color_manual(values = bfu_palette, labels = xlabels) +
  scale_shape_manual(values = c(15, 16), labels = project_label) +
  theme_classic() +
  theme(text = element_text(size = 18), legend.position = "none", legend.text = element_text(size = 16), legend.title = element_blank()) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 2))+
  stat_ellipse()

### stratified PERMANOVA control vs baseline vs follow-up
adonis2(formula = BC_dist ~ ETI_yn, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

# Repeat PERMANOVA for separate time points, paired with baseline

metadata_v1 <- metadata %>%
  filter(visit_sum %in% c("1", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v1$x_sample_id

# Filter the distance matrix to only include relevant rows and cols
BC_df <- as.matrix(BC_dist)
BC_df <- as.data.frame(BC_df)

# Filter rows based on sample IDs
df_v1 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v1 <- df_v1[, colnames(df_v1) %in% sample_ids]

# Convert to matrix
BC_v1 <- as.matrix(df_v1)

### unstratified
adonis2(formula = BC_v1 ~ visit_sum, data = metadata_v1, permutations = 999, by = "margin")
# as there are no repeated samples no need for stratification
adonis2(formula = BC_v1 ~ project, data = metadata_v1, permutations = 999, by = "margin")

# now Control vs all follow-ups together (without baseline)

metadata_fu <- metadata %>%
  filter(ETI_yn %in% c("yes", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_fu$x_sample_id
# Filter the distance matrix to only include relevant rows and cols
BC_df <- as.matrix(BC_dist)
BC_df <- as.data.frame(BC_df)

# Filter rows based on sample IDs
df_fu <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_fu <- df_fu[, colnames(df_fu) %in% sample_ids]

# Convert to matrix
BC_fu <- as.matrix(df_fu)

### stratified
adonis2(formula = BC_fu ~ visit_sum, data = metadata_fu, permutations = 999, by = "margin", strata = metadata_fu$id.x)
adonis2(formula = BC_fu ~ ETI_yn, data = metadata_fu, permutations = 999, by = "margin", strata = metadata_fu$id.x)

# now Control vs follow-ups separately
metadata_v2 <- metadata %>%
  filter(visit_sum %in% c("2", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v2$x_sample_id

# Filter rows based on sample IDs
df_v2 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v2 <- df_v2[, colnames(df_v2) %in% sample_ids]

# Convert to matrix
BC_v2 <- as.matrix(df_v2)

### unstratified
adonis2(formula = BC_v2 ~ visit_sum, data = metadata_v2, permutations = 999, by = "margin")
# as there are no repeated samples no need for stratification - it also gives an error
adonis2(formula = BC_v2 ~ project, data = metadata_v2, permutations = 999, by = "margin")

metadata_v3 <- metadata %>%
  filter(visit_sum %in% c("3-5", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v3$x_sample_id

# Filter rows based on sample IDs
df_v3 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v3 <- df_v3[, colnames(df_v3) %in% sample_ids]

# Convert to matrix
BC_v3 <- as.matrix(df_v3)

### unstratified
adonis2(formula = BC_v3 ~ visit_sum, data = metadata_v3, permutations = 999, by = "margin")
adonis2(formula = BC_v3 ~ project, data = metadata_v3, permutations = 999, by = "margin")
# as there are repeated samples, I will block id to verify this observation
adonis2(formula = BC_v3 ~ visit_sum, data = metadata_v3, permutations = 999, by = "margin", strata = metadata_v3$id.x)

metadata_v6 <- metadata %>%
  filter(visit_sum %in% c("6-7", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v6$x_sample_id

# Filter rows based on sample IDs
df_v6 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v6 <- df_v6[, colnames(df_v6) %in% sample_ids]

# Convert to matrix
BC_v6 <- as.matrix(df_v6)

### unstratified
adonis2(formula = BC_v6 ~ visit_sum, data = metadata_v6, permutations = 999, by = "margin")
adonis2(formula = BC_v6 ~ project, data = metadata_v6, permutations = 999, by = "margin")
# as there are repeated samples, I will block id to verify this observation
adonis2(formula = BC_v6 ~ visit_sum, data = metadata_v6, permutations = 999, by = "margin", strata = metadata_v6$id.x)

metadata_v8 <- metadata %>%
  filter(visit_sum %in% c("8-10", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v8$x_sample_id

# Filter rows based on sample IDs
df_v8 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v8 <- df_v8[, colnames(df_v8) %in% sample_ids]

# Convert to matrix
BC_v8 <- as.matrix(df_v8)

### unstratified
adonis2(formula = BC_v8 ~ visit_sum, data = metadata_v8, permutations = 999, by = "margin")
adonis2(formula = BC_v8 ~ project, data = metadata_v8, permutations = 999, by = "margin")
# as there are repeated samples, I will block id to verify this observation
adonis2(formula = BC_v8 ~ visit_sum, data = metadata_v8, permutations = 999, by = "margin", strata = metadata_v8$id.x)
```
# PERMANOVA on CF-health comparisons alone

```{r, eval=FALSE}

# I reduce the distance matrix to only have CF samples in the rows, and only healthy samples in the columns, so that only CF to healthy comparisons are retained

# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_throat_j,
                             method="bray", weighted=F)

#extract metadata
metadata<- as(sample_data(ps_full_throat_j),"data.frame")

# Use grep to find columns that contain "IMMP" in their names
cols_to_keep <- metadata %>% 
  filter(project=="IMMPIMMP") %>% 
  select(x_sample_id)
cols_to_keep <- cols_to_keep$x_sample_id

rows_to_keep <- metadata %>% 
  filter(project=="IMP") %>% 
  select(x_sample_id)
rows_to_keep <- rows_to_keep$x_sample_id
# Convert the distance object to a matrix
BC_matrix <- as.matrix(BC_dist)

# Subset the matrix to keep only those columns that are from controls
BC_matrix_f <- BC_matrix[, cols_to_keep]
BC_matrix_f <- BC_matrix[rows_to_keep, ]

# no keep in the metadata only CF samples
metadata <- metadata %>% 
  filter(project=="IMP")

### unstratified
adonis2(formula = BC_matrix_f ~ visit_sum, data = metadata, permutations = 999, by = "margin")

adonis2(formula = BC_matrix_f ~ visit_sum + id.x, data = metadata, permutations = 999, by = "margin")

### stratified PERMANOVA per id
adonis2(formula = BC_matrix_f ~ visit_sum, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

# check baseline vs follow-up samples
metadata<- metadata %>% 
  mutate(ETI_yn = case_when(visit_sum=="1"~FALSE, T~TRUE))

### unstratified
adonis2(formula = BC_matrix_f ~ ETI_yn, data = metadata, permutations = 999, by = "margin")

adonis2(formula = BC_matrix_f ~ ETI_yn + id.x, data = metadata, permutations = 999, by = "margin")

### stratified PERMANOVA per id
adonis2(formula = BC_matrix_f ~ ETI_yn, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

# Repeat PERMANOVA for separate time points, paired with baseline

md_v1v2 <- metadata %>%
  filter(visit_sum %in% c("1", "2"))

# Filter the distance matrix to only include relevant rows
BC_matrix_f <- as.data.frame(BC_matrix_f)
BC_v1v2 <- BC_matrix_f %>% 
  filter(rownames(BC_matrix_f) %in% c(md_v1v2$x_sample_id))

# Convert to matrix
BC_v1v2 <- as.matrix(BC_v1v2)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v1v2 ~ visit_sum , data = md_v1v2, permutations = 999, by = "margin")
adonis2(formula = BC_v1v2 ~ visit_sum + id.x, data = md_v1v2, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id
adonis2(formula = BC_v1v2 ~ visit_sum, data = md_v1v2, permutations = 999, by = "margin", strata = md_v1v2$id.x)

# 6-12 months + baseline
md_v1v35 <- metadata %>%
  filter(visit_sum %in% c("1", "3-5"))

# Filter the distance matrix to only include relevant rows
BC_v35 <- BC_matrix_f %>% 
  filter(rownames(BC_matrix_f) %in% c(md_v1v35$x_sample_id))

# Convert to matrix
BC_v35 <- as.matrix(BC_v35)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v35 ~ visit_sum , data = md_v1v35, permutations = 999, by = "margin")
adonis2(formula = BC_v35 ~ visit_sum + id.x, data = md_v1v35, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id
adonis2(formula = BC_v35 ~ visit_sum, data = md_v1v35, permutations = 999, by = "margin", strata = md_v1v35$id.x)
```

# compare CF and control BC-distances for throat
```{r compare CF and control BC-distances}
# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_throat_j,
                             method="bray", weighted=F)
BC_dist.throat <- as.matrix(BC_dist)
#BC_dist.throat[lower.tri(BC_dist.throat)] <- 0  # this is needed to remove the 0 latter, to keep only single pairs
BC_dist.throat_df <- reshape2::melt(BC_dist.throat)

tmp1 <- BC_dist.throat_df%>%
  filter(grepl("IMMP", Var1)) %>% #this keeps only control samples in Var 1
  filter(!grepl("IMMP", Var2))#%>%  #this filters control samples from Var 2, so that in the end only distances between COntrols and all CF samples are kept
  #filter(value!=0)#this removes the same sample distances, and the one from set 0 from above to keep a single pair

# merge with metadata to have visit_sums for CF samples
#extract metadata
metadata<- as(sample_data(ps_full_throat_j),"data.frame")
tmp1 <- tmp1 %>% 
  left_join(metadata, by=c("Var2"="x_sample_id")) %>% 
  mutate(comparison=paste(Var1, Var2, sep = "_")) %>% 
  distinct(comparison, .keep_all = T)

summary(tmp1$visit_sum)

my_comp <- list(c("1","2"), c("1","3-5"), c("1","6-7"), c("1","8-10"))

# Calculate median and IQR
summary_stats <- tmp1 %>%
  group_by(visit_sum) %>%
  summarize(
    median = median(value),
    lower = quantile(value, 0.25),
    upper = quantile(value, 0.75)
  )

tmp1 %>% 
  ggplot(aes(visit_sum, value, fill=visit_sum))+
  geom_boxplot(outlier.shape = NA, alpha=0.7)+
  #geom_point()+
  theme_classic()+
  scale_fill_manual(values = visit_sum_palette)+
  ylab("BC distance between CF and Controls")+
  theme(legend.position = "none",text=element_text(size=20))+ #
  scale_x_discrete(labels= xlabels)+
  xlab("")+
  stat_compare_means(comparisons = my_comp, method = "wilcox.test", label = "p.adj")

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/BC_throat_controls.png", width = 8, height = 6)

summary(lmerTest::lmer(value~visit_sum + (1|Var1)+ (1|id.x), data = tmp1))

# below I plot the BC distance per patient and each control, with controls being on the x-axis, one can observe the consistency of BC distance per patient across controls, with visit_sum giving a distinct signal in each patient
tmp1 %>% 
  ggplot(aes(Var1, value, fill=visit_sum))+
  #geom_boxplot(outlier.shape = NA)+
  geom_point(aes(color=visit_sum), alpha=0.5)+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  ylab("BC distance between CF and Controls")+
  theme(legend.position = "none",text=element_text(size=20))+ #
  #scale_x_discrete(labels= xlabels)+
  xlab("")+
  facet_wrap(~id.x)
```

# throat linear model + boxplot
```{r throat plot boxplot with linear model age and sex correction}

lm_boxplot_function <- function(dependent_variable, ylabel) {

  # Fit linear mixed-effects model
  print(summary(lmerTest::lmer(paste(dependent_variable, "~ visit_sum + sex + age_y + (1|id) + (1|Var1)"), data = tmp1)))
  lm <- summary(lmerTest::lmer(paste(dependent_variable, "~ visit_sum + sex + age_y + (1|id) + (1|Var1)"), data = tmp1))
  #lm <- summary(lmerTest::lmer(value ~ visit_sum + sex + age_y + (1|id), data = tmp1))
  # Extract coefficients and adjust p-values
  coefs <- data.frame(coef(lm))
  fdr <- p.adjust(coefs$Pr...t.., method = "fdr", n = nrow(coefs))
  
  # Create lm_stats table
  lm_stats <- bind_cols(coefs, fdr) %>%
    mutate(p = Pr...t..) %>%
    mutate(fdr = ...6) %>%
    select(-c(Pr...t.., ...6)) %>%
    rownames_to_column() %>%
    mutate(Months_after_ETI_start = rowname) %>%
    mutate(Months_after_ETI_start = case_when(
      Months_after_ETI_start == "(Intercept)" ~ "Baseline (Intercept)",
      Months_after_ETI_start == "visit_sum2" ~ "3 months",
      Months_after_ETI_start == "visit_sum3-5" ~ "6-12 months",
      Months_after_ETI_start == "visit_sum6-7" ~ "15-18 months",
      Months_after_ETI_start == "visit_sum8-10" ~ "21-24 months",
      Months_after_ETI_start %in% c("sex2", "age_y") ~ Months_after_ETI_start
    )) %>%
    mutate(fdr_star = case_when(
      fdr <= 0.001 ~ "***",
      fdr <= 0.01 ~ "**",
      fdr <= 0.05 ~ "*",
      fdr <= 0.1 ~ ".",
      fdr >= 0.1 ~ "ns"
    )) %>%
    select(-rowname) %>%
    select(Months_after_ETI_start, Estimate, Std..Error, df, t.value, p, fdr, fdr_star) %>%
    mutate(p = round(p, 5)) %>%
    mutate(fdr = round(fdr, 5))
  
  # Print lm_stats table
  print(lm_stats)
  
  # Save lm_stats table as HTML
  sjPlot::tab_df(lm_stats, title = paste("lmer(", dependent_variable, " ~ visit_sum + sex + age_y + (1|id), data=tmp1)"), file = paste("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/", dependent_variable, "_table.html", sep = ""))
  
  # Print boxplot
  group1 <- c("(Intercept)", rep("1", nrow(lm_stats) - 1))
  
  lm_sig<- lm_stats %>%
    bind_cols(group1) %>%
    mutate(group1 = ...9) %>%
    mutate(group2 = case_when(
      Months_after_ETI_start == "Baseline (Intercept)" ~ "1",
      Months_after_ETI_start == "3 months" ~ "2",
      Months_after_ETI_start == "6-12 months" ~ "3-5",
      Months_after_ETI_start == "15-18 months" ~ "6-7",
      Months_after_ETI_start == "21-24 months" ~ "8-10",
      Months_after_ETI_start %in% c("sex2", "age_y") ~ Months_after_ETI_start
    )) %>%
    filter(group1 != "(Intercept)") %>%
    filter(Months_after_ETI_start!="sex2") %>% 
    filter(Months_after_ETI_start!="age_y")
  
  max_y <- max(tmp1[[dependent_variable]], na.rm = TRUE)
  min_y <- min(tmp1[[dependent_variable]], na.rm = TRUE)
  
 # Calculate the adjusted y.position for stat labels
stat_labels_y_position <- max_y + 0.1 * (max_y - min_y)

  xlabels <- c("0", "3", "6-12", "15-18", "21-24")
  
# Print boxplot
boxplot <- tmp1 %>%
  ggplot(aes(get(dependent_variable), x = visit_sum)) +
  geom_boxplot(aes(fill = visit_sum), outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = visit_sum_palette) +
  theme_classic() +
  ylab(ylabel) +
  xlab("Months from ETI treatment start") +
  scale_x_discrete(labels = xlabels) +
  theme(text = element_text(size = 18), legend.position = "none") +
   stat_pvalue_manual(lm_sig, label = "fdr_star", y.position = stat_labels_y_position, step.increase = 0.05)

return(list(lm_stats = lm_stats, boxplot = boxplot))
}

throat <- lm_boxplot_function("value", "Throat: BC distance - CF & Control")
p2 <- throat$boxplot
p2
throat$lm_stats
```

```{r plot linear model coefficients throat}
library(coefplot)

# Fit your linear mixed-effects model
# For example, replace 'YourModel' and 'YourData' with your model and data, id is the CF patient id, and Var1 encodes the control id, by controlling for the pairs between CF patients and controls, implosion is controlled
model <-  lmerTest::lmer(value ~ visit_sum + sex + age_y + (1|id) + (1|Var1), data = tmp1)

# Extract coefficients
coefficients <- coef(model)

# Plot coefficients
  coef_p <- coefplot::coefplot(model, intercept = FALSE, color = "black")
  coef_p_t <- coef_p+
    theme_classic()+
    scale_y_discrete(labels= c("3 m", "6-12 m", "15-18 m", "21-24 m", "sex", "age"))+
    coord_flip()+
    xlab("Estimate")+
    ggtitle("")+
    ylab("")+
    theme(text = element_text(size = 18))
  coef_p_t
```

```{r combine plots, fig.width=13, fig.height=5.5}
ggarrange(p1,p2, widths = c(2:1))
#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/BetaandBC_throat_controls.png", width = 13, height = 5.5)
```

# PERMANOVA on CF-Control distances in throat
```{r}
# from tmp1 , the dataframe where I have all CF-Control distances listed, first I create a id for control samples

set.seed(123)

df_pm <- tmp1 %>%
  mutate(id_c = str_extract_all(comparison, "(?<=K|O)\\d+")) %>% 
  mutate(id_c = as_factor(sapply(id_c, function(x) as.integer(x[1]))))

# Select relevant columns
df <- df_pm %>%
  select(Var1, Var2, value)

df_s <- df_pm %>%
  select(Var1, Var2, value, visit_sum, id.y, id_c)

# Pivot data to wide format
df_wide <- df %>%
  pivot_wider(names_from = Var1, values_from = value)

# Set row names and remove the column used for row names
rownames(df_wide) <- df_wide$Var2

df_matrix <- df_wide %>% 
  select(-Var2)

# Convert to matrix
BC <- as.matrix(df_matrix)

# Ensure distinct rows for running adonis2
df_pm_d <- df_pm %>%
  distinct(Var2, .keep_all = TRUE)

# Run PERMANOVA for overall effect
adonis2(formula = BC ~ visit_sum + id.y, data = df_pm_d, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id.y
adonis2(formula = BC ~ visit_sum, data = df_pm_d, permutations = 999, by = "margin", strata = df_pm_d$id.y)

# Repeat PERMANOVA for separate time points, paired with baseline

df_pm_d_v1v2 <- df_pm_d %>%
  filter(visit_sum %in% c("1", "2"))

# Filter the distance matrix to only include relevant rows
df_v1v2 <- df_wide %>% 
  filter(Var2 %in% c(df_pm_d_v1v2$Var2))
df_v1v2$Var2<- NULL
# Convert to matrix
BC_v1v2 <- as.matrix(df_v1v2)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v1v2 ~ visit_sum + id.y, data = df_pm_d_v1v2, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id.y
adonis2(formula = BC_v1v2 ~ visit_sum, data = df_pm_d_v1v2, permutations = 999, by = "margin", strata = df_pm_d_v1v2$id.y)

# visits 6-12 months
df_pm_d_v1v35 <- df_pm_d %>%
  filter(visit_sum %in% c("1", "3-5"))

# Filter the distance matrix to only include relevant rows
df_v1v35 <- df_wide %>% 
  filter(Var2 %in% c(df_pm_d_v1v35$Var2))
df_v1v35$Var2<- NULL
# Convert to matrix
BC_v1v35 <- as.matrix(df_v1v35)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v1v35 ~ visit_sum + id.y, data = df_pm_d_v1v35, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id.y
adonis2(formula = BC_v1v35 ~ visit_sum, data = df_pm_d_v1v35, permutations = 999, by = "margin", strata = df_pm_d_v1v35$id.y)

# visits 15-18 months
df_pm_d_v1v67 <- df_pm_d %>%
  filter(visit_sum %in% c("1", "6-7"))

# Filter the distance matrix to only include relevant rows
df_v1v67 <- df_wide %>% 
  filter(Var2 %in% c(df_pm_d_v1v67$Var2))
df_v1v67$Var2<- NULL
# Convert to matrix
BC_v1v67 <- as.matrix(df_v1v67)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v1v67 ~ visit_sum + id.y, data = df_pm_d_v1v67, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id.y
adonis2(formula = BC_v1v67 ~ visit_sum, data = df_pm_d_v1v67, permutations = 999, by = "margin", strata = df_pm_d_v1v67$id.y)

# visits 21-24 months
df_pm_d_v1v89 <- df_pm_d %>%
  filter(visit_sum %in% c("1", "8-10"))

# Filter the distance matrix to only include relevant rows
df_v1v89 <- df_wide %>% 
  filter(Var2 %in% c(df_pm_d_v1v89$Var2))
df_v1v89$Var2<- NULL
# Convert to matrix
BC_v1v89 <- as.matrix(df_v1v89)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v1v89 ~ visit_sum + id.y, data = df_pm_d_v1v89, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id.y
adonis2(formula = BC_v1v89 ~ visit_sum, data = df_pm_d_v1v89, permutations = 999, by = "margin", strata = df_pm_d_v1v89$id.y)
```

# Analysis on most abundant taxa per patient in stool
```{r}
#Get top taxa per patient
#find.top.taxa2 is sourced from functions.R
top.stool<- find.top.taxa2(ps_full_stool_glom, "Genus",1)
top.stool$Species<- NULL

rslt <- top.stool[, "taxa"]
dd <- matrix(unlist(rslt), nrow=1)
colnames(dd) <- rownames(top.stool)
top.stool <- t(dd)

top.stool_df <- data.frame(x1 = row.names(top.stool), top.stool)%>%
  mutate(dominantGenus = top.stool)
top.stool_df$top.stool<- NULL

# add clinical data to p

##Add dominant Genus to ps_full_stool_glom sample data
ps_full_stool_glom_j <- microViz::ps_join(ps_full_stool_glom, top.stool_df, by = "x1")

##Add dominant Genus to ps_full_stool_glom sample data
ps_full_stool_j <- microViz::ps_join(ps_full_stool, top.stool_df, by = "x1")

### plot principal component analysis by visit_sum
PCA_all_visit_stool_dominantGenus <- plot_ordination(ps_full_stool_j, ordinate(ps_full_stool_j, "MDS"), color = "visit_sum", shape="project")

project_label <- c("Control","CF")

p5 <- PCA_all_visit_stool_dominantGenus +
  geom_point(size = 4, alpha = 0.7) +
  scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  scale_color_manual(values = visit_sum_palette, labels = xlabels) +
  scale_shape_manual(values = c(15, 16), labels = project_label) +
  theme_classic() +
  theme(text = element_text(size = 18), legend.position = "none", legend.text = element_text(size = 16), legend.title = element_blank()) +
  stat_ellipse()
p5

#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/beta_stool_controls.png", width = 10, height = 6)
```

# create density plots for axis 1 and 2
```{r, fig.width=4, fig.height=4}
# Perform PCoA
ord_obj <- ordinate(ps_full_stool_j, "MDS")

# Extract coordinates from PCoA
pcoa_coords <- ord_obj$vectors

# Create a data frame for ggplot
df <- data.frame(x1 = pcoa_coords[, 1], x2 = pcoa_coords[, 2], visit_sum = sample_data(ps_full_stool_j)$visit_sum, age_y=sample_data(ps_full_stool_j)$age_y, sex=sample_data(ps_full_stool_j)$sex, id.x=sample_data(ps_full_stool_j)$id.x)

# Create a named vector for custom labels
custom_labels <- c("1" = "0", 
                   "2" = "3", 
                   "3-5" = "6-12", 
                   "6-7" = "15-18", 
                   "8-10" = "21-24", 
                   "Control" = "Control")

# Create density plots on the x and y axes
plot_x_st <- ggplot(df, aes(x = x1, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  #clean_theme()+
  scale_x_continuous(breaks = c(-0.3, 0, 0.3), labels = c("-0.3", "0", "0.3"))+
  scale_fill_manual(values = visit_sum_palette, labels = xlabels)+
  theme(legend.position = "none", text = element_text(size = 18), legend.title = element_blank())

# Define the common y-axis limits
y_axis_limits <- c(0, 6.5)  # Adjust these values as needed

# Density plot with fixed y-axis range
axis1_den <- ggplot(df, aes(x = x1, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.8) +
  theme_classic() +
  scale_x_continuous(breaks = c(-0.3, 0, 0.3), labels = c("-0.3", "0", "0.3")) +
    scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  theme(
    legend.position = "none",
    text = element_text(size = 18),
    legend.title = element_blank(),
    strip.text.y = element_text(angle = 0)  # Rotate the facet labels
  ) +
  facet_grid(rows = vars(visit_sum), labeller =  as_labeller(custom_labels)) +
  ylim(y_axis_limits)+
  labs( x = "Axis 1")
axis1_den

# Boxplot comparing x between the 5 groups
axis1_bp <- ggplot(df, aes(x = as.factor(visit_sum), y = x1, fill = as.factor(visit_sum))) +
  geom_boxplot(alpha=0.8) +
  theme_classic() +
  scale_x_discrete(labels = xlabels) +
  scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  theme(
    legend.position = "none",
    text = element_text(size = 18),
    legend.title = element_blank()
  ) +
  labs(x = "Months from tt start", y = "Axis 1")
axis1_bp

plot_y_st <- ggplot(df, aes(x = x2, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  #clean_theme()+
  xlim(c(-0.6, 0.4))+
  scale_fill_manual(values = visit_sum_palette, labels = xlabels)+
  theme(legend.position = "none",text = element_text(size = 18), legend.title = element_blank())+
  coord_flip()

# Define the common y-axis limits
y_axis_limits <- c(0, 5)  # Adjust these values as needed

# Density plot with fixed y-axis range
axis2_den <- ggplot(df, aes(x = x2, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.8) +
  theme_classic() +
  scale_x_continuous(breaks = c(-0.6, 0, 0.4), labels = c("-0.6", "0", "0.4")) +
  scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  theme(
    legend.position = "none",
    text = element_text(size = 18),
    legend.title = element_blank(),
    strip.text.y = element_text(angle = 0)  # Rotate the facet labels
  ) +
  facet_grid(rows = vars(visit_sum), labeller =  as_labeller(custom_labels)) +
  ylim(y_axis_limits)+
  labs(x = "Axis 2")
axis2_den

# Boxplot comparing x between the 5 groups
axis2_bp <- ggplot(df, aes(x = as.factor(visit_sum), y = x2, fill = as.factor(visit_sum))) +
  geom_boxplot(alpha=0.8) +
  theme_classic() +
  scale_x_discrete(labels = xlabels) +
  scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  theme(
    legend.position = "none",
    text = element_text(size = 18),
    legend.title = element_blank(),
  ) +
  labs(x = "Months from tt start", y = "Axis 2")
axis2_bp

# Extract the legend. Returns a gtable
leg <- get_legend(p1_leg)

# Convert to a ggplot and print
legend <- as_ggplot(leg)

# Arranging the plot

axis <- ggarrange(axis1_den, axis2_den, axis1_bp, axis2_bp, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(1, 1),
          common.legend = F)
axis

stool_density <- ggarrange(plot_x_th, NULL, p1, plot_y_th, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(2, 1), heights = c(1, 2),
          common.legend = F)
stool_density
```

# add linear models on axis boxplots
```{r}
# Load required libraries
# Run function on  data
axis1 <- lm_corrected_pvalues_boxplot(df$x1, "Axis 1")
axis1_lm_stats <- axis1$lm_stats # Access lm_stats
p3 <- axis1$boxplot # Access boxplot

# Print the boxplot
print(p3)

axis2 <- lm_corrected_pvalues_boxplot(df$x2, "Axis 2")
axis2_lm_stats <- axis2$lm_stats # Access lm_stats
p4 <- axis2$boxplot # Access boxplot

# Print the boxplot
print(p4)

ggarrange(p3,p4)

# run statistics on controls vs all CF samples,
# in order to make this feasible in a linear model I have to revert the order of visit_sum
#
df$visit_sum_reorder <- factor(df$visit_sum, levels = c("Control", "1", "2", "3-5", "6-7", "8-10"))

# run CF vs Controls
#Axis 1
axis1.2 <- lm_corrected_pvalues_boxplot_reorder(df$x1, "Axis 1")
axis1_lm_stats_reorder <- axis1.2$lm_stats # Access lm_stats
axis1_reorder <- axis1.2$boxplot # Access boxplot
axis1_reorder

axi1_wC <- axis1$boxplot +
  geom_signif(
    y_position = c(-0.6, -0.55, -0.5, -0.45),  
    xmin = c(2, 3, 4, 5), 
    xmax = c(6, 6, 6, 6),
    annotation = c("***", "***", "***", "***"), 
    tip_length = -0.03, 
    vjust = 1,
    size = 0.35
  )
axi1_wC

#Axis 2
axis2.2 <- lm_corrected_pvalues_boxplot_reorder(df$x2, "Axis 2")
axis2_lm_stats_reorder <- axis2.2$lm_stats # Access lm_stats
axis2_reorder <- axis2.2$boxplot # Access boxplot
axis2_reorder
axi2_wC <- axis2$boxplot +
  geom_signif(
    y_position = c(-0.4, -0.35, -0.3, -0.25), 
    xmin = c(2, 3, 4, 5), 
    xmax = c(6, 6, 6, 6),
    annotation = c("ns", "ns", "ns", "ns"), 
    tip_length = -0.03, 
    vjust = 1,
    size = 0.35
  )
axi2_wC

# Arranging the plot

axis_stool <- ggarrange(axis1_den, axis2_den, axi1_wC, axi2_wC, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(1, 1),
          common.legend = F)
axis_stool


```

# Check nutrition and stool consistency on stool beta-diversity
```{r nutrition, fig.width=12}

### plot principal component analysis by nutrition
PCA_all_visit_stool_nutrition <- plot_ordination(ps_full_stool_j, ordinate(ps_full_stool_j, "MDS"), color = "basic_nutrition", shape="project")

project_label <- c("Control","CF")

PCA_all_visit_stool_nutrition+
  geom_point(size = 4)+
  ggtitle("Variation by time point with healthy controls")+
  #scale_color_manual(values = visit_sum_palette)+
  scale_shape_manual(values = c(23,15),labels = project_label)+
  theme_bw()+
  theme(text=element_text(size=24), legend.position = "right", legend.text = element_text(size=16), legend.title = element_text(size=18))+
  guides(color = guide_legend(title = "time points"))+
   guides(shape = guide_legend(title = "CF vs Control"))+
  stat_ellipse()

### plot principal component analysis by bristol
PCA_all_visit_stool_bristol <- plot_ordination(ps_full_stool_j, ordinate(ps_full_stool_j, "MDS"), color = "bristol", shape="project")

PCA_all_visit_stool_bristol+
  geom_point(size = 4)+
  ggtitle("Variation by time point with healthy controls")+
  #scale_color_manual(values = visit_sum_palette)+
  scale_shape_manual(values = c(23,15),labels = project_label)+
  theme_bw()+
  theme(text=element_text(size=24), legend.position = "right", legend.text = element_text(size=16), legend.title = element_text(size=18))+
  guides(color = guide_legend(title = "Bristol stool scale"))+
   guides(shape = guide_legend(title = "CF vs Control"))+
  stat_ellipse()

 #calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool_j,
                             method="bray", weighted=F)
#extract metadata
metadata<- as(sample_data(ps_full_stool_j),"data.frame")
metadata$id.x

nutri <- vegan::adonis2(BC_dist ~ basic_nutrition + bristol + sex+ id.x,
              permutations = 999, na.action=na.exclude, data = metadata)
nutri
```


# PERMANOVA in stool
```{r calculate bray curtis distance for all stool samples and run PERMANOVA}

# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool_j,
                             method="bray", weighted=F)
#extract metadata
metadata<- as(sample_data(ps_full_stool_j),"data.frame")

### stratified PERMANOVA per visit group
adonis2(formula = BC_dist ~ visit_sum + sex + age_y, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

### stratified PERMANOVA health vs CF
adonis2(formula = BC_dist ~ project + sex + age_y, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)
```

# compare CF and control BC-distances for stool
```{r compare CF and control BC-distances stool}
# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool_j,
                             method="bray", weighted=F)
BC_dist.stool <- as.matrix(BC_dist)
#BC_dist.stool[lower.tri(BC_dist.stool)] <- 0  # this is needed to remove the 0 latter, to keep only single pairs
BC_dist.stool_df <- reshape2::melt(BC_dist.stool)

tmp1 <- BC_dist.stool_df%>%
  filter(grepl("IMMP", Var1)) %>% #this keeps only control samples in Var 1
  filter(!grepl("IMMP", Var2))#%>%  #this filters control samples from Var 2, so that in the end only distances between COntrols and all CF samples are kept
  #filter(value!=0)#this removes the same sample distances, and the one from set 0 from above to keep a single pair

# merge with metadata to have visit_sums for CF samples
#extract metadata
metadata<- as(sample_data(ps_full_stool_j),"data.frame")
tmp1 <- tmp1 %>% 
  left_join(metadata, by=c("Var2"="x_sample_id")) %>% 
  mutate(comparison=paste(Var1, Var2, sep = "_")) %>% 
  distinct(comparison, .keep_all = T)

summary(tmp1$visit_sum)

my_comp <- list(c("1","2"), c("1","3-5"), c("1","6-7"), c("1","8-10"))

# Calculate median and IQR
summary_stats <- tmp1 %>%
  group_by(visit_sum) %>%
  summarize(
    median = median(value),
    lower = quantile(value, 0.25),
    upper = quantile(value, 0.75)
  )

tmp1 %>% 
  ggplot(aes(visit_sum, value, fill=visit_sum))+
  geom_boxplot(outlier.shape = NA, alpha=0.7)+
  #geom_point()+
  theme_classic()+
  scale_fill_manual(values = visit_sum_palette)+
  ylab("BC distance: CF and Controls")+
  theme(legend.position = "none",text=element_text(size=20))+ #
  scale_x_discrete(labels= xlabels)+
  xlab("")+
  stat_compare_means(comparisons = my_comp, method = "wilcox.test", label = "p.adj")

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/BC_stool_controls.png", width = 6, height = 5)

summary(lmerTest::lmer(value~visit_sum + age_y+sex+ (1|id.x)+ (1|Var1), data = tmp1))

# below I plot the BC distance per patient and each control, with controls being on the x-axis, one can observe the consistency of BC distance per patient across controls, with visit_sum giving a distinct signal in each patient
tmp1 %>% 
  ggplot(aes(Var1, value, fill=visit_sum))+
  #geom_boxplot(outlier.shape = NA)+
  geom_point(aes(color=visit_sum), alpha=0.5)+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  ylab("BC distance between CF and Controls")+
  theme(legend.position = "none",text=element_text(size=20))+ #
  #scale_x_discrete(labels= xlabels)+
  xlab("")+
  facet_wrap(~id.x)
```

# stool linear model + boxplot
```{r stool plot boxplot with linear model age and sex correction}

lm_boxplot_function <- function(dependent_variable, ylabel) {
  
  # Fit linear mixed-effects model
  print(summary(lmerTest::lmer(paste(dependent_variable, "~ visit_sum + sex + age_y + (1|id) + (1|Var1)"), data = tmp1)))
  lm <- summary(lmerTest::lmer(paste(dependent_variable, "~ visit_sum + sex + age_y + (1|id) + (1|Var1)"), data = tmp1))
  #lm <- summary(lmerTest::lmer(value ~ visit_sum + sex + age_y + (1|id), data = tmp1))
  # Extract coefficients and adjust p-values
  coefs <- data.frame(coef(lm))
  fdr <- p.adjust(coefs$Pr...t.., method = "fdr", n = nrow(coefs))
  
  # Create lm_stats table
  lm_stats <- bind_cols(coefs, fdr) %>%
    mutate(p = Pr...t..) %>%
    mutate(fdr = ...6) %>%
    select(-c(Pr...t.., ...6)) %>%
    rownames_to_column() %>%
    mutate(Months_after_ETI_start = rowname) %>%
    mutate(Months_after_ETI_start = case_when(
      Months_after_ETI_start == "(Intercept)" ~ "Baseline (Intercept)",
      Months_after_ETI_start == "visit_sum2" ~ "3 months",
      Months_after_ETI_start == "visit_sum3-5" ~ "6-12 months",
      Months_after_ETI_start == "visit_sum6-7" ~ "15-18 months",
      Months_after_ETI_start == "visit_sum8-10" ~ "21-24 months",
      Months_after_ETI_start %in% c("sex2", "age_y") ~ Months_after_ETI_start
    )) %>%
    mutate(fdr_star = case_when(
      fdr <= 0.001 ~ "***",
      fdr <= 0.01 ~ "**",
      fdr <= 0.05 ~ "*",
      fdr <= 0.1 ~ ".",
      fdr >= 0.1 ~ "ns"
    )) %>%
    select(-rowname) %>%
    select(Months_after_ETI_start, Estimate, Std..Error, df, t.value, p, fdr, fdr_star) %>%
    mutate(p = round(p, 5)) %>%
    mutate(fdr = round(fdr, 5))
  
  # Print lm_stats table
  print(lm_stats)
  
  # Save lm_stats table as HTML
  sjPlot::tab_df(lm_stats, title = paste("lmer(", dependent_variable, " ~ visit_sum + sex + age_y + (1|id), data=tmp1)"), file = paste("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/", dependent_variable, "_table.html", sep = ""))
  
  # Print boxplot
  group1 <- c("(Intercept)", rep("1", nrow(lm_stats) - 1))
  
  lm_sig <- lm_stats %>%
    bind_cols(group1) %>%
    mutate(group1 = ...9) %>%
    mutate(group2 = case_when(
      Months_after_ETI_start == "Baseline (Intercept)" ~ "1",
      Months_after_ETI_start == "3 months" ~ "2",
      Months_after_ETI_start == "6-12 months" ~ "3-5",
      Months_after_ETI_start == "15-18 months" ~ "6-7",
      Months_after_ETI_start == "21-24 months" ~ "8-10",
      Months_after_ETI_start %in% c("sex2", "age_y") ~ Months_after_ETI_start
    )) %>%
    filter(group1 != "(Intercept)") %>%
    filter(Months_after_ETI_start!="sex2") %>% 
    filter(Months_after_ETI_start!="age_y")
  
  max_y <- max(tmp1[[dependent_variable]], na.rm = TRUE)
  min_y <- min(tmp1[[dependent_variable]], na.rm = TRUE)
  
 # Calculate the adjusted y.position for stat labels
stat_labels_y_position <- max_y + 0.1 * (max_y - min_y)

  xlabels <- c("0", "3", "6-12", "15-18", "21-24")
  
# Print boxplot
boxplot <- tmp1 %>%
  ggplot(aes(get(dependent_variable), x = visit_sum)) +
  geom_boxplot(aes(fill = visit_sum), outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = visit_sum_palette) +
  theme_classic() +
  ylab(ylabel) +
  xlab("Months from ETI treatment start") +
  scale_x_discrete(labels = xlabels) +
  theme(text = element_text(size = 18), legend.position = "none") +
   stat_pvalue_manual(lm_sig, label = "fdr_star", y.position = stat_labels_y_position, step.increase = 0.05)

return(list(lm_stats = lm_stats, boxplot = boxplot))
}

stool <- lm_boxplot_function("value", "Stool: BC distance - CF & Control")
p4 <- stool$boxplot
stool$lm_stats
```

```{r plot linear model coefficients stool}
# Fit your linear mixed-effects model
# For example, replace 'YourModel' and 'YourData' with your model and data
model <-  lmerTest::lmer(value ~ visit_sum + sex + age_y + (1|id) + (1|Var1), data = tmp1)

# Extract coefficients
coefficients <- coef(model)

# Extract coefficients
coefficients <- coef(model)

# Plot coefficients
  coef_p <- coefplot::coefplot(model, intercept = FALSE, color = "black")
  coef_p_s<- coef_p+
    theme_classic()+
    scale_y_discrete(labels= c("3 m", "6-12 m", "15-18 m", "21-24 m", "sex", "age"))+
    coord_flip()+
    xlab("Estimate")+
    ggtitle("")+
    ylab("")+
    theme(text = element_text(size = 18))
  coef_p_s
```

# PERMANOVA on CF-Control distances in stool
```{r}
# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool_j,
                             method="bray", weighted=F)
#extract metadata
metadata<- as(sample_data(ps_full_stool_j),"data.frame")

### unstratified

adonis2(formula = BC_dist ~ visit_sum, data = metadata, permutations = 999, by = "margin")

adonis2(formula = BC_dist ~ visit_sum + sex + age_y, data = metadata, permutations = 999, by = "margin")

adonis2(formula = BC_dist ~ visit_sum + sex + age_y +id.x, data = metadata, permutations = 999, by = "margin") # sex and age are not significant when including id.x as a explantory variable

adonis2(formula = BC_dist ~ visit_sum +id.x, data = metadata, permutations = 999, by = "margin") # sex and age are not significant when including id.x as a explantory variable

### stratified PERMANOVA per id
adonis2(formula = BC_dist ~ visit_sum , data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

adonis2(formula = BC_dist ~ visit_sum + project , data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

### stratified PERMANOVA per project (health vs CF)
adonis2(formula = BC_dist ~ visit_sum + id.x, data = metadata, permutations = 999, by = "margin", strata = metadata$project)

### unstratified PERMANOVA health vs CF
adonis2(formula = BC_dist ~ project, data = metadata, permutations = 999, by = "margin")
adonis2(formula = BC_dist ~ project + id.x, data = metadata, permutations = 999, by = "margin")
adonis2(formula = BC_dist ~ project + visit_sum, data = metadata, permutations = 999, by = "margin")
### stratified PERMANOVA health vs CF
adonis2(formula = BC_dist ~ project + visit_sum + sex + age_y, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

adonis2(formula = BC_dist ~ project + id.x, data = metadata, permutations = 999, by = "margin", strata = metadata$visit_sum)

adonis2(formula = BC_dist ~ visit_sum + id.x, data = metadata, permutations = 999, by = "margin", strata = metadata$project)
```

# PERMANOVA per time point comparison

```{r}

# check baseline vs follow-up samples
metadata<- metadata %>% 
  mutate(ETI_yn = as_factor(case_when(visit_sum=="1"~"no", visit_sum=="Control"~"Control", T~"yes")))

summary(metadata$ETI_yn)

### plot principal component analysis by visit_sum
PCA_all_visit_stool_dominantGenus <- plot_ordination(ps_full_stool_j, ordinate(ps_full_stool_j, "MDS"), color = "visit_sum", shape="project")

project_label <- c("Control","CF")

labels <- c("Baseline", "ETI", "ETI", "ETI", "ETI", "Control")

bfu_palette <- c("black", "#55AD89FF","#55AD89FF","#55AD89FF","#55AD89FF","#8176AA" )

PCA_all_visit_stool_dominantGenus +
  geom_point(size = 4, alpha = 0.7) +
  scale_color_manual(values = bfu_palette, labels = xlabels) +
  scale_shape_manual(values = c(15, 16), labels = project_label) +
  theme_classic() +
  theme(text = element_text(size = 18), legend.position = "none", legend.text = element_text(size = 16), legend.title = element_blank()) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 2))+
  stat_ellipse()

### stratified PERMANOVA control vs baseline vs follow-up
adonis2(formula = BC_dist ~ ETI_yn, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

# Repeat PERMANOVA for separate time points, paired with baseline

# as there are no repeated samples no need for stratification
adonis2(formula = BC_v1 ~ project, data = metadata_v1, permutations = 999, by = "margin")

# now Control vs all follow-ups together (without baseline)

metadata_fu <- metadata %>%
  filter(ETI_yn %in% c("yes", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_fu$x_sample_id
# Filter the distance matrix to only include relevant rows and cols
BC_df <- as.matrix(BC_dist)
BC_df <- as.data.frame(BC_df)

# Filter rows based on sample IDs
df_fu <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_fu <- df_fu[, colnames(df_fu) %in% sample_ids]

# Convert to matrix
BC_fu <- as.matrix(df_fu)

### stratified
adonis2(formula = BC_fu ~ visit_sum, data = metadata_fu, permutations = 999, by = "margin", strata = metadata_fu$id.x)
adonis2(formula = BC_fu ~ ETI_yn, data = metadata_fu, permutations = 999, by = "margin", strata = metadata_fu$id.x)

# now Control vs follow-ups separately
metadata_v2 <- metadata %>%
  filter(visit_sum %in% c("2", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v2$x_sample_id

# Filter rows based on sample IDs
df_v2 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v2 <- df_v2[, colnames(df_v2) %in% sample_ids]

# Convert to matrix
BC_v2 <- as.matrix(df_v2)

### unstratified
adonis2(formula = BC_v2 ~ visit_sum, data = metadata_v2, permutations = 999, by = "margin")
# as there are no repeated samples no need for stratification - it also gives an error
adonis2(formula = BC_v2 ~ project, data = metadata_v2, permutations = 999, by = "margin")

metadata_v3 <- metadata %>%
  filter(visit_sum %in% c("3-5", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v3$x_sample_id

# Filter rows based on sample IDs
df_v3 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v3 <- df_v3[, colnames(df_v3) %in% sample_ids]

# Convert to matrix
BC_v3 <- as.matrix(df_v3)

### unstratified
adonis2(formula = BC_v3 ~ visit_sum, data = metadata_v3, permutations = 999, by = "margin")
adonis2(formula = BC_v3 ~ project, data = metadata_v3, permutations = 999, by = "margin")
# as there are repeated samples, I will block id to verify this observation
adonis2(formula = BC_v3 ~ visit_sum, data = metadata_v3, permutations = 999, by = "margin", strata = metadata_v3$id.x)

metadata_v6 <- metadata %>%
  filter(visit_sum %in% c("6-7", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v6$x_sample_id

# Filter rows based on sample IDs
df_v6 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v6 <- df_v6[, colnames(df_v6) %in% sample_ids]

# Convert to matrix
BC_v6 <- as.matrix(df_v6)

### unstratified
adonis2(formula = BC_v6 ~ visit_sum, data = metadata_v6, permutations = 999, by = "margin")
adonis2(formula = BC_v6 ~ project, data = metadata_v6, permutations = 999, by = "margin")
# as there are repeated samples, I will block id to verify this observation
adonis2(formula = BC_v6 ~ visit_sum, data = metadata_v6, permutations = 999, by = "margin", strata = metadata_v6$id.x)

metadata_v8 <- metadata %>%
  filter(visit_sum %in% c("8-10", "Control"))
# Extract relevant sample IDs
sample_ids <- metadata_v8$x_sample_id

# Filter rows based on sample IDs
df_v8 <- BC_df[rownames(BC_df) %in% sample_ids, ]
# Filter columns based on sample IDs
df_v8 <- df_v8[, colnames(df_v8) %in% sample_ids]

# Convert to matrix
BC_v8 <- as.matrix(df_v8)

### unstratified
adonis2(formula = BC_v8 ~ visit_sum, data = metadata_v8, permutations = 999, by = "margin")
adonis2(formula = BC_v8 ~ project, data = metadata_v8, permutations = 999, by = "margin")
# as there are repeated samples, I will block id to verify this observation
adonis2(formula = BC_v8 ~ visit_sum, data = metadata_v8, permutations = 999, by = "margin", strata = metadata_v8$id.x)
```

# PERMANOVA on CF-health comparisons alone

```{r}

# I reduce the distance matrix to only have CF samples in the rows, and only healthy samples in the columns, so that only CF to healthy comparisons are retained

# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool_j,
                             method="bray", weighted=F)

#extract metadata
metadata<- as(sample_data(ps_full_stool_j),"data.frame")

# Use grep to find columns that contain "IMMP" in their names
cols_to_keep <- metadata %>% 
  filter(project=="IMMPIMMP") %>% 
  select(x_sample_id)
cols_to_keep <- cols_to_keep$x_sample_id

rows_to_keep <- metadata %>% 
  filter(project=="IMP") %>% 
  select(x_sample_id)
rows_to_keep <- rows_to_keep$x_sample_id
# Convert the distance object to a matrix
BC_matrix <- as.matrix(BC_dist)

# Subset the matrix to keep only those columns that are from controls
BC_matrix_f <- BC_matrix[, cols_to_keep]
BC_matrix_f <- BC_matrix[rows_to_keep, ]

# no keep in the metadata only CF samples
metadata <- metadata %>% 
  filter(project=="IMP")

### unstratified
adonis2(formula = BC_matrix_f ~ visit_sum, data = metadata, permutations = 999, by = "margin")

adonis2(formula = BC_matrix_f ~ visit_sum + id.x, data = metadata, permutations = 999, by = "margin")

### stratified PERMANOVA per id
adonis2(formula = BC_matrix_f ~ visit_sum, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

# check the impact of number of treatment days
adonis2(formula = BC_matrix_f ~ delta_ETI_merged, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

# check baseline vs follow-up samples
metadata<- metadata %>% 
  mutate(ETI_yn = case_when(visit_sum=="1"~FALSE, T~TRUE))

### unstratified
adonis2(formula = BC_matrix_f ~ ETI_yn, data = metadata, permutations = 999, by = "margin")

adonis2(formula = BC_matrix_f ~ ETI_yn + id.x, data = metadata, permutations = 999, by = "margin")

### stratified PERMANOVA per id
adonis2(formula = BC_matrix_f ~ ETI_yn, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

# Repeat PERMANOVA for separate time points, paired with baseline

md_v1v2 <- metadata %>%
  filter(visit_sum %in% c("1", "2"))

# Filter the distance matrix to only include relevant rows
BC_matrix_f <- as.data.frame(BC_matrix_f)
BC_v1v2 <- BC_matrix_f %>% 
  filter(rownames(BC_matrix_f) %in% c(md_v1v2$x_sample_id))

# Convert to matrix
BC_v1v2 <- as.matrix(BC_v1v2)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v1v2 ~ visit_sum , data = md_v1v2, permutations = 999, by = "margin")
adonis2(formula = BC_v1v2 ~ visit_sum + id.x, data = md_v1v2, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id
adonis2(formula = BC_v1v2 ~ visit_sum, data = md_v1v2, permutations = 999, by = "margin", strata = md_v1v2$id.x)

# 6-12 months + baseline
md_v1v35 <- metadata %>%
  filter(visit_sum %in% c("1", "3-5"))

# Filter the distance matrix to only include relevant rows
BC_v35 <- BC_matrix_f %>% 
  filter(rownames(BC_matrix_f) %in% c(md_v1v35$x_sample_id))

# Convert to matrix
BC_v35 <- as.matrix(BC_v35)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v35 ~ visit_sum , data = md_v1v35, permutations = 999, by = "margin")
adonis2(formula = BC_v35 ~ visit_sum + id.x, data = md_v1v35, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id
adonis2(formula = BC_v35 ~ visit_sum, data = md_v1v35, permutations = 999, by = "margin", strata = md_v1v35$id.x)

# 21-24 months + baseline
md_v1v89 <- metadata %>%
  filter(visit_sum %in% c("1", "8-10"))

# Filter the distance matrix to only include relevant rows
BC_v89 <- BC_matrix_f %>% 
  filter(rownames(BC_matrix_f) %in% c(md_v1v89$x_sample_id))

# Convert to matrix
BC_v89 <- as.matrix(BC_v89)

# Run PERMANOVA for specific time points
adonis2(formula = BC_v89 ~ visit_sum , data = md_v1v89, permutations = 999, by = "margin")
adonis2(formula = BC_v89 ~ visit_sum + id.x, data = md_v1v89, permutations = 999, by = "margin")

# Run PERMANOVA for visit_sum, controlling for id
adonis2(formula = BC_v89 ~ visit_sum, data = md_v1v89, permutations = 999, by = "margin", strata = md_v1v89$id.x)


```

# combine plots
```{r, fig.width=16, fig.height=18}
library(gridExtra)
p1 <- p1+
  ggtitle("Throat")

p5 <- p5+
  ggtitle("Stool")

grid_arranged_t <- grid.arrange(
  p1+ggtitle(""),
  NULL,
  plot_x_th+theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks = element_blank())+ggtitle("Throat")+labs(tag = "A") ,
  plot_y_th+theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks = element_blank()),
  p5+ggtitle(""),
  NULL,
  plot_x_st+theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks = element_blank())+ggtitle("Stool")+labs(tag = "D") ,
  plot_y_st+theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks = element_blank()),
  coef_p_t + labs(tag = "B"),
  p2 + labs(tag = "C"),
  coef_p_s + labs(tag = "E"),
  p4 + labs(tag = "F"),
  legend,
  ncol = 3, nrow = 5,
  heights = c(1, 2, 1,2, 0.5),
  widths=c(2,0.5,1.5),
  layout_matrix = rbind(c(3,2,9),
                        c(1,4,10),
                        c(7,6,11),
                        c(5,8,12),
                        c(13,13,13)
                      )
)

# Print or save the arranged plot
#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/Submission_CHM/Fig3.tiff",grid_arranged_t, dpi = 300, width = 16, height = 18)

# combine stats table

throat$lm_stats <- throat$lm_stats %>% 
  mutate(Sample="Throat") %>% 
  select(Sample, everything())

stool$lm_stats <- stool$lm_stats %>% 
  mutate(Sample="Stool") %>% 
  select(Sample, everything())

bc_stats <- rbind(throat$lm_stats, stool$lm_stats)
kable(bc_stats)
#write_csv(bc_stats, "/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/lmm_bc_distance_control.csv")
```
```{r}
sessionInfo()
```

