---
title: "CF-Control distance analysis"
author: "Rebecca L. Knoll"
date: " last edit `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: show
    number_sections: yes
    smart: no
    toc: yes
    df_print: paged
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries and run functions script}
library(microViz)

pacman::p_load(tidyverse, phyloseq, magrittr, janitor, microbiome, knitr, lubridate, naniar, readxl, ggplot2, ggpubr)

library(dendextend)
library(vegan)

ps_full <- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/ps_IMProveCF_patientsAndControls_Run1-23_18102023.rds")

source("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/IMMProveCF/functions_full.R")

visit_sum_palette <- c("black", "#C3BC3FFF", "#6388B4FF", "#BB7693FF", "#55AD89FF", "#8176AA")

dom_palette <- c(Streptococcus="#69b3a2", Staphylococcus = "#8175AA", Fusobacterium="#6FB899FF", Prevotella_7="#31A1B3FF", Rothia="#027B8EFF", Pseudomonas="#EE6AA7", Eikenella="#94D0C0FF", Haemophilus="#CCB22BFF", Achromobacter="#9F8F12FF", Gemella= "#97CFD0" , Moraxella= "#6FB899", `missing sample` = "#CFCFCF", `Burkholderia-Caballeronia-Paraburkholderia` = "#E5C494")
```

```{r, prepare data}

ps_full_relab <- transform_sample_counts(ps_full, function(x) x/sum(x)*100)

#subset materials
ps_full_throat <- subset_samples(ps_full_relab, material== "Throat")
ps_full_stool <- subset_samples(ps_full_relab, material== "Stool")

ps_full_throat_glom <- tax_glom(ps_full_throat, taxrank = "Genus")
ps_full_stool_glom <- tax_glom(ps_full_stool, taxrank = "Genus")
```

# Analysis on most abundant taxa per patient in throat
```{r}
#Get top taxa per patient
#find.top.taxa2 is sourced from functions.R
top.throat<- find.top.taxa2(ps_full_throat_glom, "Genus",1)
top.throat$Species<- NULL

rslt <- top.throat[, "taxa"]
dd <- matrix(unlist(rslt), nrow=1)
colnames(dd) <- rownames(top.throat)
top.throat <- t(dd)

top.throat_df <- data.frame(x1 = row.names(top.throat), top.throat)%>%
  mutate(dominantGenus = top.throat)
top.throat_df$top.throat<- NULL

# add clinical data to p

##Add dominant Genus to ps_full_throat_glom sample data
ps_full_throat_glom_j <- microViz::ps_join(ps_full_throat_glom, top.throat_df, by = "x1")

##Add dominant Genus to ps_full_throat sample data
ps_full_throat_j <- microViz::ps_join(ps_full_throat, top.throat_df, by = "x1")

# Control's dominant taxa
summary(as_factor(sample_data(ps_full_throat_j)$dominantGenus[sample_data(ps_full_throat_j)$project=="IMMPIMMP"]))

# Patients's dominant taxa
summary(as_factor(sample_data(ps_full_throat_j)$dominantGenus[sample_data(ps_full_throat_j)$project!="IMMPIMMP"]))

 ### plot principal component analysis by dominant Genus
PCA_all_visit_throat_dominantGenus <- plot_ordination(ps_full_throat_j, ordinate(ps_full_throat_j, "MDS"), color = "dominantGenus", shape="visit_sum")

xlabels <- c("0","3","6-12","15-18","21-24","Control")

PCA_all_visit_throat_dominantGenus+
  geom_line(aes(group=id.x), color="grey")+
  geom_point(size = 4)+
  ggtitle("Variation by dominant Genus")+
  scale_color_manual(values = dom_palette)+
  scale_shape_manual(values = c(15, 16, 17, 18, 12, 23),labels = xlabels)+
  theme_bw()+
  theme(text=element_text(size=24), legend.position = "right", legend.text = element_text(size=16), legend.title = element_text(size=18))+
  guides(color = guide_legend(title = "dominant genus"))+
   guides(shape = guide_legend(title = "months from treatment start"))

### plot principal component analysis by visit_sum
PCA_all_visit_throat_dominantGenus <- plot_ordination(ps_full_throat_j, ordinate(ps_full_throat_j, "MDS"), color = "visit_sum", shape="project")

project_label <- c("Control","CF")

xlabels <- c("0", "3", "6-12", "15-18", "21-24", "Control")

p1 <- PCA_all_visit_throat_dominantGenus +
  geom_point(size = 4, alpha = 0.7) +
  scale_color_manual(values = visit_sum_palette, labels = xlabels) +
  scale_shape_manual(values = c(15, 16), labels = project_label) +
  theme_classic() +
  theme(text = element_text(size = 18), legend.position = "none", legend.text = element_text(size = 16), legend.title = element_blank()) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 2))+
  stat_ellipse()
p1

p1_leg <- PCA_all_visit_throat_dominantGenus +
  geom_point(size = 4, alpha = 0.7) +
  scale_color_manual(values = visit_sum_palette, labels = xlabels) +
  scale_shape_manual(values = c(15, 16), labels = project_label) +
  theme_classic() +
  theme(text = element_text(size = 16), legend.position = "bottom", legend.text = element_text(size = 16)) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 6),shape = guide_legend(title = "", title.position = "left", ncol = 2))+
  stat_ellipse()
p1_leg

#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/beta_throat_controls.png", width = 10, height = 6)

```
# create density plots for axis 1 and 2
```{r, fig.width=4, fig.height=4}
# Perform PCoA
ord_obj <- ordinate(ps_full_throat_j, "MDS")

# Extract coordinates from PCoA
pcoa_coords <- ord_obj$vectors

# Create a data frame for ggplot
df <- data.frame(x = pcoa_coords[, 1], y = pcoa_coords[, 2], visit_sum = sample_data(ps_full_throat_j)$visit_sum)

# Create density plots on the x and y axes
plot_x_th <- ggplot(df, aes(x = x, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  #clean_theme()+
  scale_x_continuous(breaks = c(-0.3, 0, 0.3), labels = c("-0.3", "0", "0.3"))+
  scale_fill_manual(values = visit_sum_palette, labels = xlabels)+
  theme(legend.position = "none", text = element_text(size = 18), legend.title = element_blank())


plot_y_th <- ggplot(df, aes(x = y, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  #clean_theme()+
  xlim(c(-0.6, 0.4))+
  scale_fill_manual(values = visit_sum_palette, labels = xlabels)+
  theme(legend.position = "none",text = element_text(size = 18), legend.title = element_blank())+
  coord_flip()

# Extract the legend. Returns a gtable
leg <- get_legend(p1_leg)

# Convert to a ggplot and print
legend <- as_ggplot(leg)

# Arranging the plot
throat_density <- ggarrange(plot_x_th, NULL, p1, plot_y_th, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(2, 1), heights = c(1, 2),
          common.legend = F)
throat_density
```


# PERMANOVA in throat
```{r calculate bray curtis distance for all throat samples and run PERMANOVA}

# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_throat_j,
                             method="bray", weighted=F)
#extract metadata
metadata<- as(sample_data(ps_full_throat_j),"data.frame")

### stratified PERMANOVA per visit group
adonis2(formula = BC_dist ~ visit_sum + sex + age_y, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

### stratified PERMANOVA health vs CF
adonis2(formula = BC_dist ~ project + sex + age_y, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)
```

# compare CF and control BC-distances for throat
```{r compare CF and control BC-distances}
# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_throat_j,
                             method="bray", weighted=F)
BC_dist.throat <- as.matrix(BC_dist)
#BC_dist.throat[lower.tri(BC_dist.throat)] <- 0  # this is needed to remove the 0 latter, to keep only single pairs
BC_dist.throat_df <- reshape2::melt(BC_dist.throat)

tmp1 <- BC_dist.throat_df%>%
  filter(grepl("IMMP", Var1)) %>% #this keeps only control samples in Var 1
  filter(!grepl("IMMP", Var2))#%>%  #this filters control samples from Var 2, so that in the end only distances between COntrols and all CF samples are kept
  #filter(value!=0)#this removes the same sample distances, and the one from set 0 from above to keep a single pair

# merge with metadata to have visit_sums for CF samples
#extract metadata
metadata<- as(sample_data(ps_full_throat_j),"data.frame")
tmp1 <- tmp1 %>% 
  left_join(metadata, by=c("Var2"="x_sample_id")) %>% 
  mutate(comparison=paste(Var1, Var2, sep = "_")) %>% 
  distinct(comparison, .keep_all = T)

summary(tmp1$visit_sum)

my_comp <- list(c("1","2"), c("1","3-5"), c("1","6-7"), c("1","8-10"))

# Calculate median and IQR
summary_stats <- tmp1 %>%
  group_by(visit_sum) %>%
  summarize(
    median = median(value),
    lower = quantile(value, 0.25),
    upper = quantile(value, 0.75)
  )

tmp1 %>% 
  ggplot(aes(visit_sum, value, fill=visit_sum))+
  geom_boxplot(outlier.shape = NA, alpha=0.7)+
  #geom_point()+
  theme_classic()+
  scale_fill_manual(values = visit_sum_palette)+
  ylab("BC distance between CF and Controls")+
  theme(legend.position = "none",text=element_text(size=20))+ #
  scale_x_discrete(labels= xlabels)+
  xlab("")+
  stat_compare_means(comparisons = my_comp, method = "wilcox.test", label = "p.adj")

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/BC_throat_controls.png", width = 8, height = 6)

summary(lmerTest::lmer(value~visit_sum + (1|Var1)+ (1|id.x), data = tmp1))

# below I plot the BC distance per patient and each control, with controls being on the x-axis, one can observe the consistency of BC distance per patient across controls, with visit_sum giving a distinct signal in each patient
tmp1 %>% 
  ggplot(aes(Var1, value, fill=visit_sum))+
  #geom_boxplot(outlier.shape = NA)+
  geom_point(aes(color=visit_sum), alpha=0.5)+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  ylab("BC distance between CF and Controls")+
  theme(legend.position = "none",text=element_text(size=20))+ #
  #scale_x_discrete(labels= xlabels)+
  xlab("")+
  facet_wrap(~id.x)
```

# throat linear model + boxplot
```{r throat plot boxplot with linear model age and sex correction}

lm_boxplot_function <- function(dependent_variable, ylabel) {

  # Fit linear mixed-effects model
  print(summary(lmerTest::lmer(paste(dependent_variable, "~ visit_sum + sex + age_y + (1|id) + (1|Var1)"), data = tmp1)))
  lm <- summary(lmerTest::lmer(paste(dependent_variable, "~ visit_sum + sex + age_y + (1|id) + (1|Var1)"), data = tmp1))
  #lm <- summary(lmerTest::lmer(value ~ visit_sum + sex + age_y + (1|id), data = tmp1))
  # Extract coefficients and adjust p-values
  coefs <- data.frame(coef(lm))
  fdr <- p.adjust(coefs$Pr...t.., method = "fdr", n = nrow(coefs))
  
  # Create lm_stats table
  lm_stats <- bind_cols(coefs, fdr) %>%
    mutate(p = Pr...t..) %>%
    mutate(fdr = ...6) %>%
    select(-c(Pr...t.., ...6)) %>%
    rownames_to_column() %>%
    mutate(Months_after_ETI_start = rowname) %>%
    mutate(Months_after_ETI_start = case_when(
      Months_after_ETI_start == "(Intercept)" ~ "Baseline (Intercept)",
      Months_after_ETI_start == "visit_sum2" ~ "3 months",
      Months_after_ETI_start == "visit_sum3-5" ~ "6-12 months",
      Months_after_ETI_start == "visit_sum6-7" ~ "15-18 months",
      Months_after_ETI_start == "visit_sum8-10" ~ "21-24 months",
      Months_after_ETI_start %in% c("sex2", "age_y") ~ Months_after_ETI_start
    )) %>%
    mutate(fdr_star = case_when(
      fdr <= 0.001 ~ "***",
      fdr <= 0.01 ~ "**",
      fdr <= 0.05 ~ "*",
      fdr <= 0.1 ~ ".",
      fdr >= 0.1 ~ "ns"
    )) %>%
    select(-rowname) %>%
    select(Months_after_ETI_start, Estimate, Std..Error, df, t.value, p, fdr, fdr_star) %>%
    mutate(p = round(p, 5)) %>%
    mutate(fdr = round(fdr, 5))
  
  # Print lm_stats table
  print(lm_stats)
  
  # Save lm_stats table as HTML
  sjPlot::tab_df(lm_stats, title = paste("lmer(", dependent_variable, " ~ visit_sum + sex + age_y + (1|id), data=tmp1)"), file = paste("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/", dependent_variable, "_table.html", sep = ""))
  
  # Print boxplot
  group1 <- c("(Intercept)", rep("1", nrow(lm_stats) - 1))
  
  lm_sig<- lm_stats %>%
    bind_cols(group1) %>%
    mutate(group1 = ...9) %>%
    mutate(group2 = case_when(
      Months_after_ETI_start == "Baseline (Intercept)" ~ "1",
      Months_after_ETI_start == "3 months" ~ "2",
      Months_after_ETI_start == "6-12 months" ~ "3-5",
      Months_after_ETI_start == "15-18 months" ~ "6-7",
      Months_after_ETI_start == "21-24 months" ~ "8-10",
      Months_after_ETI_start %in% c("sex2", "age_y") ~ Months_after_ETI_start
    )) %>%
    filter(group1 != "(Intercept)") %>%
    filter(Months_after_ETI_start!="sex2") %>% 
    filter(Months_after_ETI_start!="age_y")
  
  max_y <- max(tmp1[[dependent_variable]], na.rm = TRUE)
  min_y <- min(tmp1[[dependent_variable]], na.rm = TRUE)
  
 # Calculate the adjusted y.position for stat labels
stat_labels_y_position <- max_y + 0.1 * (max_y - min_y)

  xlabels <- c("0", "3", "6-12", "15-18", "21-24")
  
# Print boxplot
boxplot <- tmp1 %>%
  ggplot(aes(get(dependent_variable), x = visit_sum)) +
  geom_boxplot(aes(fill = visit_sum), outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = visit_sum_palette) +
  theme_classic() +
  ylab(ylabel) +
  xlab("Months from ETI treatment start") +
  scale_x_discrete(labels = xlabels) +
  theme(text = element_text(size = 18), legend.position = "none") +
   stat_pvalue_manual(lm_sig, label = "fdr_star", y.position = stat_labels_y_position, step.increase = 0.05)

return(list(lm_stats = lm_stats, boxplot = boxplot))
}

throat <- lm_boxplot_function("value", "Throat: BC distance - CF & Control")
p2 <- throat$boxplot
p2
throat$lm_stats
```
```{r plot linear model coefficients throat}
library(coefplot)

# Fit your linear mixed-effects model
# For example, replace 'YourModel' and 'YourData' with your model and data, id is the CF patient id, and Var1 encodes the control id, by controlling for the pairs between CF patients and controls, implosion is controlled
model <-  lmerTest::lmer(value ~ visit_sum + sex + age_y + (1|id) + (1|Var1), data = tmp1)

# Extract coefficients
coefficients <- coef(model)

# Plot coefficients
  coef_p <- coefplot::coefplot(model, intercept = FALSE, color = "black")
  coef_p_t <- coef_p+
    theme_classic()+
    scale_y_discrete(labels= c("3 m", "6-12 m", "15-18 m", "21-24 m", "sex", "age"))+
    coord_flip()+
    xlab("Estimate")+
    ggtitle("")+
    ylab("")+
    theme(text = element_text(size = 18))
  coef_p_t
```

```{r combine plots, fig.width=13, fig.height=5.5}
ggarrange(p1,p2, widths = c(2:1))
#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/BetaandBC_throat_controls.png", width = 13, height = 5.5)
```

# Analysis on most abundant taxa per patient in stool
```{r}
#Get top taxa per patient
#find.top.taxa2 is sourced from functions.R
top.stool<- find.top.taxa2(ps_full_stool_glom, "Genus",1)
top.stool$Species<- NULL

rslt <- top.stool[, "taxa"]
dd <- matrix(unlist(rslt), nrow=1)
colnames(dd) <- rownames(top.stool)
top.stool <- t(dd)

top.stool_df <- data.frame(x1 = row.names(top.stool), top.stool)%>%
  mutate(dominantGenus = top.stool)
top.stool_df$top.stool<- NULL

# add clinical data to p

##Add dominant Genus to ps_full_stool_glom sample data
ps_full_stool_glom_j <- microViz::ps_join(ps_full_stool_glom, top.stool_df, by = "x1")

##Add dominant Genus to ps_full_stool_glom sample data
ps_full_stool_j <- microViz::ps_join(ps_full_stool, top.stool_df, by = "x1")

### plot principal component analysis by visit_sum
PCA_all_visit_stool_dominantGenus <- plot_ordination(ps_full_stool_j, ordinate(ps_full_stool_j, "MDS"), color = "visit_sum", shape="project")

project_label <- c("Control","CF")

p3 <- PCA_all_visit_stool_dominantGenus +
  geom_point(size = 4, alpha = 0.7) +
  scale_fill_manual(values = visit_sum_palette, labels = xlabels) +
  scale_color_manual(values = visit_sum_palette, labels = xlabels) +
  scale_shape_manual(values = c(15, 16), labels = project_label) +
  theme_classic() +
  theme(text = element_text(size = 18), legend.position = "none", legend.text = element_text(size = 16), legend.title = element_blank()) +
  stat_ellipse()
p3

#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/beta_stool_controls.png", width = 10, height = 6)
```
# create density plots for axis 1 and 2 stool
```{r, fig.width=4, fig.height=4}
# Perform PCoA
ord_obj <- ordinate(ps_full_stool_j, "MDS")

# Extract coordinates from PCoA
pcoa_coords <- ord_obj$vectors

# Create a data frame for ggplot
df <- data.frame(x = pcoa_coords[, 1], y = pcoa_coords[, 2], visit_sum = sample_data(ps_full_stool_j)$visit_sum)

# Create density plots on the x and y axes
plot_x <- ggplot(df, aes(x = x, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  #clean_theme()+
  scale_y_continuous(breaks = c(0,1.0,2.0,3.0,4.0), labels = c("0.0", "1.0", "2.0","3.0","4.0"))+
  scale_fill_manual(values = visit_sum_palette, labels = xlabels)+
  theme(legend.position = "none", text = element_text(size = 18), legend.title = element_blank())

plot_y <- ggplot(df, aes(x = y, fill = as.factor(visit_sum))) +
  geom_density(alpha = 0.5) +
  theme_classic() +
  #clean_theme()+
  #xlim(c(-0.6, 0.4))+
  scale_fill_manual(values = visit_sum_palette, labels = xlabels)+
  theme(legend.position = "none", text = element_text(size = 18), legend.title = element_blank())+
  coord_flip()

# Extract the legend. Returns a gtable
leg <- get_legend(p1)

# Convert to a ggplot and print
legend <- as_ggplot(leg)

# Combine the density plots with the PCA plot
# Arranging the plot
stool_density <- ggarrange(plot_x, NULL, p3, plot_y, 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(2, 1), heights = c(1, 2),
          common.legend = F)
stool_density 
```

# Check nutrition and stool consistency on stool beta-diversity
```{r nutrition, fig.width=12}
### plot principal component analysis by nutrition
PCA_all_visit_stool_nutrition <- plot_ordination(ps_full_stool_j, ordinate(ps_full_stool_j, "MDS"), color = "basic_nutrition", shape="project")

project_label <- c("Control","CF")

PCA_all_visit_stool_nutrition+
  geom_point(size = 4)+
  ggtitle("Variation by time point with healthy controls")+
  #scale_color_manual(values = visit_sum_palette)+
  scale_shape_manual(values = c(23,15),labels = project_label)+
  theme_bw()+
  theme(text=element_text(size=24), legend.position = "right", legend.text = element_text(size=16), legend.title = element_text(size=18))+
  guides(color = guide_legend(title = "time points"))+
   guides(shape = guide_legend(title = "CF vs Control"))+
  stat_ellipse()

### plot principal component analysis by bristol
PCA_all_visit_stool_bristol <- plot_ordination(ps_full_stool_j, ordinate(ps_full_stool_j, "MDS"), color = "bristol", shape="project")

PCA_all_visit_stool_bristol+
  geom_point(size = 4)+
  ggtitle("Variation by time point with healthy controls")+
  #scale_color_manual(values = visit_sum_palette)+
  scale_shape_manual(values = c(23,15),labels = project_label)+
  theme_bw()+
  theme(text=element_text(size=24), legend.position = "right", legend.text = element_text(size=16), legend.title = element_text(size=18))+
  guides(color = guide_legend(title = "Bristol stool scale"))+
   guides(shape = guide_legend(title = "CF vs Control"))+
  stat_ellipse()

 #calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool_j,
                             method="bray", weighted=F)
#extract metadata
metadata<- as(sample_data(ps_full_stool_j),"data.frame")
metadata$id.x

nutri <- vegan::adonis2(BC_dist ~ basic_nutrition + bristol + sex+ id.x,
              permutations = 999, na.action=na.exclude, data = metadata)
nutri
```


# PERMANOVA in stool
```{r calculate bray curtis distance for all stool samples and run PERMANOVA}

# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool_j,
                             method="bray", weighted=F)
#extract metadata
metadata<- as(sample_data(ps_full_stool_j),"data.frame")

### stratified PERMANOVA per visit group
adonis2(formula = BC_dist ~ visit_sum + sex + age_y, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)

### stratified PERMANOVA health vs CF
adonis2(formula = BC_dist ~ project + sex + age_y, data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)
```

# compare CF and control BC-distances for stool
```{r compare CF and control BC-distances stool}
# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool_j,
                             method="bray", weighted=F)
BC_dist.stool <- as.matrix(BC_dist)
#BC_dist.stool[lower.tri(BC_dist.stool)] <- 0  # this is needed to remove the 0 latter, to keep only single pairs
BC_dist.stool_df <- reshape2::melt(BC_dist.stool)

tmp1 <- BC_dist.stool_df%>%
  filter(grepl("IMMP", Var1)) %>% #this keeps only control samples in Var 1
  filter(!grepl("IMMP", Var2))#%>%  #this filters control samples from Var 2, so that in the end only distances between COntrols and all CF samples are kept
  #filter(value!=0)#this removes the same sample distances, and the one from set 0 from above to keep a single pair

# merge with metadata to have visit_sums for CF samples
#extract metadata
metadata<- as(sample_data(ps_full_stool_j),"data.frame")
tmp1 <- tmp1 %>% 
  left_join(metadata, by=c("Var2"="x_sample_id")) %>% 
  mutate(comparison=paste(Var1, Var2, sep = "_")) %>% 
  distinct(comparison, .keep_all = T)

summary(tmp1$visit_sum)

my_comp <- list(c("1","2"), c("1","3-5"), c("1","6-7"), c("1","8-10"))

# Calculate median and IQR
summary_stats <- tmp1 %>%
  group_by(visit_sum) %>%
  summarize(
    median = median(value),
    lower = quantile(value, 0.25),
    upper = quantile(value, 0.75)
  )

tmp1 %>% 
  ggplot(aes(visit_sum, value, fill=visit_sum))+
  geom_boxplot(outlier.shape = NA, alpha=0.7)+
  #geom_point()+
  theme_classic()+
  scale_fill_manual(values = visit_sum_palette)+
  ylab("BC distance: CF and Controls")+
  theme(legend.position = "none",text=element_text(size=20))+ #
  scale_x_discrete(labels= xlabels)+
  xlab("")+
  stat_compare_means(comparisons = my_comp, method = "wilcox.test", label = "p.adj")

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/BC_stool_controls.png", width = 6, height = 5)

summary(lmerTest::lmer(value~visit_sum + age_y+sex+ (1|id.x)+ (1|Var1), data = tmp1))

# below I plot the BC distance per patient and each control, with controls being on the x-axis, one can observe the consistency of BC distance per patient across controls, with visit_sum giving a distinct signal in each patient
tmp1 %>% 
  ggplot(aes(Var1, value, fill=visit_sum))+
  #geom_boxplot(outlier.shape = NA)+
  geom_point(aes(color=visit_sum), alpha=0.5)+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  ylab("BC distance between CF and Controls")+
  theme(legend.position = "none",text=element_text(size=20))+ #
  #scale_x_discrete(labels= xlabels)+
  xlab("")+
  facet_wrap(~id.x)
```

# stool linear model + boxplot
```{r stool plot boxplot with linear model age and sex correction}

lm_boxplot_function <- function(dependent_variable, ylabel) {
  
  # Fit linear mixed-effects model
  print(summary(lmerTest::lmer(paste(dependent_variable, "~ visit_sum + sex + age_y + (1|id) + (1|Var1)"), data = tmp1)))
  lm <- summary(lmerTest::lmer(paste(dependent_variable, "~ visit_sum + sex + age_y + (1|id) + (1|Var1)"), data = tmp1))
  #lm <- summary(lmerTest::lmer(value ~ visit_sum + sex + age_y + (1|id), data = tmp1))
  # Extract coefficients and adjust p-values
  coefs <- data.frame(coef(lm))
  fdr <- p.adjust(coefs$Pr...t.., method = "fdr", n = nrow(coefs))
  
  # Create lm_stats table
  lm_stats <- bind_cols(coefs, fdr) %>%
    mutate(p = Pr...t..) %>%
    mutate(fdr = ...6) %>%
    select(-c(Pr...t.., ...6)) %>%
    rownames_to_column() %>%
    mutate(Months_after_ETI_start = rowname) %>%
    mutate(Months_after_ETI_start = case_when(
      Months_after_ETI_start == "(Intercept)" ~ "Baseline (Intercept)",
      Months_after_ETI_start == "visit_sum2" ~ "3 months",
      Months_after_ETI_start == "visit_sum3-5" ~ "6-12 months",
      Months_after_ETI_start == "visit_sum6-7" ~ "15-18 months",
      Months_after_ETI_start == "visit_sum8-10" ~ "21-24 months",
      Months_after_ETI_start %in% c("sex2", "age_y") ~ Months_after_ETI_start
    )) %>%
    mutate(fdr_star = case_when(
      fdr <= 0.001 ~ "***",
      fdr <= 0.01 ~ "**",
      fdr <= 0.05 ~ "*",
      fdr <= 0.1 ~ ".",
      fdr >= 0.1 ~ "ns"
    )) %>%
    select(-rowname) %>%
    select(Months_after_ETI_start, Estimate, Std..Error, df, t.value, p, fdr, fdr_star) %>%
    mutate(p = round(p, 5)) %>%
    mutate(fdr = round(fdr, 5))
  
  # Print lm_stats table
  print(lm_stats)
  
  # Save lm_stats table as HTML
  sjPlot::tab_df(lm_stats, title = paste("lmer(", dependent_variable, " ~ visit_sum + sex + age_y + (1|id), data=tmp1)"), file = paste("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/", dependent_variable, "_table.html", sep = ""))
  
  # Print boxplot
  group1 <- c("(Intercept)", rep("1", nrow(lm_stats) - 1))
  
  lm_sig <- lm_stats %>%
    bind_cols(group1) %>%
    mutate(group1 = ...9) %>%
    mutate(group2 = case_when(
      Months_after_ETI_start == "Baseline (Intercept)" ~ "1",
      Months_after_ETI_start == "3 months" ~ "2",
      Months_after_ETI_start == "6-12 months" ~ "3-5",
      Months_after_ETI_start == "15-18 months" ~ "6-7",
      Months_after_ETI_start == "21-24 months" ~ "8-10",
      Months_after_ETI_start %in% c("sex2", "age_y") ~ Months_after_ETI_start
    )) %>%
    filter(group1 != "(Intercept)") %>%
    filter(Months_after_ETI_start!="sex2") %>% 
    filter(Months_after_ETI_start!="age_y")
  
  max_y <- max(tmp1[[dependent_variable]], na.rm = TRUE)
  min_y <- min(tmp1[[dependent_variable]], na.rm = TRUE)
  
 # Calculate the adjusted y.position for stat labels
stat_labels_y_position <- max_y + 0.1 * (max_y - min_y)

  xlabels <- c("0", "3", "6-12", "15-18", "21-24")
  
# Print boxplot
boxplot <- tmp1 %>%
  ggplot(aes(get(dependent_variable), x = visit_sum)) +
  geom_boxplot(aes(fill = visit_sum), outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = visit_sum_palette) +
  theme_classic() +
  ylab(ylabel) +
  xlab("Months from ETI treatment start") +
  scale_x_discrete(labels = xlabels) +
  theme(text = element_text(size = 18), legend.position = "none") +
   stat_pvalue_manual(lm_sig, label = "fdr_star", y.position = stat_labels_y_position, step.increase = 0.05)

return(list(lm_stats = lm_stats, boxplot = boxplot))
}

stool <- lm_boxplot_function("value", "Stool: BC distance - CF & Control")
p4 <- stool$boxplot
stool$lm_stats
```
```{r plot linear model coefficients stool}
# Fit your linear mixed-effects model
# For example, replace 'YourModel' and 'YourData' with your model and data
model <-  lmerTest::lmer(value ~ visit_sum + sex + age_y + (1|id) + (1|Var1), data = tmp1)

# Extract coefficients
coefficients <- coef(model)

# Extract coefficients
coefficients <- coef(model)

# Plot coefficients
  coef_p <- coefplot::coefplot(model, intercept = FALSE, color = "black")
  coef_p_s<- coef_p+
    theme_classic()+
    scale_y_discrete(labels= c("3 m", "6-12 m", "15-18 m", "21-24 m", "sex", "age"))+
    coord_flip()+
    xlab("Estimate")+
    ggtitle("")+
    ylab("")+
    theme(text = element_text(size = 18))
  coef_p_s
```

# combine plots
```{r, fig.width=16, fig.height=18}
library(gridExtra)
p1 <- p1+
  ggtitle("Throat")

p3 <- p3+
  ggtitle("Stool")

grid_arranged_t <- grid.arrange(
  p1+ggtitle(""),
  NULL,
  plot_x_th+theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks = element_blank())+ggtitle("Throat")+labs(tag = "A") ,
  plot_y_th+theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks = element_blank()),
  p3+ggtitle(""),
  NULL,
  plot_x+theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks = element_blank())+ggtitle("Stool")+labs(tag = "D") ,
  plot_y+theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks = element_blank()),
  coef_p_t + labs(tag = "B"),
  p2 + labs(tag = "C"),
  coef_p_s + labs(tag = "E"),
  p4 + labs(tag = "F"),
  legend,
  ncol = 3, nrow = 5,
  heights = c(1, 2, 1,2, 0.5),
  widths=c(2,0.5,1.5),
  layout_matrix = rbind(c(3,2,9),
                        c(1,4,10),
                        c(7,6,11),
                        c(5,8,12),
                        c(13,13,13)
                      )
)

# Print or save the arranged plot
ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/Submission_CHM/Fig3.tiff",grid_arranged_t, dpi = 600, width = 16, height = 18)

# combine stats table

throat$lm_stats <- throat$lm_stats %>% 
  mutate(Sample="Throat") %>% 
  select(Sample, everything())

stool$lm_stats <- stool$lm_stats %>% 
  mutate(Sample="Stool") %>% 
  select(Sample, everything())

bc_stats <- rbind(throat$lm_stats, stool$lm_stats)
kable(bc_stats)
#write_csv(bc_stats, "/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/lmm_bc_distance_control.csv")
```
```{r}
sessionInfo()
```

