---
title: "Mediation analysis"
author: "Rebecca L. Knoll"
date: "last edit `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    code_folding: show
    number_sections: yes
    smart: no
    toc: yes
    df_print: paged
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

pacman::p_load(rmarkdown, tidyverse, phyloseq, magrittr, janitor, microbiome, knitr, lubridate, naniar,  ggplot2, microViz)
#install.packages("mediation")
library(mediation) #Mediation package
library(rockchalk) #Graphing simple slopes; moderation
library(multilevel) #Sobel Test
library(bda) #Another Sobel Test option
library(gvlma) #Testing Model Assumptions 
library(stargazer) #Handy regression tables
```
# load data
```{r}
ps_full <- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/ps_IMProveCF_patients_Run1-23_18102023.rds")

ps_full_sputum <- subset_samples(ps_full, material== "Sputum")
ps_full_stool <- subset_samples(ps_full, material== "Stool")
ps_full_throat <- subset_samples(ps_full, material== "Throat")

ps_full_sputum <- tax_filter(ps_full_sputum, min_prevalence = 1) # has 632 taxa

set.seed(100)
ps_full_sputum_rar <- rarefy_even_depth(ps_full_sputum, rngseed=FALSE, sample.size=8000, replace=T, trimOTUs = T, verbose = T) # has 597 taxa
ps_sputum_filtered <- tax_filter(ps_full_sputum, min_prevalence = 0.25) # has 101 taxa
ps_sputum_filtered_rar <- tax_filter(ps_full_sputum_rar, min_prevalence = 0.25) # has 39 taxa

ps_full_throat <- tax_filter(ps_full_throat, min_prevalence = 1) # removes 0s, has 2237 taxa
ps_full_throat_rar <- rarefy_even_depth(ps_full_throat, rngseed=FALSE, sample.size=5500, replace=T, trimOTUs = T, verbose = T)
ps_throat_filtered <- tax_filter(ps_full_throat, min_prevalence = 0.25) # has to be prevalent in 20% of samples = 107 taxa
ps_throat_filtered_rar <- tax_filter(ps_full_throat_rar, min_prevalence = 0.25) # has 70 taxa


ps_full_stool <- tax_filter(ps_full_stool, min_prevalence = 1) # removes 0s, has 2341 taxa
ps_full_stool_rar <- rarefy_even_depth(ps_full_stool, rngseed=FALSE, sample.size=6500, replace=T, trimOTUs = T, verbose = T) # has 1935 taxa
ps_stool_filtered <- tax_filter(ps_full_stool, min_prevalence = 0.25) # has to be prevalent in 20% of samples = 124 taxa
ps_stool_filtered_rar <- tax_filter(ps_full_stool_rar, min_prevalence = 0.25) # has to be prevalent in 20% of samples = 123 taxa

ps_full_relab <- transform_sample_counts(ps_full, function(x) x/sum(x))
ps_sputum_filtered_relab <- transform_sample_counts(ps_sputum_filtered, function(x) x/sum(x))
ps_throat_filtered_relab <- transform_sample_counts(ps_throat_filtered, function(x) x/sum(x))
ps_stool_filtered_relab <- transform_sample_counts(ps_stool_filtered, function(x) x/sum(x))

# extract Staphylococcus abundance from Sputum

sputum_df <- psmelt(ps_sputum_filtered_relab)
sputum_df <- sputum_df %>% 
  mutate(Abundance=Abundance*100) # transform relative abundance into percentages

staph_df <- sputum_df%>% 
  filter(Genus=="Staphylococcus")

tabyl(staph_df, id.x,visit_cal_cor) # summary of sputum samples and to check that there is only 1 sample per patient per visit
```
# Mediation analysis on Staphylococcus abundance mediated by sweat chloride, dependent on ETI treatment duration (days) using`JSmediation` package, as in Morgan Essex Gespic Paper

```{r}
# Install and load necessary packages
#install.packages("JSmediation")
#install.packages("lme4")
library(JSmediation)
library(lme4)

# Fit the mixed-effects models
fitM_staph <- lmer(chlorid_mmol_cl_l ~ delta_ETI_merged + sex + age_y + (1 | id.x), data = staph_df)
fitY_staph <- lmer(Abundance ~ delta_ETI_merged + chlorid_mmol_cl_l + sex + age_y + (1 | id.x), data = staph_df)

# Perform the mediation analysis using mdt_simple
mediation_result <- mdt_simple(staph_df, # data set
                               delta_ETI_merged, # predictor
                               Abundance, # outcome
                               chlorid_mmol_cl_l) # mediator
mediation_result
med_output <- add_index(mediation_result)
med_output$type
med_output$method
med_output$params
med_output$paths
med_output$indirect_index
med_output$indirect_index_infos
med_output$js_models

# Summarize the mediation analysis
summary(mediation_result)

```






Interpretation of mediation analysis using the `mdt_simple()` function from the `JSmediation` package, lets break down the key components:

### Variables:
- **IV (Independent Variable)**: `delta_ETI_merged`
- **DV (Dependent Variable)**: `Abundance`
- **M (Mediator)**: `chlorid_mmol_cl_l`

### Paths:
- **Path a (IV to Mediator)**:
  - Point Estimate: -0.07241581
  - Standard Error: 0.01501267
  - t(31) = 4.82, p < .001
  - Interpretation: There is a significant negative effect of `delta_ETI_merged` on `chlorid_mmol_cl_l`. This means that as `delta_ETI_merged` increases, `chlorid_mmol_cl_l` decreases significantly.

- **Path b (Mediator to DV, controlling for IV)**:
  - Point Estimate: 0.331141
  - Standard Error: 0.1500046
  - t(30) = 2.21, p = .035
  - Interpretation: There is a significant positive effect of `chlorid_mmol_cl_l` on `Abundance`, controlling for `delta_ETI_merged`. This means that as `chlorid_mmol_cl_l` increases, `Abundance` increases significantly.

- **Path c (Total effect of IV on DV)**:
  - Point Estimate: -0.02906967
  - Standard Error: 0.007989211
  - t(55) = 3.64, p < .001
  - Interpretation: There is a significant negative total effect of `delta_ETI_merged` on `Abundance`. This means that as `delta_ETI_merged` increases, `Abundance` decreases significantly.

- **Path c (Direct effect of IV on DV, controlling for Mediator)**:
  - Point Estimate: -0.0097406
  - Standard Error: 0.01658947
  - t(30) = 0.59, p = .561
  - Interpretation: The direct effect of `delta_ETI_merged` on `Abundance`, controlling for `chlorid_mmol_cl_l`, is not significant. This suggests that the relationship between `delta_ETI_merged` and `Abundance` is largely mediated by `chlorid_mmol_cl_l`.

### Indirect Effect:
- **Indirect effect (a * b)**:
  - Point Estimate: -0.024
  - Confidence Interval (95%): [-0.0498, -0.00286]
  - Method: Monte Carlo (5000 iterations)
  - Interpretation: The indirect effect of `delta_ETI_merged` on `Abundance` through `chlorid_mmol_cl_l` is significant (since the confidence interval does not include zero). This indicates that `chlorid_mmol_cl_l` significantly mediates the relationship between `delta_ETI_merged` and `Abundance`.

### Summary:
- **Significant Indirect Effect**: The mediation analysis reveals that `chlorid_mmol_cl_l` significantly mediates the effect of `delta_ETI_merged` on `Abundance`. This is supported by the significant paths a and b, as well as the significant indirect effect.
- **Non-significant Direct Effect**: The direct effect of `delta_ETI_merged` on `Abundance`, controlling for `chlorid_mmol_cl_l`, is not significant. This suggests that the effect of `delta_ETI_merged` on `Abundance` is primarily through its effect on `chlorid_mmol_cl_l`.

### Reporting the Results

The mediation analysis examined whether `chlorid_mmol_cl_l` mediated the relationship between `delta_ETI_merged` and `Abundance`. 

The analysis revealed a significant negative total effect of `delta_ETI_merged` on `Abundance` (c path: β = -0.029, SE = 0.008, t(55) = 3.64, p < .001). The direct effect of `delta_ETI_merged` on `Abundance`, controlling for `chlorid_mmol_cl_l`, was not significant (c`'` path: β = -0.010, SE = 0.017, t(30) = 0.59, p = .561).

The effect of `delta_ETI_merged` on `chlorid_mmol_cl_l` was significant and negative (a path: β = -0.072, SE = 0.015, t(31) = 4.82, p < .001). Additionally, the effect of `chlorid_mmol_cl_l` on `Abundance`, controlling for `delta_ETI_merged`, was significant and positive (b path: β = 0.331, SE = 0.150, t(30) = 2.21, p = .035).

The indirect effect of `delta_ETI_merged` on `Abundance` through `chlorid_mmol_cl_l` was significant (β = -0.024, 95% CI [-0.0498, -0.00286]), indicating that `chlorid_mmol_cl_l` significantly mediated the relationship between `delta_ETI_merged` and `Abundance`.

These results suggest that the impact of `delta_ETI_merged` on `Abundance` is largely mediated by changes in `chlorid_mmol_cl_l`, highlighting the important role of `chlorid_mmol_cl_l` in this relationship.

# Mediation analysis with the mediate package on Staphylococcus abundance mediated by sweat chloride, dependent on ETI treatment duration (days)
```{r}

# perform mediation analysis for Staphylococcus
fitM_staph <- lm(staph_df$chlorid_mmol_cl_l ~ staph_df$delta_ETI_merged + staph_df$sex + staph_df$age_y)   ####### M: sweat chloride predicting X: tt time
fitY_staph <- lm(staph_df$Abundance ~ staph_df$delta_ETI_merged + staph_df$chlorid_mmol_cl_l + staph_df$sex + staph_df$age_y) 
                 ###### X: tt time and M: sweat chloride predicting Y:Staph abundance

summary(fitM_staph)
summary(fitY_staph)
gvlma(fitM_staph)
gvlma(fitY_staph)

fitMed_staph <- mediate(fitM_staph, fitY_staph, treat="staph_df$delta_ETI_merged", mediator="staph_df$chlorid_mmol_cl_l")
summary(fitMed_staph)
plot(fitMed_staph)

```
Mediation Analysis Overview

In a mediation analysis, we are interested in understanding whether the effect of an independent variable (treatment) on a dependent variable (outcome) is mediated through a third variable (mediator).

Here, the variables are:

    Treatment (X): staph_df$delta_ETI_merged (tt time)
    Mediator (M): staph_df$chlorid_mmol_cl_l (sweat chloride)
    Outcome (Y): staph_df$Abundance (Staphylococcus abundance)

Fit Models

    Mediator Model (fitM_staph):
    Mediator (M)=β0+β1⋅Treatment (X)+β2⋅sex+β3⋅age
    Mediator (M)=β0​+β1​⋅Treatment (X)+β2​⋅sex+β3​⋅age

    This model examines how the treatment (tt time) affects the mediator (sweat chloride).

    Outcome Model (fitY_staph):
    Outcome (Y)=γ0+γ1⋅Treatment (X)+γ2⋅Mediator (M)+γ3⋅sex+γ4⋅age
    Outcome (Y)=γ0​+γ1​⋅Treatment (X)+γ2​⋅Mediator (M)+γ3​⋅sex+γ4​⋅age

    This model examines how both the treatment (tt time) and the mediator (sweat chloride) affect the outcome (Staphylococcus abundance).

Mediation Analysis Results

The mediate function combines the results from these models to decompose the total effect of the treatment on the outcome into:

    ACME (Average Causal Mediation Effect): The effect of the treatment on the outcome that is mediated through the mediator.
    ADE (Average Direct Effect): The effect of the treatment on the outcome that is not mediated through the mediator.
    Total Effect: The combined effect of ACME and ADE.
    Proportion Mediated: The proportion of the total effect that is mediated through the mediator.

Interpretation of Results

    ACME: Estimate = -0.02675, 95% CI = [-0.05391, 0.00], p-value = 0.022
        The average causal mediation effect is significant at the 5% level (p-value < 0.05), suggesting that the treatment's effect on the outcome is mediated through the sweat chloride levels.

    ADE: Estimate = -0.00621, 95% CI = [-0.04021, 0.03], p-value = 0.720
        The average direct effect is not significant (p-value > 0.05), indicating that the treatment does not have a significant direct effect on the outcome when controlling for the mediator.

    Total Effect: Estimate = -0.03296, 95% CI = [-0.05993, 0.00], p-value = 0.014
        The total effect is significant at the 5% level (p-value < 0.05), indicating that the treatment has a significant overall effect on the outcome.

    Proportion Mediated: Estimate = 0.80567, 95% CI = [0.08188, 3.86], p-value = 0.036
        The proportion of the total effect that is mediated through sweat chloride is significant, with about 80.57% of the total effect being mediated.

Conclusion

The results suggest that the effect of tt time on Staphylococcus abundance is largely mediated through sweat chloride levels. This means that changes in sweat chloride due to the treatment (tt time) play a crucial role in influencing Staphylococcus abundance. The direct effect of the treatment on Staphylococcus abundance, independent of sweat chloride levels, is not significant.

# revers the mediation effect for confirmation
```{r}
# perform mediation analysis for Staphylococcus
fitM_staph_r <- lm(staph_df$delta_ETI_merged ~ staph_df$chlorid_mmol_cl_l + staph_df$sex + staph_df$age_y)   ####### M: sweat chloride predicting X: tt time
fitY_staph <- lm(staph_df$Abundance ~ staph_df$delta_ETI_merged + staph_df$chlorid_mmol_cl_l + staph_df$sex + staph_df$age_y) 
                 ###### X: tt time and M: sweat chloride predicting Y:Staph abundance

summary(fitM_staph_r)
summary(fitY_staph)
gvlma(fitM_staph_r)
gvlma(fitY_staph)

fitMed_staph_r <- mediate(fitM_staph_r, fitY_staph, treat= "staph_df$chlorid_mmol_cl_l", mediator= "staph_df$delta_ETI_merged")
summary(fitMed_staph_r)
plot(fitMed_staph)
```

# add random effects for repeated measures:

```{r}
# perform mediation analysis for Staphylococcus
fit.totaleffect <- lmer(Abundance ~ delta_ETI_merged + (1|id.x) + sex + age_y, data = staph_df) # this is computed for all sputum samples (N=57) and not truely needed for the subsequent analysis
fit.mediator <- lmer(chlorid_mmol_cl_l ~ delta_ETI_merged + (1|id.x) + sex + age_y, data = staph_df)
fitY<- lmer(Abundance ~ delta_ETI_merged + (1|id.x) + chlorid_mmol_cl_l + sex + age_y, data = staph_df) 

summary(fit.totaleffect)
summary(fit.mediator)
summary(fitY)

fitMed_staph <- mediate(fit.mediator, fitY, treat= "delta_ETI_merged", mediator=  "chlorid_mmol_cl_l")
summary(fitMed_staph)
plot(fitMed_staph)
```
# function for mediation analysis with repeated measures
```{r}
library(lme4)
library(mediation)

# Define the function for mediation analysis
perform_mediation_analysis <- function(data, mediator, outcome, predictor, id_var, covariates) {
  # Construct the covariates string
  covariates_str <- paste(covariates, collapse = " + ")
  
  # Construct formulas for the models
  mediator_formula <- as.formula(paste(mediator, "~", predictor, "+ (1|", id_var, ") +", covariates_str))
  outcome_formula <- as.formula(paste(outcome, "~", predictor, "+ (1|", id_var, ") +", mediator, "+", covariates_str))
  
  # Fit the models
  fit.mediator <- lmer(mediator_formula, data = data)
  fitY <- lmer(outcome_formula, data = data)
  
  # Print summaries of the models
  print(summary(fit.mediator))
  print(summary(fitY))
  
  # Perform mediation analysis
  fitMed <- mediate(fit.mediator, fitY, treat = predictor, mediator = mediator)
  
  # Print summary and plot the results
  print(summary(fitMed))
  plot(fitMed)
  
  return(fitMed)
}
```

# Staphylococcus abundance in Sputum, mediated by sweat chloride, corrected for id as repeated measure
```{r}
# Define the dataset and variables
data <- staph_df
mediator <- "chlorid_mmol_cl_l"
outcome <- "Abundance"
predictor <- "delta_ETI_merged"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis
fitMed_result_staph <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

summary(fitMed_result_staph)
```
# Staphylococcus abundance in Sputum, mediated by sweat chloride, corrected for id as repeated measure
```{r}
# Define the dataset and variables
data <- staph_df
mediator <- "chlorid_mmol_cl_l"
outcome <- "Abundance"
predictor <- "delta_ETI_merged"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis
fitMed_result_staph <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

summary(fitMed_result_staph)
```

# reversed
```{r}
# Define the dataset and variables
data <- staph_df
mediator <-  "delta_ETI_merged"
outcome <- "Abundance"
predictor <- "chlorid_mmol_cl_l"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis
fitMed_result_staph_r <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

summary(fitMed_result_staph_r)
```

# Test sputum diversity in mediation
```{r}
alphadiversity_df_sputum <- microbiome::alpha(ps_full_sputum_rar, index = "all")
alphadiversity_df_sputum$x_sample_id <- rownames(alphadiversity_df_sputum)
alphadiversity_df_sputum_sel <- alphadiversity_df_sputum %>% 
  dplyr::select(diversity_shannon, observed, evenness_pielou, dominance_dbp, x_sample_id)

ps_full_sputum_df <- as.data.frame(sample_data(ps_full_sputum))

diversity_df_sputum_metadata<- alphadiversity_df_sputum%>%
left_join(ps_full_sputum_df, by = "x_sample_id")

summary(diversity_df_sputum_metadata$diversity_shannon)
# Define the dataset and variables
data <- diversity_df_sputum_metadata
mediator <- "chlorid_mmol_cl_l"
outcome <- "diversity_shannon"
predictor <- "delta_ETI_merged"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis for shannon and sweat chloride
fitMed_result_shannon <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for observed and sweat chloride
outcome <- "observed"
fitMed_result_obs <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for diversity and levofloxacin
diversity_df_sputum_metadata <- diversity_df_sputum_metadata %>% 
  mutate(levofloxacin_inhal=case_when(levofloxacin_inhal==0~0, levofloxacin_inhal==2~1))
summary(diversity_df_sputum_metadata$levofloxacin_inhal)
data <- diversity_df_sputum_metadata
outcome <- "diversity_shannon"
mediator <- "levofloxacin_inhal"
fitMed_result_shannon_levofloxacin <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for richness and levofloxacin
outcome <- "observed"
mediator <- "levofloxacin_inhal"
fitMed_result_richness_levofloxacin <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for richness and levofloxacin, predictor: lung function
outcome <- "diversity_shannon"
mediator <- "levofloxacin_inhal"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y")
fitMed_result_diversity_levofloxacin <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for richness and levofloxacin, predictor: lung function
outcome <- "observed"
mediator <- "levofloxacin_inhal"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y")
fitMed_result_richness_levofloxacin <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for richness and levofloxacin, predictor: lung function, include treatment days as an additional fixed effect
outcome <- "diversity_shannon"
mediator <- "levofloxacin_inhal"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y", "delta_ETI_merged")
fitMed_result_diversity_levofloxacin_eti <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for richness and levofloxacin, predictor: lung function, include treatment time as a fixed effect
outcome <- "observed"
mediator <- "levofloxacin_inhal"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y", "delta_ETI_merged")
fitMed_result_richness_levofloxacin_eti <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# test lung function as predictor of diversity independent of levofloxacin
model <- lmer(diversity_shannon ~ pp_fev_percent + (1|id.x) + sex + age_y, data = diversity_df_sputum_metadata)

# in metadeconfoundR the confounder of this is sweat chloride concentration, so I rerun mediation analysis inspecting: diversity (outcome) ~ lung function (predictor) + sweat chloride (mediator)
outcome <- "diversity_shannon"
mediator <- "chlorid_mmol_cl_l"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y")
fitMed_result_diversity_chloride <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

outcome <- "observed"
mediator <- "chlorid_mmol_cl_l"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y")
fitMed_result_richness_chloride <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# next I include sweat chloride as a covariate, as it has been identified as a confounder, thus should be included in the model
outcome <- "diversity_shannon"
mediator <- "levofloxacin_inhal"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y","chlorid_mmol_cl_l")
fitMed_result_diversity_levo_chloride <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Rerun for richness (richness prediced from lung function has no confounders in metadeconfoundR reported, as it was also not significant)
outcome <- "observed"
mediator <- "levofloxacin_inhal"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y")
fitMed_result_richness_levo <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# chloride as mediator
outcome <- "observed"
mediator <- "chlorid_mmol_cl_l"
predictor <- "pp_fev_percent"
covariates <- c("sex", "age_y")
fitMed_result_richness_chloride <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

```
# Stool
## investigating the effect of total number of antibiotic treatment days on alpha diversity
```{r}
alphadiversity_df_stool <- microbiome::alpha(ps_full_stool_rar, index = "all")
alphadiversity_df_stool$x_sample_id <- rownames(alphadiversity_df_stool)
alphadiversity_df_stool_sel <- alphadiversity_df_stool %>% 
  dplyr::select(diversity_shannon, observed, evenness_pielou, dominance_dbp, x_sample_id)

ps_full_stool_df <- as.data.frame(sample_data(ps_full_stool))

diversity_df_stool_metadata<- alphadiversity_df_stool%>%
left_join(ps_full_stool_df, by = "x_sample_id")

summary(diversity_df_stool_metadata$diversity_shannon)
# Define the dataset and variables
data <- diversity_df_stool_metadata
mediator <- "chlorid_mmol_cl_l"
outcome <- "diversity_shannon"
predictor <- "delta_ETI_merged"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis for shannon and sweat chloride
fitMed_result_shannon <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for observed and sweat chloride
outcome <- "observed"
fitMed_result_obs <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for diversity and antibiotic treatment days
summary(diversity_df_stool_metadata$treatmentdays_365prior_visit)
data <- diversity_df_stool_metadata
outcome <- "diversity_shannon"
mediator <- "treatmentdays_365prior_visit"
covariates <- c("sex", "age_y")

fitMed_result_shannon_atb <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for richness and antibiotic treatment days
outcome <- "observed"
mediator <- "treatmentdays_365prior_visit"
fitMed_result_richness_levofloxacin <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)
```
```{r}

# extract Escherichia abundance from Stool

stool_df <- psmelt(ps_stool_filtered_relab)
stool_df <- stool_df %>% 
  mutate(Abundance=Abundance*100) # transform relative abundance into percentages

esch_df <- stool_df%>% 
  filter(Genus=="Escherichia_Shigella")

tabyl(esch_df, id.x,visit_cal_cor) # summary of stool samples and to check that there is only 1 sample per patient per visit

# Define the dataset and variables
data <- esch_df
mediator <- "calprotectin_amg_g_stuhl"
outcome <- "Abundance"
predictor <- "delta_ETI_merged"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis 
fitMed_result_esch_calp <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Define the dataset and variables
data <- esch_df
mediator <- "delta_ETI_merged"
outcome <- "Abundance"
predictor <- "calprotectin_amg_g_stuhl"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis 
fitMed_result_reversed <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Define the dataset and variables
data <- esch_df
mediator <- "calprotectin_amg_g_stuhl"
outcome <- "Abundance"
predictor <- "BMI"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis 
fitMed_result_reversed <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Define the dataset and variables
data <- esch_df
mediator <- "BMI" 
outcome <- "Abundance"
predictor <- "calprotectin_amg_g_stuhl"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis 
fitMed_result_reversed <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

summary(lmerTest::lmer(Abundance~BMI+calprotectin_amg_g_stuhl+age_y+sex+(1|id.x), data = esch_df))
summary(lmerTest::lmer(Abundance~log10(calprotectin_amg_g_stuhl)+age_y+sex+(1|id.x), data = esch_df))


# Define the dataset and variables
data <- esch_df
mediator <- "BMI"
outcome <- "Abundance"
predictor <- "delta_ETI_merged"
id_var <- "id.x"
covariates <- c("sex", "age_y")

# Perform the mediation analysis 
fitMed_result_reversed <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for PPI
mediator <- "ppi"
fitMed_result_esch_ppi <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)

# Perform the mediation analysis for PPI on calprotectin
mediator <- "ppi"
outcome <- "Abundance"
predictor <- "calprotectin_amg_g_stuhl"
id_var <- "id.x"
covariates <- c("sex", "age_y")
fitMed_result_esch_ppi_calp <- perform_mediation_analysis(data, mediator, outcome, predictor, id_var, covariates)
```



