---
title: "MetadeconfoundR full data set"
author: "Rebecca Luise Knoll"
date: "last edit `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    code_folding: show
    number_sections: yes
    smart: no
    toc: yes
    df_print: paged
    toc_float: yes
---
```{r setup 2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r run import script, include=FALSE}
pacman::p_load(rmarkdown, tidyverse, phyloseq, magrittr, janitor, microbiome, knitr, lubridate, naniar, readxl, ggplot2, ggpubr,rstatix, metadeconfoundR, microViz)

#renv::install("tillbirkner/MetadeconfoundR")
ps_full <- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/ps_IMProveCF_patients_Run1-23_18102023.rds")

samples <- read_csv("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/data/samples_sequencedTF.csv")

```

```{r subset per sample type}

ps_full_sputum <- subset_samples(ps_full, material== "Sputum")
ps_full_stool <- subset_samples(ps_full, material== "Stool")
ps_full_throat <- subset_samples(ps_full, material== "Throat")

ps_full_sputum <- tax_filter(ps_full_sputum, min_prevalence = 1) # has 632 taxa

set.seed(100)
ps_full_sputum_rar <- rarefy_even_depth(ps_full_sputum, rngseed=FALSE, sample.size=8000, replace=T, trimOTUs = T, verbose = T) # has 597 taxa
ps_sputum_filtered <- tax_filter(ps_full_sputum, min_prevalence = 0.25) # has 101 taxa
ps_sputum_filtered_rar <- tax_filter(ps_full_sputum_rar, min_prevalence = 0.25) # has 39 taxa

ps_full_throat <- tax_filter(ps_full_throat, min_prevalence = 1) # removes 0s, has 2237 taxa
ps_full_throat_rar <- rarefy_even_depth(ps_full_throat, rngseed=FALSE, sample.size=5500, replace=T, trimOTUs = T, verbose = T)
ps_throat_filtered <- tax_filter(ps_full_throat, min_prevalence = 0.25) # has to be prevalent in 20% of samples = 107 taxa
ps_throat_filtered_rar <- tax_filter(ps_full_throat_rar, min_prevalence = 0.25) # has 70 taxa


ps_full_stool <- tax_filter(ps_full_stool, min_prevalence = 1) # removes 0s, has 2341 taxa
ps_full_stool_rar <- rarefy_even_depth(ps_full_stool, rngseed=FALSE, sample.size=6500, replace=T, trimOTUs = T, verbose = T) # has 1935 taxa
ps_stool_filtered <- tax_filter(ps_full_stool, min_prevalence = 0.25) # has to be prevalent in 20% of samples = 124 taxa
ps_stool_filtered_rar <- tax_filter(ps_full_stool_rar, min_prevalence = 0.25) # has to be prevalent in 20% of samples = 123 taxa

ps_full_relab <- transform_sample_counts(ps_full, function(x) x/sum(x))
ps_sputum_filtered_relab <- transform_sample_counts(ps_sputum_filtered, function(x) x/sum(x))
ps_throat_filtered_relab <- transform_sample_counts(ps_throat_filtered, function(x) x/sum(x))
ps_stool_filtered_relab <- transform_sample_counts(ps_stool_filtered, function(x) x/sum(x))

```

# MetadeconfoundR on alpha diversity
## MetadeconfoundR for Sputum diversity
```{r MetadeconfoundR for Sputum diversity unrarefied, fig.height=3, fig.width=5.5}
alphadiversity_df_sputum <- microbiome::alpha(ps_full_sputum_rar, index = "all")
alphadiversity_df_sputum$x_sample_id <- rownames(alphadiversity_df_sputum)
alphadiversity_df_sputum_sel <- alphadiversity_df_sputum %>% 
  select(diversity_shannon, observed, evenness_pielou, dominance_dbp, x_sample_id)

ps_full_sputum_df <- as.data.frame(sample_data(ps_full_sputum))

view(ps_full_sputum_df)
diversity_df_sputum_metadata<- alphadiversity_df_sputum%>%
left_join(ps_full_sputum_df, by = "x_sample_id")

xlabels <- c("0", "3", "6-12", "15-18", "21-24")

# create metadata, to make MetadeconfoundR run properly, I need to recode all variables I want to input into MetadeconfoundR into numeric variables and I will select only the relevant variables to reduce computational time

metadata_x_sputum_selected <- diversity_df_sputum_metadata %>%
  select(sex, x_sample_id, id.y, mutation_status, age_y, BMI, azithromycin_oral, levofloxacin_inhal, colistin_inhal, dpi_colistin, tobramycin_inhal, azetronam_inhal, hypertonic_na_cl, isotonic_na_cl, anticholinergic, laba_inhal, saba_inhal, dn_ase, mannitol, cortison_inhal, cortison_nasal, cortison_oral, antihistaminikum, nasenspray, na_cl_oral,  ursodesoxychols_a_ure, ppi, orale_zusatznahrung, tranexams_a_ure, nsaid, laxantien, acetylcystein, insulin, antidiabetic_oral, furosemid, ssri, ab_15dprior_visit, treatmentdays_365prior_visit, chlorid_mmol_cl_l, hb_a1c_percent, crp_mg_l, interleukin6_pg_ml, interleukin8_pg_ml, ig_g_g_l, calprotectin_amg_g_stuhl, leukozyten_nl, gpt_alat_u_l, got_asat_u_l, pp_fev_percent, pp_fvc_percent, cftr_prior_v1, total_reads, delta_ETI_merged, bristol, basic_nutrition, staph_pos_throat, staph_pos_sputum, pseudo_pos_sputum, pseudo_pos_throat) 


xlabels= c("tobramycin_inhal" = "Tobramycin Inhalation", "isotonic_na_cl"="Isotonic NaCl Inhalation", "saba_inhal"="(SABA) Inhaler", "cortison_inhal" = "Cortisone Inhalation", "antihistaminikum"="Oral Antihistamine", "orale_zusatznahrung"="Oral Nutritional Supplement", "laxantien"= "Laxatives", "antidiabetic_oral"="Antidiabetic Medication (Oral)", "ab_15dprior_visit"="Antibiotic 15 Days before Visit", "hb_a1c_percent"="Hb A1c %", "interleukin8_pg_ml"="Interleukin-8 (pg/ml)", "leukozyten_nl"="Leukocyte Count /nl", "pp_fev_percent"="Lung function (ppFEV1)", "total_reads"="Total Reads", "sex"="Sex", "age_y"="Age", "levofloxacin_inhal"="Levofloxacin Inhalation", "azetronam_inhal"="Aztreonam Inhalation", "anticholinergic"="Anticholinergic Inhalation", "dn_ase"="DNAse Inhalation", "cortison_nasal"="Cortisone Nasal Spray", "nasenspray"= "Nasal Spray", "ursodesoxychols_a_ure"="Ursodeoxycholic Acid", "tranexams_a_ure"="Tranexamic Acid", "acetylcystein"="Acetylcysteine", "furosemid"="Furosemide", "treatmentdays_365prior_visit"="N Antibiotic tt Days", "crp_mg_l"="CRP mg/l", "ig_g_g_l"="IgG g/l", "gpt_alat_u_l"="GPT/ALAT U/l", "pp_fvc_percent"= "Lung function (ppFVC)", "delta_ETI_merged"="N ETI tt Days", "id.y"="ID", "BMI"="BMI", "colistin_inhal"="Colistin Inhalation",
"hypertonic_na_cl"="Hypertonic Saline Inhalation", "laba_inhal"= "Long-Acting Beta-Agonist (LABA) Inhaler", "mannitol"="Mannitol Inhalation", "cortison_oral"= "Cortisone", "na_cl_oral"="Sodium suppl.", "ppi"="PPI", "nsaid"="NSAIDs", "insulin"="Insulin", "ssri"="SSRIs", "chlorid_mmol_cl_l"="Sweat chloride (mmol/l)", "interleukin6_pg_ml"="Interleukin-6 (pg/ml)", "calprotectin_amg_g_stuhl"= "Calprotectin (μg/g Stool)", "got_asat_u_l"="GOT/ASAT U/l", "cftr_prior_v1"="prior CFTR-modulator tt", "mutation_status"= "Mutation (homo- vs heterozygous", "azithromycin_oral"="Azithromycin (Oral)", "dpi_colistin" =  "DPI Colistin Inhalation", "bristol"="Bristol Score", "basic_nutrition"="Basic nutrition", "staph_pos_throat"= "Staphylococcus positive throat", "staph_pos_sputum" = "Staphylococcus positive sputum", "pseudo_pos_sputum"= "Pseudomonas positive sputum", "pseudo_pos_throat" = "Pseudomonas positive throat")

# create a dataframe for the investigated metadata
# Split each string at the equal sign
split_values <- strsplit(xlabels, "=")

# Create a data frame with two columns
metaVariable_df <- data.frame(
  Column1 = sapply(split_values, `[`, 1),
  Column2 = sapply(split_values, `[`, 2)
)

write_csv(metaVariable_df, "/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/metaVariables_metadeconfoundR.csv")

ylabels<- c("evenness_pielou"="Evenness (Pielou)", "diversity_shannon"="Diversity (Shannon)", "dominance_dbp" ="Dominance (BP)", "observed"="Observed ASVs")

metadata <- metadata_x_sputum_selected
rownames(metadata) <- metadata$x_sample_id
metadata$insulin[metadata$x1 == "18IMP16V1Spu"]<-"0"
glimpse(metadata)
metadata <- metadata %>%
  mutate(across(where(is.character), as.numeric))

metadata$sex<- metadata$sex-1

# exclude variables which only represent 1 patient
metadata <- metadata %>%
  select(-c(x_sample_id))#, allopurinol, azithromycin_oral, nmh_enoxaparin, immunsuppressiva, na_ssa,dopaminagonist))

#create other input
features <- alphadiversity_df_sputum_sel %>% 
  select(-x_sample_id)


library(metadeconfoundR)

sputum_metad_div<- MetaDeconfound(featureMat = features, metaMat = metadata, randomVar = c("id.y"), nnodes=9, returnLong = T)

view(sputum_metad_div)

view(sputum_metad_div %>% 
  filter(Qs<=0.1&feature=="diversity_shannon") )

view(sputum_metad_div %>% 
    filter(Qs<=0.1) %>% 
    filter(abs(Ds)>=0.33)%>% 
    filter(status!="NS"))

tab_sputum <- BuildHeatmap(sputum_metad_div,intermedData = T,showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.2, keepMeta = c("chlorid_mmol_cl_l", "delta_ETI_merged", "interleukin6_pg_ml", "levofloxacin_inhal", "chlorid_mmol_cl_l","crp_mg_l", "leukozyten_nl", "pp_fev_percent", "delta_ETI_merged", "cortison_oral", "total_reads", "treatmentdays_365prior_visit" ))

tab_sputum <- tab_sputum %>% 
  mutate(Sample="Sputum")

stats_sputum <- sputum_metad_div %>% 
    filter(Qs<=0.1) %>% # filters fdr
    filter(abs(Ds)>=0.2)%>% #filters effect size
    filter(status!="NS") %>% 
    mutate(Sample="Sputum")

tab_sputum%>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size (Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(sputum_metad_div$Ds), max(sputum_metad_div$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=12, angle = 40, hjust = 1), axis.text.y=element_text(size=12, face="italic"), legend.position = "right")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)+
  scale_y_discrete(labels = ylabels)


metadata %>% 
  rownames_to_column(var="x_sample_id")%>% 
  left_join(alphadiversity_df_sputum, by="x_sample_id") %>% 
  ggplot(aes(x=levofloxacin_inhal, y=diversity_shannon))+
  #geom_boxplot()+
  geom_point()+
  geom_smooth(method = "lm")

metadata %>% 
  rownames_to_column(var="x_sample_id")%>% 
  left_join(alphadiversity_df_sputum, by="x_sample_id") %>% 
  ggplot(aes(x=levofloxacin_inhal, y=chlorid_mmol_cl_l))+
  #geom_boxplot()+
  geom_point()+
  geom_smooth(method = "lm")
  

#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/sputum_metaD_selected.png", width = 5.5, height = 3)
```

## MetadeconfoundR for throat diversity
```{r MetadeconfoundR for throat diversity rarefied, fig.height=4, fig.width=8}

alphadiversity_df_throat <- microbiome::alpha(ps_full_throat_rar, index = "all")
alphadiversity_df_throat$x_sample_id <- rownames(alphadiversity_df_throat)
alphadiversity_df_throat_sel <- alphadiversity_df_throat %>% 
  select(diversity_shannon, observed, evenness_pielou, dominance_dbp, x_sample_id)

ps_full_throat_df <- as.data.frame(sample_data(ps_full_throat))

diversity_df_throat_metadata<- alphadiversity_df_throat%>%
left_join(ps_full_throat_df, by = "x_sample_id")

# create metadata, to make MetadeconfoundR run properly, I need to recode all variables I want to input into MetadeconfoundR into numeric variables and I will select only the relevant variables to reduce computational time

metadata_x_throat_selected <- diversity_df_throat_metadata %>%
  select(sex, x_sample_id, id.y, mutation_status, age_y, BMI, azithromycin_oral, levofloxacin_inhal, colistin_inhal, dpi_colistin, tobramycin_inhal, azetronam_inhal, hypertonic_na_cl, isotonic_na_cl, anticholinergic, laba_inhal, saba_inhal, dn_ase, mannitol, cortison_inhal, cortison_nasal, cortison_oral, antihistaminikum, nasenspray, na_cl_oral,  ursodesoxychols_a_ure, ppi, orale_zusatznahrung, tranexams_a_ure, nsaid, laxantien, acetylcystein, insulin, antidiabetic_oral, furosemid, ssri, ab_15dprior_visit, treatmentdays_365prior_visit, chlorid_mmol_cl_l, hb_a1c_percent, crp_mg_l, interleukin6_pg_ml, interleukin8_pg_ml, ig_g_g_l, calprotectin_amg_g_stuhl, leukozyten_nl, gpt_alat_u_l, got_asat_u_l, pp_fev_percent, pp_fvc_percent, cftr_prior_v1, total_reads, delta_ETI_merged, bristol, basic_nutrition, staph_pos_throat, staph_pos_sputum, pseudo_pos_sputum, pseudo_pos_throat)  

metadata <- metadata_x_throat_selected
rownames(metadata) <- metadata$x_sample_id
metadata$insulin[metadata$x1 == "18IMP16V1Spu"]<-"0"
glimpse(metadata)
metadata <- metadata %>%
  mutate(across(where(is.character), as.numeric))

metadata$sex<- metadata$sex-1

# exclude variables which only represent 1 patient
metadata <- metadata %>%
  select(-c(x_sample_id))#, allopurinol, azithromycin_oral, nmh_enoxaparin, immunsuppressiva, na_ssa,dopaminagonist))

#create other input
features <- alphadiversity_df_throat_sel %>% 
  select(-x_sample_id)

throat_metad_div<- MetaDeconfound(featureMat = features, metaMat = metadata, randomVar = c("id.y"),  nnodes=9, returnLong = T)

tab_throat <- BuildHeatmap(throat_metad_div,intermedData = T,showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.2, keepMeta = c("chlorid_mmol_cl_l", "delta_ETI_merged", "interleukin6_pg_ml", "levofloxacin_inhal", "chlorid_mmol_cl_l","crp_mg_l", "leukozyten_nl", "pp_fev_percent", "delta_ETI_merged", "cortison_oral", "treatmentdays_365prior_visit"))

tab_throat_2 <- BuildHeatmap(throat_metad_div,intermedData = T,showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.2)

tab_throat <- tab_throat %>% 
  mutate(Sample="Throat")

stats_throat <- throat_metad_div%>% 
    filter(Qs<=0.1) %>% 
    filter(abs(Ds)>=0.2)%>% 
    filter(status!="NS") %>% 
    mutate(Sample="Throat")

throat_div_hm <- tab_throat_2%>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size (Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(sputum_metad_div$Ds), max(sputum_metad_div$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=12, angle = 40, hjust = 1), axis.text.y=element_text(size=12, face="italic"), legend.position = "right")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)+
  scale_y_discrete(labels = ylabels)
throat_div_hm

saveRDS(throat_div_hm, "/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/outputs/throat_div_hm.rds")

tabyl(metadata, insulin, id.y) # 7 patients had insulin
tabyl(metadata, ssri, id.y) #  3 patients took SSRIs

tab_all<-rbind(tab_throat, tab_sputum)

tab_all %>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size (Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(sputum_metad_div$Ds), max(sputum_metad_div$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=12, angle = 40, hjust = 1), axis.text.y=element_text(size=12, face="italic"), legend.position = "right")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)+
  scale_y_discrete(labels = ylabels)+
  facet_wrap(~Sample, nrow = 8)

```

## MetadeconfoundR for stool diversity
```{r MetadeconfoundR for stool diversity unrarefied, fig.height=5.5, fig.width=4.5}

alphadiversity_df_stool <- microbiome::alpha(ps_full_stool_rar, index = "all")
alphadiversity_df_stool$x_sample_id <- rownames(alphadiversity_df_stool)
alphadiversity_df_stool_sel <- alphadiversity_df_stool %>% 
  select(diversity_shannon, observed, evenness_pielou, dominance_dbp, x_sample_id)

ps_full_stool_df <- as.data.frame(sample_data(ps_full_stool))

diversity_df_stool_metadata<- alphadiversity_df_stool%>%
left_join(ps_full_stool_df, by = "x_sample_id")

# create metadata, to make MetadeconfoundR run properly, I need to recode all variables I want to input into MetadeconfoundR into numeric variables and I will select only the relevant variables to reduce computational time

metadata_x_stool_selected <- diversity_df_stool_metadata %>%
  select(sex, x_sample_id, id.y, mutation_status, age_y, BMI, azithromycin_oral, levofloxacin_inhal, colistin_inhal, dpi_colistin, tobramycin_inhal, azetronam_inhal, hypertonic_na_cl, isotonic_na_cl, anticholinergic, laba_inhal, saba_inhal, dn_ase, mannitol, cortison_inhal, cortison_nasal, cortison_oral, antihistaminikum, nasenspray, na_cl_oral,  ursodesoxychols_a_ure, ppi, orale_zusatznahrung, tranexams_a_ure, nsaid, laxantien, acetylcystein, insulin, antidiabetic_oral, furosemid, ssri, ab_15dprior_visit, treatmentdays_365prior_visit, chlorid_mmol_cl_l, hb_a1c_percent, crp_mg_l, interleukin6_pg_ml, interleukin8_pg_ml, ig_g_g_l, calprotectin_amg_g_stuhl, leukozyten_nl, gpt_alat_u_l, got_asat_u_l, pp_fev_percent, pp_fvc_percent, cftr_prior_v1, total_reads, delta_ETI_merged, bristol, basic_nutrition, staph_pos_throat, staph_pos_sputum, pseudo_pos_sputum, pseudo_pos_throat) 

metadata <- metadata_x_stool_selected
rownames(metadata) <- metadata$x_sample_id
metadata$insulin[metadata$x1 == "18IMP16V1Spu"]<-"0"
glimpse(metadata)
metadata <- metadata %>%
  mutate(across(where(is.character), as.numeric))

metadata$sex<- metadata$sex-1

# exclude variables which only represent 1 patient
metadata <- metadata %>%
  select(-c(x_sample_id))#, allopurinol, azithromycin_oral, nmh_enoxaparin, immunsuppressiva, na_ssa,dopaminagonist))

#create other input
features <- alphadiversity_df_stool_sel %>% 
  select(-x_sample_id)

stool_metad_div<- MetaDeconfound(featureMat = features, metaMat = metadata, nnodes=9,returnLong = T) # randomVar = list("+ (1|id.y)", c("id.y")),

tab_stool<- BuildHeatmap(stool_metad_div,intermedData = T,showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.2, keepMeta = c("chlorid_mmol_cl_l", "delta_ETI_merged", "interleukin6_pg_ml", "levofloxacin_inhal", "chlorid_mmol_cl_l","crp_mg_l", "leukozyten_nl", "pp_fev_percent", "delta_ETI_merged", "cortison_oral", "treatmentdays_365prior_visit"))

view(stool_metad_div)

tab_stool<- tab_stool%>% 
  mutate(Sample="Stool")

stats_stool <- stool_metad_div%>% 
    filter(Qs<=0.1) %>% # filters fdr
    filter(abs(Ds)>=0.2)%>% #filters effect size
    filter(Ds!="Inf")%>%
    filter(status!="NS") %>% 
    mutate(Sample="Stool")

tab_stool %>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size (Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(sputum_metad_div$Ds), max(sputum_metad_div$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=10, angle = 40, hjust = 1), axis.text.y=element_text(size=10, face="italic"), legend.position = "right")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)+
  scale_y_discrete(labels = ylabels)+
  facet_wrap(~Sample, scales = "free_y")

tab_all<-rbind(tab_throat, tab_sputum, tab_stool)

tab_all<-tab_all %>% 
  select(Sample, everything())

write_csv(tab_all, "/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/diversity_all_metaD.csv")

stats_all<-rbind(stats_throat, stats_sputum, stats_stool)

stats_all<-stats_all %>% 
  select(Sample, everything())

write_csv(stats_all, "/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/rarified_stats_diversity_all_metaD.csv")

tab_all$Sample <- factor(tab_all$Sample, levels = c("Sputum","Throat","Stool"))

diversity_hm<-tab_all %>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size\n(Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(stool_metad_div$Ds), max(stool_metad_div$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=10, angle = 40, hjust = 1, vjust = 1.05), axis.text.y=element_text(size=10, face="italic"), legend.title = element_text(size=10), legend.position = "bottom")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)+
  scale_y_discrete(labels = ylabels)+
  facet_wrap(~Sample, nrow=12,scales = "free_y")
diversity_hm

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/all_diversity_metaD.png", diversity_hm, width = 4.5, height = 5.5)

metadata %>% 
  filter(!is.na(insulin)) %>% 
  ggplot(aes(x=as.factor(insulin), y=treatmentdays_365prior_visit))+
  geom_boxplot()+
  geom_point()+
  stat_compare_means()

metadata %>%
  group_by(insulin) %>%
  summarise(
    median_treatmentdays = median(treatmentdays_365prior_visit, na.rm = TRUE),
    iqr_treatmentdays = IQR(treatmentdays_365prior_visit, na.rm = TRUE),
    mean_treatmentdays = mean(treatmentdays_365prior_visit, na.rm = TRUE)
  )

tab_stool_2<- BuildHeatmap(stool_metad_div,intermedData = T,showConfounded = F, q_cutoff = 0.1, d_cutoff = 0.2, keepMeta = c("delta_ETI_merged", "cortison_oral", "treatmentdays_365prior_visit"))

stool_div_hm <- tab_stool_2%>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size (Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(stool_metad_div$Ds), max(stool_metad_div$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=10, angle = 40, hjust = 1), axis.text.y=element_text(size=10, face="italic"), legend.position = "right")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)+
  scale_y_discrete(labels = ylabels)

stool_div_hm

saveRDS(stool_div_hm, "/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/outputs/stool_div_hm.rds")
```

## MetadeconfoundR for Sputum glomerulated taxa
```{r MetadeconfoundR for Sputum glomerulated taxa, fig.width=8, fig.height=8}

# for Sputum glomerulated taxa
ps_x_sputum_glom <- tax_glom(ps_sputum_filtered_relab, "Genus")

# take metadata from chunk above
metadata_x_sputum_selected <- diversity_df_sputum_metadata %>%
  select(sex, x_sample_id, id.y, mutation_status, age_y, BMI, azithromycin_oral, levofloxacin_inhal, colistin_inhal, dpi_colistin, tobramycin_inhal, azetronam_inhal, hypertonic_na_cl, isotonic_na_cl, anticholinergic, laba_inhal, saba_inhal, dn_ase, mannitol, cortison_inhal, cortison_nasal, cortison_oral, antihistaminikum, nasenspray, na_cl_oral,  ursodesoxychols_a_ure, ppi, orale_zusatznahrung, tranexams_a_ure, nsaid, laxantien, acetylcystein, insulin, antidiabetic_oral, furosemid, ssri, ab_15dprior_visit, treatmentdays_365prior_visit, chlorid_mmol_cl_l, hb_a1c_percent, crp_mg_l, interleukin6_pg_ml, interleukin8_pg_ml, ig_g_g_l, calprotectin_amg_g_stuhl, leukozyten_nl, gpt_alat_u_l, got_asat_u_l, pp_fev_percent, pp_fvc_percent, cftr_prior_v1, total_reads, delta_ETI_merged, bristol, basic_nutrition, staph_pos_throat, staph_pos_sputum, pseudo_pos_sputum, pseudo_pos_throat)

metadata <- metadata_x_sputum_selected
rownames(metadata) <- metadata$x_sample_id
metadata$insulin[metadata$x1 == "18IMP16V1Spu"]<-"0"
glimpse(metadata)
metadata <- metadata %>%
  mutate(across(where(is.character), as.numeric))

metadata$sex<- metadata$sex-1

# exclude variables which only represent 1 patient
metadata <- metadata %>%
  select(-c(x_sample_id))#, allopurinol, azithromycin_oral, nmh_enoxaparin, immunsuppressiva, na_ssa,dopaminagonist))

#create other input
features <- as.data.frame(otu_table(ps_x_sputum_glom))
taxtable <- as.data.frame(tax_table(ps_x_sputum_glom))

#create taxatbale to crorrectly annotate asvs in the heatmap
taxtable <- taxtable%>%
  select(Genus)
features_t <- as.data.frame(t(features))

features_named <- features_t
features_named <- merge(features_named, taxtable, by = 0)
row.names(features_named) <- features_named$Genus
features_named$Row.names <- NULL
features_named$Genus <- NULL
features_named <- as.data.frame(t(features_named))

features_named_sputum <- features_named

sputum_metad_glom<- MetaDeconfound(featureMat = features_named,metaMat = metadata, randomVar = c("id.y"),  nnodes=9, returnLong = T)

sputum_metad_glom <- sputum_metad_glom %>% 
  mutate(Sample="Sputum")

hm_05 <- BuildHeatmap(sputum_metad_glom, intermedData = T, showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.45)

view(hm_05)

sputum_hm <- hm_05 %>% 
  filter(metaVariable!="age_y") %>% 
  filter(metaVariable!="colistin_inhal") %>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size\n(Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(sputum_metad_glom$Ds), max(sputum_metad_glom$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=10, angle = 40, hjust = 1, vjust = 1.05), axis.text.y=element_text(size=10, face="italic"), legend.title = element_text(size=10), legend.position = "bottom")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)+
  scale_y_discrete(labels = ylabels)+
  ggtitle("Sputum")
sputum_hm
#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/sputum_metaD_randomVarID.pdf", width = 18, height = 21)
saveRDS(sputum_hm, "/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/outputs/sputum_metaD_25percentprevalence.rds")
```
## Staphylococcus
```{r direct correlations with sputum rel abundances, fig.height=6, fig.width=6}

metadata_features_combined_sputum <- rownames_to_column(features_named_sputum)%>%
  left_join(rownames_to_column(ps_full_sputum_df), by=c("rowname"="rowname"))

lm_visit_sum <- summary(lmerTest::lmer(Staphylococcus ~ visit_sum + age_y + sex +(1|id), data=metadata_features_combined_sputum))
lm_chloride <- summary(lmerTest::lmer(Staphylococcus ~ chlorid_mmol_cl_l + (1|id), data=metadata_features_combined_sputum))
lm_full <- summary(lmerTest::lmer(Staphylococcus ~ visit_sum + chlorid_mmol_cl_l + (1|id), data=metadata_features_combined_sputum))

anova(lmerTest::lmer(Staphylococcus ~ visit_sum + chlorid_mmol_cl_l + (1|id), data=metadata_features_combined_sputum), lmerTest::lmer(Staphylococcus ~ chlorid_mmol_cl_l + (1|id), data=metadata_features_combined_sputum))

# a model knowing about sweat chloride and visit_sum does not perform better than a model knowing about sweat chloride alone, hence the effect on Staph abundance is sweat chloride mediated

  coefs <- data.frame(coef(lm_chloride))

  lm_stats <- coefs%>%
    mutate(p = Pr...t..)
  
  lm_stats <- format(lm_stats, digits = 2)
  lm_stats <- rownames_to_column(lm_stats)%>%
    mutate(variable=rowname)%>%
    select(c(variable, Estimate, Std..Error, df, t.value, p))
  print(lm_stats)
  
  lm_stats<- lm_stats[2,]
  lm_stats$p <- format(as.numeric(lm_stats$p), scientific=F)
  
  library(ggtext)
  
visit_sum_palette <- c("black", "#C3BC3FFF","#6388B4FF","#BB7693FF","#55AD89FF") 
x2labels <- c("0", "3", "6-12", "15-18", "21-24")


p1 <- metadata_features_combined_sputum %>%
  ggplot(aes(y = Staphylococcus * 100 + 0.01, x=chlorid_mmol_cl_l)) +
  geom_point(aes(color = visit_sum), size = 4) +
  geom_smooth(method = "lm", color = "black", se = T) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0", "0.1", "1", "10", "100"), limits = c(0.005, 100)) +
  xlab("Sweat chloride mmol/l") +
  ylab("*Staphylococcus* (%) in Sputum") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 14),
    axis.title.x  = ggtext::element_markdown(size = 14),
    legend.title = element_text(size = 14),
    text = element_text(size = 14),
    legend.position = "bottom",
    axis.text = element_text(size = 14)
  ) +
  annotate("text", x = 90, y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from\ntreatment start", title.position = "left", ncol = 3)) 
p1

saveRDS(p1, "/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/cor_staph_chloride_sputum.rds")
#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/cor_staph_chloride_sputum.pdf", width = 7.5, height = 7)


```
## Pseudomonas

```{r direct correlations with sputum rel abundances, fig.height=8, fig.width=12}

pseudomonas <- metadata_features_combined_sputum %>%
  ggplot(aes(y = Pseudomonas * 100 + 0.01, x=leukozyten_nl)) +
  geom_point(aes(color = visit_sum), size = 4) +
  geom_smooth(method = "lm", color = "black", se = T) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_y_log10(
   breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0", "0.1", "1", "10", "100"), limits = c(0.005, 100)) +
  scale_x_log10()+
  xlab("Leucocytes /nl") +
  ylab("*Pseudomonas* (%) in Sputum") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "bottom",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = , y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  #geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  #geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5)) 
pseudomonas

pseu_igg <- metadata_features_combined_sputum %>%
  ggplot(aes(y = Pseudomonas * 100 + 0.01, x=ig_g_g_l)) +
  geom_point(aes(color = visit_sum), size = 4) +
  geom_smooth(method = "lm", color = "black", se = T) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0", "0.1", "1", "10", "100"), limits = c(0.005, 100)) +
  #scale_x_log10()+
  xlab("IgG g/l") +
  ylab("*Pseudomonas* (%) in Sputum") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "bottom",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = , y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  #geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  #geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5)) 

igg_leuco <- metadata_features_combined_sputum %>%
  ggplot(aes(y = leukozyten_nl, x=ig_g_g_l)) +
  geom_point(aes(color = visit_sum), size = 4) +
  geom_smooth(method = "lm", color = "black", se = T) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  #scale_y_log10(
    #breaks = c(0.01, 0.1, 1, 10, 100),
    #labels = c("0", "0.1", "1", "10", "100"), limits = c(0.005, 100)) +
  #scale_x_log10()+
  xlab("IgG g/l") +
  ylab("Leucocytes /nl") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "bottom",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = , y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  #geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  #geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5)) 


leu <- metadata_features_combined_sputum %>%
  ggplot(aes(y = leukozyten_nl, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  xlab("Months from treatment start") +
  ylab("Leucocytes /nl") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = , y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  #geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  #geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5)) 

igg <- metadata_features_combined_sputum %>%
  filter(!is.na(Pseudomonas)) %>% 
  ggplot(aes(y = ig_g_g_l, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  xlab("Months from treatment start") +
  ylab("IgG g/l") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = , y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  #geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  #geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5))

pseu_ab <- metadata_features_combined_sputum %>%
  filter(!is.na(Pseudomonas)) %>% 
  ggplot(aes(y = Pseudomonas * 100 + 0.01, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0", "0.1", "1", "10", "100"), limits = c(0.005, 100)) +
  xlab("Months from treatment start") +
  ylab("*Pseudomonas* (%) in Sputum") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = , y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  #geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  #geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5))


summary(lmerTest::lmer(Pseudomonas ~ visit_sum + age_y + sex +(1|id), data=metadata_features_combined_sputum))
summary(lmerTest::lmer(leukozyten_nl ~ visit_sum + age_y + sex +(1|id), data=metadata_features_combined_sputum))
summary(lmerTest::lmer(ig_g_g_l ~ visit_sum + age_y + sex +(1|id), data=metadata_features_combined_sputum))


ggarrange(pseudomonas, pseu_igg, igg_leuco, leu, igg, pseu_ab, labels = c("A","B","C","D","E","F"), nrow = 2, ncol = 3, legend = "bottom", common.legend = T)

#ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/pseudo_inflammation.png", width = 12, height = 8)
```

## perform little subgroup analysis within pseudomonas positive individuals only for inflammation parameters
```{r}


pseudo_pat_v1 <-  metadata_features_combined_sputum %>%
  filter(Pseudomonas>0&visit=="1") %>% 
  select(x_sample_id)

pseu_only_data <- metadata_features_combined_sputum %>%
  filter(Pseudomonas>0 | x_sample_id%in% c(pseudo_pat_v1$x_sample_id) )

summary(pseu_only_data$age_y)

pseu_ab <- pseu_only_data %>% 
ggplot(aes(y = Pseudomonas * 100 + 0.01, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0", "0.1", "1", "10", "100"), limits = c(0.005, 100)) +
  xlab("Months from treatment start") +
  ylab("*Pseudomonas* (%) in Sputum") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = , y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  #geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  #geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5))

leu <- pseu_only_data %>% 
  ggplot(aes(y = leukozyten_nl, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  #geom_line(aes(group=id), color="black", alpha=0.5)+
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  xlab("Months from treatment start") +
  ylab("Leucocytes /nl") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5)) 

summary(lmerTest::lmer(Pseudomonas ~ visit_sum + age_y + sex +(1|id), data=pseu_only_data))
summary(lmerTest::lmer(leukozyten_nl ~ visit_sum + age_y + sex +(1|id), data=pseu_only_data))
summary(lmerTest::lmer(ig_g_g_l ~ visit_sum + age_y + sex +(1|id), data=pseu_only_data))

igg <- pseu_only_data %>% 
  ggplot(aes(y = ig_g_g_l, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  #geom_line(aes(group=id), color="black", alpha=0.5)+
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  xlab("Months from treatment start") +
  ylab("IgG g/l") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5)) 

ggarrange(igg_leuco, pseudomonas, pseu_igg,pseu_ab, leu, igg, labels = c("a","b","c","d","e","f"), nrow = 2, ncol = 3, legend = "bottom", common.legend = T)

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/pseudo_inflammation.png", width = 12, height = 8)
```
## compare Pseu pos sputum vs non-pos sputum samples
```{r}

metadata_features_combined_sputum <- metadata_features_combined_sputum %>%
  mutate(pseudo_pos_sputum_16S = case_when(Pseudomonas>0~T, T~F)) 

summary(lmerTest::lmer(leukozyten_nl ~ pseudo_pos_sputum_16S +age_y + sex +(1|id), data=metadata_features_combined_sputum))
summary(lmerTest::lmer(leukozyten_nl ~ pseudo_pos_sputum_16S+visit_sum+ +age_y + sex +(1|id), data=metadata_features_combined_sputum))
summary(lmerTest::lmer(leukozyten_nl ~ visit_sum +age_y + sex +(1|id) + (1|pseudo_pos_sputum_16S), data=metadata_features_combined_sputum))

summary(lmerTest::lmer(leukozyten_nl ~ visit_sum +age_y + sex +(1|id), data=pseu_only_data))

pseu_neg_data <- metadata_features_combined_sputum %>% 
  filter(pseudo_pos_sputum_16S==F)

summary(lmerTest::lmer(leukozyten_nl ~ visit_sum +age_y + sex +(1|id), data=pseu_neg_data))


metadata_features_combined_sputum %>%
 ggplot(aes(y = leukozyten_nl, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  #geom_line(aes(group=id), color="black", alpha=0.5)+
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  xlab("Months from treatment start") +
  ylab("Leucocytes /nl") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5))+
  facet_grid(~pseudo_pos_sputum_16S)

metadata_features_combined_sputum %>%
 ggplot(aes(y = ig_g_g_l, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  #geom_line(aes(group=id), color="black", alpha=0.5)+
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  xlab("Months from treatment start") +
  ylab("IgG") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5))+
  facet_grid(~pseudo_pos_sputum_16S)

summary(lmerTest::lmer(ig_g_g_l ~ pseudo_pos_sputum_16S +age_y + sex +(1|id), data=metadata_features_combined_sputum))
summary(lmerTest::lmer(ig_g_g_l ~ visit_sum +age_y + sex +(1|id) + (1|pseudo_pos_sputum_16S), data=metadata_features_combined_sputum))

# only pseu pos. samples:
summary(lmerTest::lmer(ig_g_g_l~ visit_sum +age_y + sex +(1|id), data=pseu_only_data))
# only pseu neg. samples:
summary(lmerTest::lmer(ig_g_g_l ~ visit_sum +age_y + sex +(1|id), data=pseu_neg_data))

igg_pn <- ggplot(metadata_features_combined_sputum, aes(x = visit_sum, y = ig_g_g_l, color = pseudo_pos_sputum_16S)) +
  geom_point(position = position_jitter(width = 0.2), size=3) +
  geom_smooth(method = "lm", aes(group = pseudo_pos_sputum_16S), se = TRUE, line=2) +
  labs(x="Months from treatment start", y = "IgG (g/L)", color = "Pseudomonas") +
  theme_classic() +
  scale_color_manual(values = c("#94D0C0FF","#EE6AA7"), labels = c("negative", "positive"))+scale_x_discrete(labels= x2labels)+theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "bottom",
    axis.text = element_text(size = 12)
  ) 
igg_pn

legend2 <- as_ggplot(get_legend(igg_pn))

leu_pn <- ggplot(metadata_features_combined_sputum, aes(x = visit_sum, y = leukozyten_nl, color = pseudo_pos_sputum_16S)) +
  geom_point(position = position_jitter(width = 0.2), size=3) +
  geom_smooth(method = "lm", aes(group = pseudo_pos_sputum_16S), se = TRUE, line=2) +
  labs(x = "Months from treatment start", y = "Leucocytes /nl", color = "Pseudomonas")+
  theme_classic() +
  scale_color_manual(values = c("#94D0C0FF","#EE6AA7"), labels = c("negative", "positive"))+scale_x_discrete(labels= x2labels)+theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) 
leu_pn

legend1 <- as_ggplot(get_legend(igg_leuco))

ggarrange(igg_leuco, pseudomonas, pseu_igg,pseu_ab, leu_pn, igg_pn, labels = c("a","b","c","d","e","f"), nrow = 2, ncol = 3, legend = "none", common.legend = T)
ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/pseudo_inflammation_new.png", width = 10, height = 7)

Legend_c <- ggarrange(legend1, legend2)

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/pseudo_inflammation_legend.png", Legend_c, width = 12, height = 2)

summary(metadata_features_combined_sputum$pseudo_pos_sputum_16S)
```

## perform little subgroup analysis within Staphylococcus positive individuals only for inflammation parameters
```{r}
staph_pat_v1 <-  metadata_features_combined_sputum %>%
  filter(Staphylococcus>0&visit=="1") %>% 
  select(x_sample_id)

staph_only_data <- metadata_features_combined_sputum %>%
  filter(Staphylococcus>0 | x_sample_id%in% c(staph_pat_v1$x_sample_id) )

staph_ch <- staph_only_data %>%
  distinct(id, .keep_all = TRUE) %>%
  summarise(
    min_age = min(age_y, na.rm = TRUE),
    max_age = max(age_y, na.rm = TRUE),
    mean_age = mean(age_y, na.rm = TRUE),
    median_age = median(age_y, na.rm = TRUE),
    sd_age = sd(age_y, na.rm = TRUE),
    count = n()
  )%>%
  pivot_longer(
    cols = everything(),
    names_to = "Statistic",
    values_to = "Staphylococcus sputum+"
  )

metadata_features_combined_sputum %>%
  filter(Staphylococcus>0 | Pseudomonas>0) %>% 
  ggplot(aes(y=age_y))

metadata_features_combined_sputum %>%
  filter(Staphylococcus>0 & Pseudomonas>0) %>% 
  distinct(id)



pseu_ch <- pseu_only_data%>%
  distinct(id, .keep_all = TRUE) %>%
  summarise(
    min_age = min(age_y, na.rm = TRUE),
    max_age = max(age_y, na.rm = TRUE),
    mean_age = mean(age_y, na.rm = TRUE),
    median_age = median(age_y, na.rm = TRUE),
    sd_age = sd(age_y, na.rm = TRUE),
    count = n()
  )%>%
  pivot_longer(
    cols = everything(),
    names_to = "Statistic",
    values_to = "Pseudomonas sputum+"
  )

characteristics <- cbind(staph_ch, pseu_ch)
  

staph_only_data %>% 
ggplot(aes(y = Staphylococcus * 100 + 0.01, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0", "0.1", "1", "10", "100"), limits = c(0.005, 100)) +
  xlab("Months from treatment start") +
  ylab("*Staphylococcus* (%) in Sputum") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = , y = 200, label = "fdr=0.015, Ds=0.57", size = 5) +
  #geom_vline(xintercept = 30, linetype = 'dotted', col = 'red') +
  #geom_vline(xintercept = 60, linetype = 'dotted', col = 'red') +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5))


staph_only_data %>% 
  ggplot(aes(y = leukozyten_nl, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  #geom_line(aes(group=id), color="black", alpha=0.5)+
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  xlab("Months from treatment start") +
  ylab("Leucocytes /nl") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5)) 

summary(lmerTest::lmer(Staphylococcus ~ visit_sum + age_y + sex +(1|id), data=staph_only_data))
summary(lmerTest::lmer(leukozyten_nl ~ visit_sum + age_y + sex +(1|id), data=staph_only_data))
summary(lmerTest::lmer(ig_g_g_l ~ visit_sum + age_y + sex +(1|id), data=staph_only_data))

staph_only_data %>% 
  ggplot(aes(y = ig_g_g_l, x=visit_sum)) +
  geom_boxplot()+
  geom_point(aes(color = visit_sum), size = 4) +
  geom_line(aes(group=id), color="black", alpha=0.5)+
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_discrete(labels= x2labels)+
  xlab("Months from treatment start") +
  ylab("IgG g/l") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "none",
    axis.text = element_text(size = 12)
  ) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 5)) 

```

## add in vitro testing results
```{r}

ic<- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/IC25_Annes_Version.rds")

ic<-ic+
   theme(
    text = element_text(size = 10),
    axis.text = element_text(size = 10))+
  ggtitle("High-throughput testing of bacteria-drug interaction")
  
```


```{r arrange plots for figure 4, fig.height=13, fig.width=11}
library(gridExtra)
# Arrange the ggplot objects in a grid
grid_arranged <- grid.arrange(
sputum_hm+ labs(tag = "A") + theme(legend.margin = margin(0, 0, 0, 0), legend.box.spacing = unit(0, units = "pt"), legend.box.margin = margin(0, 0, 0, 0)),
  diversity_hm + labs(tag = "B"),
  p1 + labs(tag = "C"),
  ic + labs(tag = "D"),
  ncol = 2, nrow = 4,
  heights = c(0.75, 0.75, 1.2,1.2),
  widths=c(1.7,1),
  layout_matrix = rbind(c(1,2),
                        c(1,2),
                        c(1,3),
                        c(4,4))
)

# Print or save the arranged plot
print(grid_arranged)

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/fig4_metaD_diversity_sputum.png",grid_arranged, dpi = 600, width = 11, height = 13)
```

## MetadeconfoundR for throat glomerulated taxa
```{r MetadeconfoundR for throat glomerulated taxa, fig.width=7, fig.height=7}

# for throat glomerulated taxa
ps_x_throat_glom <- tax_glom(ps_throat_filtered_relab, "Genus")

# create metadata, to make MetadeconfoundR run properly, I need to recode all variables I want to input into MetadeconfoundR into numeric variables and I will select only the relevant variables to reduce computational time
metadata_x_throat <- as(sample_data(ps_full_throat),"data.frame")

metadata_x_throat_selected <- metadata_x_throat %>%
  select(sex, x_sample_id, id.y, mutation_status, age_y, BMI, azithromycin_oral, levofloxacin_inhal, colistin_inhal, dpi_colistin, tobramycin_inhal, azetronam_inhal, hypertonic_na_cl, isotonic_na_cl, anticholinergic, laba_inhal, saba_inhal, dn_ase, mannitol, cortison_inhal, cortison_nasal, cortison_oral, antihistaminikum, nasenspray, na_cl_oral, ursodesoxychols_a_ure, ppi, orale_zusatznahrung, tranexams_a_ure, nsaid, laxantien, acetylcystein, insulin, antidiabetic_oral, furosemid, ssri, ab_15dprior_visit, treatmentdays_365prior_visit, chlorid_mmol_cl_l, hb_a1c_percent, crp_mg_l, interleukin6_pg_ml, interleukin8_pg_ml, ig_g_g_l, calprotectin_amg_g_stuhl, leukozyten_nl, gpt_alat_u_l, got_asat_u_l, pp_fev_percent, pp_fvc_percent, cftr_prior_v1, total_reads, delta_ETI_merged, bristol, basic_nutrition, staph_pos_throat, staph_pos_sputum, pseudo_pos_sputum, pseudo_pos_throat) 

metadata <- metadata_x_throat_selected
rownames(metadata) <- metadata$x_sample_id
metadata$insulin[metadata$x1 == "18IMP16V1Spu"]<-"0"
glimpse(metadata)
metadata <- metadata %>%
  mutate(across(where(is.character), as.numeric))
metadata$sex<- metadata$sex-1

# exclude variables which only represent 1 patient
metadata <- metadata %>%
  select(-c(x_sample_id))

#create other input
features <- as.data.frame(otu_table(ps_x_throat_glom))
taxtable <- as.data.frame(tax_table(ps_x_throat_glom))

#create taxatbale to crorrectly annotate asvs in the heatmap
taxtable <- taxtable%>%
  select(Genus)
features_t <- as.data.frame(t(features))

features_named <- features_t
features_named <- merge(features_named, taxtable, by = 0)
row.names(features_named) <- features_named$Genus
features_named$Row.names <- NULL
features_named$Genus <- NULL
features_named <- as.data.frame(t(features_named))

throat_metad_glom<- MetaDeconfound(featureMat = features_named,metaMat = metadata, randomVar = c("id.y"),  nnodes=9, returnLong = T)
hm_05 <- BuildHeatmap(throat_metad_glom, intermedData = T, showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.3)

throat_metad_glom <- throat_metad_glom %>% 
  mutate(Sample="Throat")

view(hm_05)

throat_hm <- hm_05 %>% 
  filter(metaVariable!="colistin_inhal") %>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size\n(Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(sputum_metad_div$Ds), max(sputum_metad_div$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=10, angle = 40, hjust = 1, vjust = 1.05), axis.text.y=element_text(size=10, face="italic"), legend.title = element_text(size=10), legend.position = "right")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)+
  scale_y_discrete(labels = ylabels)+
  ggtitle("Throat")
throat_hm 

saveRDS(throat_hm, "/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/outputs/throat_metaD_25percentprevalence.rds")
ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/throat_metaD_randomVarID.png", width = 7, height = 7)

```

## MetadeconfoundR for stool glomerulated taxa
```{r MetadeconfoundR for stool glomerulated taxa, fig.width=7, fig.height=7}

# for stool glomerulated taxa
ps_x_stool_glom <- tax_glom(ps_stool_filtered_relab, "Genus")

# create metadata, to make MetadeconfoundR run properly, I need to recode all variables I want to input into MetadeconfoundR into numeric variables and I will select only the relevant variables to reduce computational time
metadata_x_stool <- as(sample_data(ps_full_stool),"data.frame")

metadata_x_stool_selected <- metadata_x_stool %>%
  select(sex, x_sample_id, id.y, mutation_status, age_y, BMI, azithromycin_oral, levofloxacin_inhal, colistin_inhal, dpi_colistin, tobramycin_inhal, azetronam_inhal, hypertonic_na_cl, isotonic_na_cl, anticholinergic, laba_inhal, saba_inhal, dn_ase, mannitol, cortison_inhal, cortison_nasal, cortison_oral, antihistaminikum, nasenspray, na_cl_oral, ursodesoxychols_a_ure, ppi, orale_zusatznahrung, tranexams_a_ure, nsaid, laxantien, acetylcystein, insulin, antidiabetic_oral, furosemid, ssri, ab_15dprior_visit, treatmentdays_365prior_visit, chlorid_mmol_cl_l, hb_a1c_percent, crp_mg_l, interleukin6_pg_ml, interleukin8_pg_ml, ig_g_g_l, calprotectin_amg_g_stuhl, leukozyten_nl, gpt_alat_u_l, got_asat_u_l, pp_fev_percent, pp_fvc_percent, cftr_prior_v1, total_reads, delta_ETI_merged, bristol, basic_nutrition, staph_pos_throat, staph_pos_sputum, pseudo_pos_sputum, pseudo_pos_throat) 

metadata <- metadata_x_stool_selected
rownames(metadata) <- metadata$x_sample_id
metadata$insulin[metadata$x1 == "18IMP16V1Spu"]<-"0"
glimpse(metadata)
metadata <- metadata %>%
  mutate(across(where(is.character), as.numeric))
metadata$sex<- metadata$sex-1

# exclude variables which only represent 1 patient
metadata <- metadata %>%
  select(-c(x_sample_id))

#create other input
features <- as.data.frame(otu_table(ps_x_stool_glom))
taxtable <- as.data.frame(tax_table(ps_x_stool_glom))

#create taxatbale to crorrectly annotate asvs in the heatmap
taxtable <- taxtable%>%
  select(Genus)
features_t <- as.data.frame(t(features))

features_named <- features_t
features_named <- merge(features_named, taxtable, by = 0)
row.names(features_named) <- features_named$Genus
features_named$Row.names <- NULL
features_named$Genus <- NULL
features_named <- as.data.frame(t(features_named))

stool_metad_glom<- MetaDeconfound(featureMat = features_named,metaMat = metadata, randomVar = c("id.y"),  nnodes=9, returnLong = T)
hm_05 <- BuildHeatmap(stool_metad_glom, intermedData = T, showConfounded = F, q_cutoff = 0.05, d_cutoff = 0.3)

stool_metad_glom <- stool_metad_glom %>% 
  mutate(Sample="Stool")

view(hm_05)

stool_hm <- hm_05 %>% 
  #filter(metaVariable!="levofloxacin_inhal") %>% 
  ggplot(aes(x=metaVariable,y=feature,fill=Ds))+
  geom_tile() +
  theme_minimal()+
geom_tile(aes(fill = Ds), color = "black") + 
            scale_fill_gradient2(name= "Effect size\n(Cliff's delta)", low = "#746999", high = "#69b3a2", mid = "white", midpoint = 0, guide = guide_colorbar(raster = F), 
                limits = c(min(stool_metad_glom$Ds), max(stool_metad_glom$Ds))) + 
  geom_text(aes(label=stars), color = "black", size = 5)+
  theme(axis.text.x=element_text(size=10, angle = 40, hjust = 1, vjust = 1.05), axis.text.y=element_text(size=10, face="italic"), legend.title = element_text(size=10), legend.position = "right")+
  xlab("")+
  ylab("")+
  scale_x_discrete(labels = xlabels)
  #scale_y_discrete(labels = ylabels)+
stool_hm 

saveRDS(stool_hm, "/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/outputs/stool_metaD_25percentprevalence.rds")

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/stool_metaD_randomVarID.png", width = 7, height = 7)

# use of PPI does not have an impact on calprotectin
summary(lmerTest::lmer(calprotectin_amg_g_stuhl~ppi+(1|id.y), data = metadata))

# BMI does not have an impact on calprotectin
summary(lmerTest::lmer(calprotectin_amg_g_stuhl~BMI+(1|id.y), data = metadata))

# but calprotectin has an impact on BMI
summary(lmerTest::lmer(BMI~calprotectin_amg_g_stuhl+(1|id.y), data = metadata))
# but the estimate is practically 0, so this is not clinically meaningful
metadata %>% 
  ggplot(aes(BMI, calprotectin_amg_g_stuhl))+
  geom_point()+
  geom_smooth(method = "lm")

# but the estimate is practically 0, so this is not clinically meaningful
metadata %>% 
  ggplot(aes(delta_ETI_merged,calprotectin_amg_g_stuhl))+
  geom_point()+
  geom_smooth(method = "lm")

# but treatment time has an impact on BMI for the stool cohort
summary(lmerTest::lmer(BMI~delta_ETI_merged+(1|id.y), data = metadata))
metadata %>% 
  ggplot(aes(delta_ETI_merged,BMI))+
  geom_point()+
  geom_smooth(method = "lm")
```

```{r combine metadeconfoundR statistic output tables}
metad_all <- rbind(sputum_metad_glom, throat_metad_glom, stool_metad_glom)

metad_all <- metad_all %>% 
  select(Sample, everything()) %>% 
  filter(status!="NS") %>% 
  filter(status!="NA") 
  
write_csv(metad_all, "/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/metad_all_samples.csv")
```

## Escherichia in stool
```{r direct correlations with sputum rel abundances, fig.height=6, fig.width=6}

metadata_features_combined_stool <- rownames_to_column(features_named)%>%
  left_join(rownames_to_column(ps_full_stool_df), by=c("rowname"="rowname"))

lm_visit_sum <- summary(lmerTest::lmer(Escherichia_Shigella ~ visit_sum + age_y + sex +(1|id), data=metadata_features_combined_stool))
lm_cal <- summary(lmerTest::lmer(Escherichia_Shigella ~ calprotectin_amg_g_stuhl + (1|id), data=metadata_features_combined_stool))
lm_full <- summary(lmerTest::lmer(Escherichia_Shigella ~ visit_sum + calprotectin_amg_g_stuhl + (1|id), data=metadata_features_combined_stool))

metadata_features_combined_stool$calprotectin_amg_g_stuhl

anova(lmerTest::lmer(Escherichia_Shigella ~ visit_sum + calprotectin_amg_g_stuhl + (1|id), data=metadata_features_combined_stool), lmerTest::lmer(Escherichia_Shigella ~ calprotectin_amg_g_stuhl + (1|id), data=metadata_features_combined_stool))

# a model knowing about calprotectin and visit_sum does not perform better than a model knowing about calprotectin alone, hence the effect on Escherichia abundance is calprotectin mediated

  coefs <- data.frame(coef(lm_cal))

  lm_stats <- coefs%>%
    mutate(p = Pr...t..)
  
  lm_stats <- format(lm_stats, digits = 2)
  lm_stats <- rownames_to_column(lm_stats)%>%
    mutate(variable=rowname)%>%
    select(c(variable, Estimate, Std..Error, df, t.value, p))
  print(lm_stats)
  
  lm_stats<- lm_stats[2,]
  lm_stats$p <- format(as.numeric(lm_stats$p), scientific=F)
  
  library(ggtext)
  
visit_sum_palette <- c("black", "#C3BC3FFF","#6388B4FF","#BB7693FF","#55AD89FF") 
x2labels <- c("0", "3", "6-12", "15-18", "21-24")


p1 <- metadata_features_combined_stool %>%
  ggplot(aes(y = Escherichia_Shigella * 100 + 0.01, x=calprotectin_amg_g_stuhl)) +
  geom_point(aes(color = visit_sum), size = 4) +
  geom_smooth(method = "lm", color = "black", se = T) +
  scale_color_manual(values = visit_sum_palette, labels = x2labels) +
  scale_x_log10()+
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
   labels = c("0", "0.1", "1", "10", "100"), limits = c(0.005, 100)) +
  xlab("Fecal calprotectin (mg/g)") +
  ylab("*Escherichia_Shigella* (%)") +
  theme_classic() +
  theme(
    axis.title.y  = ggtext::element_markdown(size = 12),
    axis.title.x  = ggtext::element_markdown(size = 12),
    legend.title = element_text(size = 12),
    text = element_text(size = 12),
    legend.position = "bottom",
    axis.text = element_text(size = 12)
  ) +
  #annotate("text", x = 130, y = 55, label = "p=0.012, Ds=0.57", size = 5) +
  guides(color = guide_legend(title = "Months from\ntreatment start", title.position = "left", ncol = 3)) 
p1

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/cor_esch_cale_stool.pdf", width = 4.5, height = 5)

saveRDS(p1, "/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/outputs/cor_esch_cale_stool.rds")


metadata_x_stool %>% 
  ggplot(aes(as_factor(cortison_oral), calprotectin_amg_g_stuhl, color=cortison_oral))+
  geom_boxplot()+
  geom_point()+
  scale_y_log10()+
  #scale_x_discrete(labels=xlab)+
  stat_compare_means(method = "wilcox")+
  theme_bw()

metadata_features_combined_stool$cortison_oral
```
