---
title: "Permanova analysis stool"
author: "Rebecca L. Knoll"
date: " last edit `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: show
    number_sections: yes
    smart: no
    toc: yes
    df_print: paged
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries and run functions script}
library(microViz)

pacman::p_load(tidyverse, phyloseq, magrittr, janitor, microbiome, knitr, lubridate, naniar, readxl, ggplot2, ggpubr)

library(dendextend)
#library(vegan)

ps_full <- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/ps_IMProveCF_patientsAndControls_Run1-23_18102023.rds")

visit_sum_palette <- c("black", "#C3BC3FFF", "#6388B4FF", "#BB7693FF", "#55AD89FF", "#8176AA")

```

```{r, prepare data}

ps_full_relab <- transform_sample_counts(ps_full, function(x) x/sum(x)*100)

#subset materials
ps_full_stool <- subset_samples(ps_full_relab, material== "Stool")

```

### plot principal component analysi
```{r}
PCA_all_visit_stool <- plot_ordination(ps_full_stool, ordinate(ps_full_stool, "MDS"), color = "visit_sum")

xlabels <- c("0","3","6-12","15-18","21-24","Control")

p1 <- PCA_all_visit_stool +
  geom_point(size = 4, alpha = 0.7) +
  scale_color_manual(values = visit_sum_palette, labels = xlabels) +
  theme_classic() +
  theme(text = element_text(size = 18), legend.position = "left", legend.text = element_text(size = 16), legend.title = element_blank()) +
  guides(color = guide_legend(title = "Months from treatment start", title.position = "left", ncol = 2))+
  stat_ellipse()
p1

```
# PERMANOVA in stool, all timepoints and Controls together
```{r calculate bray curtis distance for all stool samples and run PERMANOVA}

# calculate BC_distance
set.seed(100)
BC_dist<- phyloseq::distance(ps_full_stool,
                             method="bray", weighted=F)
#extract metadata
metadata<- as(sample_data(ps_full_stool),"data.frame")

### unstratified
vegan::adonis2(formula = BC_dist ~ visit_sum, data = metadata, permutations = 999, by = "margin")

### stratified PERMANOVA per id
vegan::adonis2(formula = BC_dist ~ visit_sum , data = metadata, permutations = 999, by = "margin", strata = metadata$id.x)
```

# PERMANOVA per time point comparison
## compared to controls
```{r}
set.seed(100)
# Define the function
perform_permanova_by_timepoint <- function(metadata, BC_dist, timepoints_visit_sum) {
  
  # Initialize a list to store results
  results_list <- list()
  
  for (timepoint in timepoints_visit_sum) {
    # Filter metadata for the current timepoint
    metadata_v <- metadata %>%
      filter(visit_sum %in% c(timepoint, "Control"))
    
    # Extract relevant sample IDs
    sample_ids <- metadata_v$x_sample_id
    
    # Filter the distance matrix to include only relevant samples
    BC_df <- as.data.frame(as.matrix(BC_dist))
    
    # Filter rows and columns based on sample IDs
    df_v <- BC_df[rownames(BC_df) %in% sample_ids, ]
    df_v <- df_v[, colnames(df_v) %in% sample_ids]
    
    # Convert to matrix
    BC_v <- as.matrix(df_v)
    
    # Perform PERMANOVA
    permanova_result <- vegan::adonis2(formula = BC_v ~ visit_sum, data = metadata_v, permutations = 999, by = "margin")
    
    # Extract results
    result_df <- as.data.frame(permanova_result)
    result_df$timepoint <- timepoint
    
    # add the current comparison
    result_df$comparison <- "Control"
    
    # Add to results list
    results_list[[timepoint]] <- result_df
  }
  
  # Combine results into a single data frame
  results_combined <- do.call(rbind, results_list)
  
  # Write results to a table (you can specify the file path and format)
  write.csv(results_combined, file = "permanova_results.csv", row.names = TRUE)
  
  # Return results
  return(results_combined)
}

# Example usage
timepoints_visit_sum <- c("1", "2", "3-5", "6-7", "8-10")
# Assuming metadata and BC_dist are already defined
results_controls <- perform_permanova_by_timepoint(metadata, BC_dist, timepoints_visit_sum)

```

## compared to baseline (visit 1)
```{r}
set.seed(100)
# Define the function
perform_permanova_by_timepoint <- function(metadata, BC_dist, timepoints_visit_sum) {
  
  # Initialize a list to store results
  results_list <- list()
  
  for (timepoint in timepoints_visit_sum) {
    # Filter metadata for the current timepoint
    metadata_v <- metadata %>%
      filter(visit_sum %in% c(timepoint, "1"))
    
    # Extract relevant sample IDs
    sample_ids <- metadata_v$x_sample_id
    
    # Filter the distance matrix to include only relevant samples
    BC_df <- as.data.frame(as.matrix(BC_dist))
    
    # Filter rows and columns based on sample IDs
    df_v <- BC_df[rownames(BC_df) %in% sample_ids, ]
    df_v <- df_v[, colnames(df_v) %in% sample_ids]
    
    # Convert to matrix
    BC_v <- as.matrix(df_v)
    
    # Perform PERMANOVA
    permanova_result <- vegan::adonis2(formula = BC_v ~ visit_sum, data = metadata_v, permutations = 999, by = "margin",, strata = metadata_v$id.x)
    
    # Extract results
    result_df <- as.data.frame(permanova_result)
    result_df$timepoint <- timepoint
    
    # add the current comparison
    result_df$comparison <- "1"
    
    # Add to results list
    results_list[[timepoint]] <- result_df
  }
  
  # Combine results into a single data frame
  results_combined <- do.call(rbind, results_list)
  
  # Write results to a table (you can specify the file path and format)
  write.csv(results_combined, file = "permanova_results.csv", row.names = TRUE)
  
  # Return results
  return(results_combined)
}

# Example usage
timepoints_visit_sum <- c("2", "3-5", "6-7", "8-10")
# Assuming metadata and BC_dist are already defined
results_baseline <- perform_permanova_by_timepoint(metadata, BC_dist, timepoints_visit_sum)

```
## compared to 3 months (visit 2)
```{r}
set.seed(100)
# Define the function
perform_permanova_by_timepoint <- function(metadata, BC_dist, timepoints_visit_sum) {
  
  # Initialize a list to store results
  results_list <- list()
  
  for (timepoint in timepoints_visit_sum) {
    # Filter metadata for the current timepoint
    metadata_v <- metadata %>%
      filter(visit_sum %in% c(timepoint, "2"))
    
    # Extract relevant sample IDs
    sample_ids <- metadata_v$x_sample_id
    
    # Filter the distance matrix to include only relevant samples
    BC_df <- as.data.frame(as.matrix(BC_dist))
    
    # Filter rows and columns based on sample IDs
    df_v <- BC_df[rownames(BC_df) %in% sample_ids, ]
    df_v <- df_v[, colnames(df_v) %in% sample_ids]
    
    # Convert to matrix
    BC_v <- as.matrix(df_v)
    
    # Perform PERMANOVA
    permanova_result <- vegan::adonis2(formula = BC_v ~ visit_sum, data = metadata_v, permutations = 999, by = "margin", , strata = metadata_v$id.x)
    
    # Extract results
    result_df <- as.data.frame(permanova_result)
    result_df$timepoint <- timepoint
    
    # add the current comparison
    result_df$comparison <- "2"
    
    # Add to results list
    results_list[[timepoint]] <- result_df
  }
  
  # Combine results into a single data frame
  results_combined <- do.call(rbind, results_list)
  
  # Write results to a table (you can specify the file path and format)
  write.csv(results_combined, file = "permanova_results.csv", row.names = TRUE)
  
  # Return results
  return(results_combined)
}

# Example usage
timepoints_visit_sum <- c("3-5", "6-7", "8-10")
# Assuming metadata and BC_dist are already defined
results_3months <- perform_permanova_by_timepoint(metadata, BC_dist, timepoints_visit_sum)

```
## compared to 6-12 months (visit 3-5)
```{r}
set.seed(100)
# Define the function
perform_permanova_by_timepoint <- function(metadata, BC_dist, timepoints_visit_sum) {
  
  # Initialize a list to store results
  results_list <- list()
  
  for (timepoint in timepoints_visit_sum) {
    # Filter metadata for the current timepoint
    metadata_v <- metadata %>%
      filter(visit_sum %in% c(timepoint, "3-5"))
    
    # Extract relevant sample IDs
    sample_ids <- metadata_v$x_sample_id
    
    # Filter the distance matrix to include only relevant samples
    BC_df <- as.data.frame(as.matrix(BC_dist))
    
    # Filter rows and columns based on sample IDs
    df_v <- BC_df[rownames(BC_df) %in% sample_ids, ]
    df_v <- df_v[, colnames(df_v) %in% sample_ids]
    
    # Convert to matrix
    BC_v <- as.matrix(df_v)
    
    # Perform PERMANOVA
    permanova_result <- vegan::adonis2(formula = BC_v ~ visit_sum, data = metadata_v, permutations = 999, by = "margin", strata = metadata_v$id.x)
    
    # Extract results
    result_df <- as.data.frame(permanova_result)
    result_df$timepoint <- timepoint
    
    # add the current comparison
    result_df$comparison <- "3-5"
    
    # Add to results list
    results_list[[timepoint]] <- result_df
  }
  
  # Combine results into a single data frame
  results_combined <- do.call(rbind, results_list)
  
  # Write results to a table (you can specify the file path and format)
  write.csv(results_combined, file = "permanova_results.csv", row.names = TRUE)
  
  # Return results
  return(results_combined)
}

# Example usage
timepoints_visit_sum <- c("6-7", "8-10")
# Assuming metadata and BC_dist are already defined
results_6T12months <- perform_permanova_by_timepoint(metadata, BC_dist, timepoints_visit_sum)

```

## comparison between to 15-18 months and 21-24 months
```{r}
set.seed(100)
# Define the function
perform_permanova_by_timepoint <- function(metadata, BC_dist, timepoints_visit_sum) {
  
  # Initialize a list to store results
  results_list <- list()
  
  for (timepoint in timepoints_visit_sum) {
    # Filter metadata for the current timepoint
    metadata_v <- metadata %>%
      filter(visit_sum %in% c(timepoint, "6-7"))
    
    # Extract relevant sample IDs
    sample_ids <- metadata_v$x_sample_id
    
    # Filter the distance matrix to include only relevant samples
    BC_df <- as.data.frame(as.matrix(BC_dist))
    
    # Filter rows and columns based on sample IDs
    df_v <- BC_df[rownames(BC_df) %in% sample_ids, ]
    df_v <- df_v[, colnames(df_v) %in% sample_ids]
    
    # Convert to matrix
    BC_v <- as.matrix(df_v)
    
    # Perform PERMANOVA
    permanova_result <- vegan::adonis2(formula = BC_v ~ visit_sum, data = metadata_v, permutations = 999, by = "margin", strata = metadata_v$id.x)
    
    # Extract results
    result_df <- as.data.frame(permanova_result)
    result_df$timepoint <- timepoint
    
    # add the current comparison
    result_df$comparison <- "6-7"
    
    # Add to results list
    results_list[[timepoint]] <- result_df
  }
  
  # Combine results into a single data frame
  results_combined <- do.call(rbind, results_list)
  
  # Write results to a table (you can specify the file path and format)
  write.csv(results_combined, file = "permanova_results.csv", row.names = TRUE)
  
  # Return results
  return(results_combined)
}

# Example usage
timepoints_visit_sum <- c("8-10")
# Assuming metadata and BC_dist are already defined
results_15T18months <- perform_permanova_by_timepoint(metadata, BC_dist, timepoints_visit_sum)

```
# combine tables and plot a triangle heatmap displaying the comparisons
```{r}

permanova_result_combined <- rbind(results_controls, results_baseline, results_3months, results_6T12months, results_15T18months)

permanova_result_combined$comparison <- factor(permanova_result_combined $comparison, levels = c("Control","1", "2","3-5","6-7","8-10"))

tabyl(permanova_result_combined, timepoint, comparison) # all comparisons are there

permanova_result_combined <- permanova_result_combined %>% 
  mutate(p_label= as_factor(case_when(`Pr(>F)` <= 0.05 ~ "*", `Pr(>F)` <= 0.01 ~ "**", `Pr(>F)` <= 0.001 ~ "***", `Pr(>F)` <= 0.1 ~ ".", `Pr(>F)` > 0.1 ~ "ns")))

permanova_result_combined %>% 
  filter(Df==1) %>% 
  ggplot(aes(x=timepoint, y=comparison, fill = R2))+
  geom_tile()+
  coord_flip()+
  geom_text(aes(label = p_label))+
  theme_classic()

permanova_result_combined
```

# improve plot
```{r, fig.height=3.5, fig.width=5}
# Define significance levels
permanova_result_combined <- permanova_result_combined %>%
  mutate(Significance = case_when(
    `Pr(>F)` <= 0.001 ~ "≤ 0.001",
    `Pr(>F)` <= 0.01 ~ "≤ 0.01",
    `Pr(>F)` <= 0.05 ~ "≤ 0.05",
    TRUE ~ "ns"
  ))

custom_labels <- c("1" = "0", 
                   "2" = "3", 
                   "3-5" = "6-12", 
                   "6-7" = "15-18", 
                   "8-10" = "21-24", 
                   "Control" = "Control")

permanova_result_combined %>% 
  filter(Df==1) %>% 
ggplot(aes(x = comparison, y = timepoint)) +
  geom_point(aes(fill= Significance, size = R2*100), shape = 21, color="black") +
  scale_fill_manual(values = c("≤ 0.001" = "darkred", "≤ 0.01" = "red", "≤ 0.05" = "orange", "ns" = "grey"))+
  scale_size_continuous(range = c(2, 10)) +  # Adjust the range for better visualization
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(color = "p value", size = "Explained variance (%)") +
  ggtitle("Explained variance and p-values across timepoint comparisons") 

p1 <- permanova_result_combined %>% 
  filter(Df==1) %>% 
ggplot(aes(x = comparison, y = timepoint)) +
  geom_point(aes(fill= Significance, size = R2*100), shape = 21, color="black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "ns" = "grey")) +
  scale_size_continuous(range = c(2, 10)) +  # Adjust the range for better visualization
  theme_minimal() +
  scale_x_discrete(labels=custom_labels)+
  scale_y_discrete(labels=custom_labels)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(fill = "p value", size = "Explained variance (%)") +
  ylab("months from ETI treatment start")+
  xlab("months from ETI treatment start")#+
  #ggtitle("Stool: PERMANOVA between sample timepoints") 
p1
saveRDS(p1, "~/Documents/Forschung/IMMProveCF/R_analysis/rds_files/PERMANOVA_timepoints_stool.rds")

write.csv(permanova_result_combined, "~/Documents/Forschung/IMMProveCF/R_analysis/table_outputs/PERMANOVA_timepoints_stool.csv")

```
### bubble plot without controls
```{r}
p5 <- permanova_result_combined %>% 
  filter(comparison!="Control") %>% 
  filter(Df==1) %>% 
ggplot(aes(x = comparison, y = timepoint)) +
  geom_point(aes(fill= Significance, size = R2*100), shape = 21, color="black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "ns" = "grey")) +
  scale_size_continuous(range = c(2, 10)) +  # Adjust the range for better visualization
  theme_minimal() +
  scale_x_discrete(labels=custom_labels)+
  scale_y_discrete(labels=custom_labels)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(fill = "p value", size = "Explained variance (%)") +
  ylab("months from ETI treatment start")+
  xlab("months from ETI treatment start")#+
  #ggtitle("Stool: PERMANOVA between sample timepoints") 
p5
ggsave( "~/Documents/Forschung/IMMProveCF/R_analysis/figures/PERMANOVA_timepoints_stool_woControls.png", p5, height = 3.5, width = 6)
saveRDS(p5, "~/Documents/Forschung/IMMProveCF/R_analysis/rds_files/PERMANOVA_timepoints_stool_woControls.rds")

```

## univariate

```{r}
# select only CF patients

ps_p_stool <- subset_samples(ps_full_stool, project=="IMP")

features.p <- as.data.frame(otu_table(ps_p_stool))

# Variables to test (same as for the MetadeconfoundR analysis)
metadata_to_test <- metadata %>%
  filter(project=="IMP") %>% 
  select(sex, id.y, visit_sum, mutation_status, age_y, BMI, azithromycin_oral, levofloxacin_inhal, colistin_inhal, dpi_colistin, tobramycin_inhal, azetronam_inhal, hypertonic_na_cl, isotonic_na_cl, anticholinergic, laba_inhal, saba_inhal, dn_ase, mannitol, cortison_inhal, cortison_nasal, cortison_oral, antihistaminikum, nasenspray, na_cl_oral,  ursodesoxychols_a_ure, ppi, orale_zusatznahrung, tranexams_a_ure, nsaid, laxantien, acetylcystein, insulin, antidiabetic_oral, furosemid, ssri, ab_15dprior_visit, treatmentdays_365prior_visit, chlorid_mmol_cl_l, hb_a1c_percent, crp_mg_l, interleukin6_pg_ml, interleukin8_pg_ml, ig_g_g_l, calprotectin_amg_g_stuhl, leukozyten_nl, gpt_alat_u_l, got_asat_u_l, pp_fev_percent, pp_fvc_percent, cftr_prior_v1, total_reads, delta_ETI_merged, bristol, basic_nutrition, staph_pos_throat, staph_pos_sputum, pseudo_pos_sputum, pseudo_pos_throat) 

variables_test <- colnames(metadata_to_test)

# Initialize an empty data frame to store the results
results <- data.frame(variable = character(), R2 = numeric(), p_value = numeric(), N = numeric(), stringsAsFactors = FALSE)

# Loop through each variable
for (var in variables_test) {
  
  # Clean metadata by removing rows with missing data for the current variable
  metadata_clean <- na.omit(metadata_to_test[, c(var, "id.y")])
  
  # Subset features.p to match the cleaned metadata
  features_clean <- features.p[rownames(metadata_clean), ]
  
  # Construct the formula dynamically
  formula <- as.formula(paste("features_clean ~", var))
  
  # Perform PERMANOVA
  set.seed(111)
  perm_result <- vegan::adonis2(formula, 
                         data = metadata_clean, 
                         permutations = 999, 
                         method = "bray", 
                         by = "margin", 
                         strata = metadata_clean$id.y)
  
  perm_result<- clean_names(as_tibble(perm_result))
  # Extract the results: R2, p-value, and number of observations
  R2 <- perm_result[1,3]
  p_value <- perm_result[1,5]
  N <- perm_result[3,1]
  
  # Append the results to the data frame
  results <- rbind(results, data.frame(variable = var, R2 = R2, p_value = p_value, N = N))
}

# View the results
print(results)

results_filtered <- results %>% 
  filter(variable!="id.y") %>% #in the above I stratify for id, thus it's not truly tested here
  mutate(fdr = p.adjust(pr_f, method=c("fdr"))) %>%  # fdr adjust for multiple comparisons
  filter(pr_f<=0.05) # filter for p-value below 0.05

# Optionally, save the results to a CSV file
# write.csv(results, "permanova_results.csv", row.names = FALSE)
```

## plot univariate permanova
```{r}

# Reorder the 'Variable' factor levels based on 'r2'
results_filtered <- results_filtered %>%
  arrange(desc(r2)) %>%
  mutate(variable = factor(variable, levels = unique(variable))) %>% 
  mutate(y="Y") %>% 
  mutate(p_sig = case_when(pr_f <= 0.001 ~ "≤ 0.001", pr_f <= 0.01 ~"≤ 0.01", pr_f <= 0.05 ~ "≤ 0.05"))


results_filtered%>% 
ggplot(aes(x = variable, y=y)) +
  geom_point(aes(fill= p_sig, size = r2*100), shape = 21, color="black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "ns" = "grey")) +
  scale_size_continuous(range = c(2, 10)) +  # Adjust the range for better visualization
  theme_minimal() +
  #scale_x_discrete(labels=custom_labels)+
  #scale_y_discrete(labels=custom_labels)+
  theme(axis.text.x = element_text(angle = 75, hjust = 1), legend.position = "bottom", legend.title.position = "top") +
  labs(fill = "p value", size = "Explained variance (%)") +
  ylab("")+
  xlab("Variables tested")#+
 # ggtitle("Throat: PERMANOVA between sample timepoints") 

results_filtered%>% 
ggplot(aes(x = variable, y = r2 * 100, fill = p_sig)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "ns" = "grey")) +
  theme_minimal() +
  #scale_x_discrete(labels=custom_labels)+
  #scale_y_discrete(labels=custom_labels)+
  theme(axis.text.x = element_text(angle = 75, hjust = 1), legend.position = "right", legend.title.position = "top") +
  labs(fill = "p value", size = "Explained variance (%)") +
  ylab("Explained variance (%)")+
  xlab("Variables tested")

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/figures/stool_univariate_permanova_stratified.png")
```
```{r}
results_filtered_fdr <- results_filtered %>% 
  arrange(desc(r2)) %>%
  mutate(variable = factor(variable, levels = unique(variable))) %>% 
  mutate(fdr_sig = case_when(fdr <= 0.001 ~ "≤ 0.001", fdr <= 0.01 ~"≤ 0.01", fdr<= 0.05 ~ "≤ 0.05", fdr<= 0.1 ~ "≤ 0.1" )) %>% 
  filter(fdr<=0.1) 

results_filtered_fdr%>% 
  ggplot(aes(x = variable, y = r2 * 100, fill = fdr_sig)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "≤ 0.1"  = "grey")) +
  theme_minimal() +
  #scale_x_discrete(labels=custom_labels)+
  #scale_y_discrete(labels=custom_labels)+
  theme(axis.text.x = element_text(angle = 75, hjust = 1), legend.position = "right", legend.title.position = "top") +
  labs(fill = "fdr value", size = "Explained variance (%)") +
  ylab("Explained variance (%)")+
  xlab("Variables tested")

```
## combined PERMANOVA on variables that had a significant p in the univariate testing
```{r}
p_filtered <- results%>% 
  filter(pr_f<=0.05)

variables_p_sig <- c(p_filtered$variable)
variables_p_sig

# Run one single PERMANOVA through those
  metadata.p <- metadata_to_test %>% 
  select(id.y, any_of(variables_p_sig))
  
metadata.p_woNA <- metadata.p %>% na.omit() # this tests on 85 samples that have the complete metadata for the variables studied, pseudo_pos_throat is the limiting factor as only for 85 stool samples we have corresponding throat microbiology data

pat_N <- as_data_frame(tabyl(metadata.p_woNA,id.y)) %>% 
  filter(n!=0) # has 35 patients analyzed

# Subset features.p to match the cleaned metadata
  features_clean <- features.p[rownames(metadata.p_woNA), ]
  
  # Perform PERMANOVA
  set.seed(111)
  perm_result <- vegan::adonis2(features_clean ~ visit_sum+ tobramycin_inhal+azetronam_inhal+dn_ase+mannitol+cortison_oral+tranexams_a_ure+ssri+treatmentdays_365prior_visit+crp_mg_l+calprotectin_amg_g_stuhl+delta_ETI_merged+pseudo_pos_throat, 
                         data = metadata.p_woNA, 
                         permutations = 999, 
                         method = "bray", 
                         by = "margin", 
                         na.action = na.exclude,
                         strata = metadata.p_woNA$id.y)
  perm_result
  perm_result_df<- clean_names(as.data.frame(perm_result))

# Set row names
  rownames(perm_result_df) <- c(colnames(metadata.p_woNA)[c(2:14)], "Residual", "Total")
  perm_result_df$variable <- rownames(perm_result_df)
  perm_result_df<- perm_result_df %>% 
    select(variable, everything())
perm_result_df

# plot it
perm_result_df <- perm_result_df %>% 
arrange(desc(r2)) %>%
  mutate(variable = factor(variable, levels = unique(variable))) %>% 
  mutate(p_sig = case_when(pr_f <= 0.001 ~ "≤ 0.001", pr_f <= 0.01 ~"≤ 0.01", pr_f<= 0.05 ~ "≤ 0.05", pr_f<= 0.1 ~ "≤ 0.1" )) 

perm_result_df %>% 
  filter(pr_f<=0.1) %>% 
  ggplot(aes(x = variable, y = r2 * 100, fill = p_sig)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "≤ 0.1"  = "grey")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1), legend.position = "right", legend.title.position = "top") +
  labs(fill = "p value", size = "Explained variance (%)") +
  ylab("Explained variance (%)")+
  xlab("Variables tested")

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/figures/stool_combined_permanova_stratified_85Samples.png")

  ## repeat without ETI treatment days, because other parameters are often collinear to that
  perm_result_woETI <- vegan::adonis2(features_clean ~  tobramycin_inhal+azetronam_inhal+dn_ase+mannitol+cortison_oral+tranexams_a_ure+ssri+treatmentdays_365prior_visit+crp_mg_l+calprotectin_amg_g_stuhl+pseudo_pos_throat, 
                         data = metadata.p_woNA, 
                         permutations = 999, 
                         method = "bray", 
                         by = "margin", 
                         na.action = na.exclude,
                         strata = metadata.p_woNA$id.y)
  perm_result_woETI
  perm_result_df_woETI<- clean_names(as.data.frame(perm_result_woETI))
  

  # Set row names
  rownames(perm_result_df_woETI) <- c(colnames(metadata.p_woNA)[c(2:11,13)], "Residual", "Total")
  perm_result_df_woETI$variable <- rownames(perm_result_df_woETI)
  perm_result_df_woETI<- perm_result_df_woETI %>% 
    select(variable, everything())
perm_result_df_woETI

# plot it
perm_result_df_woETI <- perm_result_df_woETI %>% 
arrange(desc(r2)) %>%
  mutate(variable = factor(variable, levels = unique(variable))) %>% 
  mutate(p_sig = case_when(pr_f <= 0.001 ~ "≤ 0.001", pr_f <= 0.01 ~"≤ 0.01", pr_f<= 0.05 ~ "≤ 0.05", pr_f<= 0.1 ~ "≤ 0.1" )) 

perm_result_df_woETI %>% 
  filter(pr_f<=0.1) %>% 
  ggplot(aes(x = variable, y = r2 * 100, fill = p_sig)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "≤ 0.1"  = "grey")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1), legend.position = "right", legend.title.position = "top") +
  labs(fill = "p value", size = "Explained variance (%)") +
  ylab("Explained variance (%)")+
  xlab("Variables tested")  

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/figures/sputum_combined_permanova_stratified_31Samples:withoutETI.png")
```

### repeat above without pseudo_pos_throat, crp and calprotectin, as those reduces available data set too much

```{r}
p_filtered <- results%>% 
  filter(pr_f<=0.1) %>% 
  filter(df>=200)

variables_p_sig <- c(p_filtered$variable)
variables_p_sig

# Run one single PERMANOVA through those
  metadata.p <- metadata_to_test %>% 
  select(id.y, any_of(variables_p_sig))
  
metadata.p_woNA <- metadata.p %>% na.omit() # this tests on 202 samples that have the complete metadata for the variables studied

pat_N <- as_data_frame(tabyl(metadata.p_woNA,id.y)) %>% 
  filter(n!=0) # has 35 patients analyzed

# Subset features.p to match the cleaned metadata
  features_clean <- features.p[rownames(metadata.p_woNA), ]
  
  # Perform PERMANOVA
  set.seed(111)
  perm_result <- vegan::adonis2(features_clean ~ visit_sum+ tobramycin_inhal+azetronam_inhal+dn_ase+mannitol+cortison_oral+tranexams_a_ure+ssri+treatmentdays_365prior_visit+delta_ETI_merged, 
                         data = metadata.p_woNA, 
                         permutations = 999, 
                         method = "bray", 
                         by = "margin", 
                         na.action = na.exclude,
                         strata = metadata.p_woNA$id.y)
  perm_result
  perm_result_df<- clean_names(as.data.frame(perm_result))

# Set row names
  rownames(perm_result_df) <- c(colnames(metadata.p_woNA)[c(2:11)], "Residual", "Total")
  perm_result_df$variable <- rownames(perm_result_df)
  perm_result_df<- perm_result_df %>% 
    select(variable, everything())
perm_result_df

# plot it
perm_result_df <- perm_result_df %>% 
arrange(desc(r2)) %>%
  mutate(variable = factor(variable, levels = unique(variable))) %>% 
  mutate(p_sig = case_when(pr_f <= 0.001 ~ "≤ 0.001", pr_f <= 0.01 ~"≤ 0.01", pr_f<= 0.05 ~ "≤ 0.05", pr_f<= 0.1 ~ "≤ 0.1" )) 

perm_result_df %>% 
  filter(pr_f<=0.1) %>% 
  ggplot(aes(x = variable, y = r2 * 100, fill = p_sig)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "≤ 0.1"  = "grey")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1), legend.position = "right", legend.title.position = "top") +
  labs(fill = "p value", size = "Explained variance (%)") +
  ylab("Explained variance (%)")+
  xlab("Variables tested")

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/figures/stool_combined_permanova_stratified_202Samples.png")
    
## repeat without ETI treatment days, because other parameters are often collinear to that
  perm_result_woETI <- vegan::adonis2(features_clean ~  tobramycin_inhal+azetronam_inhal+dn_ase+mannitol+cortison_oral+tranexams_a_ure+ssri+treatmentdays_365prior_visit, 
                         data = metadata.p_woNA, 
                         permutations = 999, 
                         method = "bray", 
                         by = "margin", 
                         na.action = na.exclude,
                         strata = metadata.p_woNA$id.y)
  perm_result_woETI
  perm_result_df_woETI<- clean_names(as.data.frame(perm_result_woETI))
  

  # Set row names
  rownames(perm_result_df_woETI) <- c(colnames(metadata.p_woNA)[c(2:9)], "Residual", "Total")
  perm_result_df_woETI$variable <- rownames(perm_result_df_woETI)
  perm_result_df_woETI<- perm_result_df_woETI %>% 
    select(variable, everything())
perm_result_df_woETI

# plot it
perm_result_df_woETI <- perm_result_df_woETI %>% 
arrange(desc(r2)) %>%
  mutate(variable = factor(variable, levels = unique(variable))) %>% 
  mutate(p_sig = case_when(pr_f <= 0.001 ~ "≤ 0.001", pr_f <= 0.01 ~"≤ 0.01", pr_f<= 0.05 ~ "≤ 0.05", pr_f<= 0.1 ~ "≤ 0.1" )) 

perm_result_df_woETI %>% 
  filter(pr_f<=0.1) %>% 
  ggplot(aes(x = variable, y = r2 * 100, fill = p_sig)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("≤ 0.001" = "#765085", "≤ 0.01" = "#31A1B3FF", "≤ 0.05" = "#CCB22BFF", "≤ 0.1"  = "grey")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 75, hjust = 1), legend.position = "right", legend.title.position = "top")+
  labs(fill = "p value", size = "Explained variance (%)") +
  ylab("Explained variance (%)")+
  xlab("Variables tested")  

ggsave("/Users/rebecca/Documents/Forschung/IMMProveCF/Manuscript_16S/IMMProveCF_public/figures/stool_combined_permanova_stratified_202Samples_woETI.png")
```