---
title: "Procrustes analysis to investigate gut-lung, vs lung-throat axis"
author: "Rebecca L. Knoll"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal
Use Orthogonal Procrustes Analysis (OPA) to assess how similar the overall spatial configuration (beta diversity structure) is between:

    Sputum vs. Gut
    Throat vs. Sputum

OPA will rotate, translate, and scale one ordination to best match another and quantify similarity with a Procrustes statistic (m²).

# Interpretation

    Procrustes statistic (m²): Lower means better alignment.
    protest(): Gives a p-value testing if the alignment is better than expected by chance (random configuration).
    If the alignment is significant and tight, it supports the idea that there is a shared ecological structure, aka a gut-lung or airway continuum.

```{r run import script, include=FALSE}
# run script to tidy the data and to load packages, creates phyloseq object per material and per year; load libraries

ps_sputum_full <- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/ps_IMProveCF_patients_Run1-23_18102023.rds")

ps_full <- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/ps_IMProveCF_patientsAndControls_Run1-23_18102023.rds")

pacman::p_load(rmarkdown, tidyverse, phyloseq, magrittr, janitor, microbiome, knitr, lubridate, naniar, readxl, ggplot2, ggpubr, vegan)

metadata <- as(sample_data(ps_full), "data.frame")

metadata_v1 <- metadata %>% 
   filter(project=="IMMPIMMP") %>% 
  filter(visit==1)

id_palette <- c(IMP11="#4E79A7FF",IMP13="#A0CBE8FF",IMP15="#F28E2BFF",IMP16="#FFBE7DFF",IMP17="#59A14FFF",IMP24="#8CD17DFF", IMP25="#B6992DFF",IMP26="#F1CE63FF",IMP27="#499894FF", IMP29="#86BCB6FF", IMP32="#E15759FF",  IMP33="#FF9D9AFF", IMP35="#79706EFF", IMP5="#BAB0ACFF", IMP6="#D37295FF",IMP9="#FABFD2FF",IMP21="#B07AA1FF", IMP31="#D4A6C8FF", IMP8="black")

source("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/IMMProveCF/functions_sputhr.R")

dom_palette <- c(Streptococcus="#69b3a2", Staphylococcus = "#8175AA", Fusobacterium="#6FB899FF", Prevotella_7="#31A1B3FF", Rothia="#027B8EFF", Pseudomonas="#EE6AA7", Eikenella="#94D0C0FF", Haemophilus="#CCB22BFF", Achromobacter="#9F8F12FF", Gemella= "#97CFD0" , Moraxella= "#6FB899", `missing sample` = "#CFCFCF")
```

```{r subset per sample type, include=FALSE}
material_s_t <- c("Sputum", "Stool")

ps_Sputum_Stool_full <- subset_samples(ps_sputum_full, material %in% material_s_t)

ps_sputum_full_sputum <- subset_samples(ps_sputum_full, material== "Sputum")

id_visit_sputum <- sample_data(ps_sputum_full_sputum)$id_visit

ps_Sputum_Stool_full_unique <- subset_samples(ps_sputum_full, id_visit %in% id_visit_sputum)#includes also stool samples
ps_Sputum_Stool_full_unique <- subset_samples(ps_Sputum_Stool_full_unique, material %in% material_s_t)#includes still sputum samples which do not have a paired stool sample

# find those stool samples which have a paired sputum sample
ps_sputum_full_stool <- subset_samples(ps_sputum_full, material== "Stool")
ps_Sputum_Stool_full_unique_3 <- subset_samples(ps_sputum_full_stool, id_visit %in% id_visit_sputum)#includes also stool samples
id_visit_unique <- sample_data(ps_Sputum_Stool_full_unique_3)$id_visit

#filter for only those samples which have both sputum and stool from the same timepoint
ps_Sputum_Stool_full_unique_2 <- subset_samples(ps_Sputum_Stool_full_unique, id_visit %in% id_visit_unique)#includes still sputum samples which do not have a paired stool sample

summary(sample_data(ps_Sputum_Stool_full_unique_2)$id) # 15 participants provided both sputum & stool samples at the same timepoint

data <- as(sample_data(ps_Sputum_Stool_full_unique_2), "data.frame")

tabyl(data, id, visit_cal_cor)

#transform to relative abundance
ps_Sputum_Stool_full_unique_2_relab <- transform_sample_counts(ps_Sputum_Stool_full_unique_2 , function(x) x/sum(x))
```

# Sputum vs. stool microbiome
## Sample overview
```{r}
# plot sample overview
bp_ps_relab <- plot_bar(ps_Sputum_Stool_full_unique_2_relab, "id", fill="Phylum")

bp_ps_relab+
  geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")+
    theme(legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),legend.text = element_text(size=8), legend.position = "bottom", 
        axis.ticks = element_blank(), axis.text.x = element_blank())+
  scale_x_discrete(name="Participant")+
  scale_y_continuous(name="Relative abundance")+
  facet_grid(rows= vars(material), cols = vars(visit_cal_cor), scales ="free_x", space ="free_x")
```

```{r}
sputum_common  <- subset_samples(ps_Sputum_Stool_full_unique_2_relab, material=="Sputum")
gut_common<- subset_samples(ps_Sputum_Stool_full_unique_2_relab, material=="Stool")

# Create distance matrices (e.g., Bray-Curtis)
gut_dist <- phyloseq::distance(gut_common, method = "bray")
sputum_dist <- phyloseq::distance(sputum_common, method = "bray")

# Ordinate using PCoA (or use NMDS if you'd rather)
gut_ord <- ordinate(gut_common, method = "PCoA", distance = gut_dist)
sputum_ord <- ordinate(sputum_common, method = "PCoA", distance = sputum_dist)

# Grab the first few axes (same number)
gut_coords  <- cmdscale(gut_dist, k = 10)
sputum_coords <- cmdscale(sputum_dist, k = 10)

?cmdscale()
?scores()

# Run Procrustes analysis
proc <- procrustes(gut_coords, sputum_coords)

# Summary
summary(proc)

# Significance test
proc_test <- protest(gut_coords, sputum_coords, permutations = 999)
print(proc_test)

# Visualization
plot(proc)

```

## do it subject wise
```{r}
library(vegan)

extract_subject_id <- function(names_vector) {
  regmatches(names_vector, regexpr("IMP\\d+", names_vector))
}

extract_visit <- function(names_vector) {
  regmatches(names_vector, regexpr("V\\d+", names_vector))
}

gut_meta <- data.frame(sample_id = rownames(gut_coords))
gut_meta$subject <- extract_subject_id(gut_meta$sample_id)
gut_meta$visit <- extract_visit(gut_meta$sample_id)
gut_meta$pair_id <- paste0(gut_meta$subject, "_", gut_meta$visit)

sputum_meta <- data.frame(sample_id = rownames(sputum_coords))
sputum_meta$subject <- extract_subject_id(sputum_meta$sample_id)
sputum_meta$visit <- extract_visit(sputum_meta$sample_id)
sputum_meta$pair_id <- paste0(sputum_meta$subject, "_", sputum_meta$visit)

subjects <- unique(gut_meta$subject)

results <- lapply(subjects, function(subj) {
  gut_sub <- gut_meta[gut_meta$subject == subj, ]
  sputum_sub <- sputum_meta[sputum_meta$subject == subj, ]

  shared_pair_ids <- intersect(gut_sub$pair_id, sputum_sub$pair_id)
  
  if (length(shared_pair_ids) >= 1) {
    gut_mat <- gut_coords[gut_sub$sample_id[gut_sub$pair_id %in% shared_pair_ids], , drop = FALSE]
    sputum_mat <- sputum_coords[sputum_sub$sample_id[sputum_sub$pair_id %in% shared_pair_ids], , drop = FALSE]

    # make sure order is identical
    gut_mat <- gut_mat[order(rownames(gut_mat)), ]
    sputum_mat <- sputum_mat[order(rownames(sputum_mat)), ]

    proc <- protest(gut_mat, sputum_mat)
    return(data.frame(subject = subj, r = proc$t0, p = proc$signif))
  } else {
    return(NULL)
  }
})

gut_sputum_results <- do.call(rbind, results)
print(gut_sputum_results)

```
# Sputum vs Throat
```{r}
# Subset your phyloseq objects to overlapping samples
material_s_t <- c("Sputum", "Throat")

ps_sputum_throat_full <- subset_samples(ps_sputum_full, material %in% material_s_t)

ps_sputum_full_sputum <- subset_samples(ps_sputum_full, material== "Sputum")

id_visit_sputum <- sample_data(ps_sputum_full_sputum)$id_visit

ps_sputum_throat_full_unique <- subset_samples(ps_sputum_full, id_visit %in% id_visit_sputum)#includes also stool samples
ps_sputum_throat_full_unique <- subset_samples(ps_sputum_throat_full_unique, material %in% material_s_t)#includes still sputum samples which do not have a paired throat sample

# find those throat samples which have a paired sputum sample
ps_sputum_full_throat <- subset_samples(ps_sputum_full, material== "Throat")
ps_sputum_throat_full_unique_3 <- subset_samples(ps_sputum_full_throat, id_visit %in% id_visit_sputum)#includes also stool samples
id_visit_unique <- sample_data(ps_sputum_throat_full_unique_3)$id_visit

#filter for only those samples which have both sputum and throat from the same timepoint
ps_sputum_throat_full_unique_2 <- subset_samples(ps_sputum_throat_full_unique, id_visit %in% id_visit_unique)#includes still sputum samples which do not have a paired throat sample

summary(sample_data(ps_sputum_throat_full_unique_2)$id) # 15 participants provided both sputum & throat samples at the same timepoint

data <- as(sample_data(ps_sputum_throat_full_unique_2), "data.frame")

tabyl(data, id, visit_cal_cor)

#transform to relative abundance
ps_sputum_throat_full_unique_2_relab <- transform_sample_counts(ps_sputum_throat_full_unique_2 , function(x) x/sum(x))
```
## Sample overview
```{r}
# plot sample overview
bp_ps_relab <- plot_bar(ps_sputum_throat_full_unique_2_relab, "id", fill="Phylum")

bp_ps_relab+
  geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")+
    theme(legend.key.height= unit(3, 'mm'),
        legend.key.width= unit(3, 'mm'),legend.text = element_text(size=8), legend.position = "bottom", 
        axis.ticks = element_blank(), axis.text.x = element_blank())+
  scale_x_discrete(name="Participant")+
  scale_y_continuous(name="Relative abundance")+
  facet_grid(rows= vars(material), cols = vars(visit_cal_cor), scales ="free_x", space ="free_x")
```

```{r}
sputum_common  <- subset_samples(ps_sputum_throat_full_unique_2_relab, material=="Sputum")
throat_common<- subset_samples(ps_sputum_throat_full_unique_2_relab, material=="Throat")

# Create distance matrices (e.g., Bray-Curtis)
throat_dist <- phyloseq::distance(throat_common, method = "bray")
sputum_dist <- phyloseq::distance(sputum_common, method = "bray")

# Grab the first few axes (same number)
throat_coords  <- cmdscale(throat_dist, k=10)
sputum_coords <- cmdscale(sputum_dist, k=10)

?cmdscale()
?scores()

# Run Procrustes analysis
proc_resp <- procrustes(throat_coords, sputum_coords)
?procrustes

# Summary
summary(proc_resp)

# Significance test
proc_test <- protest(throat_coords, sputum_coords, permutations = 999)
print(proc_test)

# Visualization
plot(proc)

```

## subject wise on throat-sputum
```{r}
throat_meta <- data.frame(sample_id = rownames(throat_coords))
throat_meta$subject <- extract_subject_id(throat_meta$sample_id)
throat_meta$visit <- extract_visit(throat_meta$sample_id)
throat_meta$pair_id <- paste0(throat_meta$subject, "_", throat_meta$visit)

sputum_meta <- data.frame(sample_id = rownames(sputum_coords))
sputum_meta$subject <- extract_subject_id(sputum_meta$sample_id)
sputum_meta$visit <- extract_visit(sputum_meta$sample_id)
sputum_meta$pair_id <- paste0(sputum_meta$subject, "_", sputum_meta$visit)

subjects <- unique(throat_meta$subject)

results <- lapply(subjects, function(subj) {
  throat_sub <- throat_meta[throat_meta$subject == subj, ]
  sputum_sub <- sputum_meta[sputum_meta$subject == subj, ]

  shared_pair_ids <- intersect(throat_sub$pair_id, sputum_sub$pair_id)
  
  if (length(shared_pair_ids) >= 1) {
    throat_mat <- throat_coords[throat_sub$sample_id[throat_sub$pair_id %in% shared_pair_ids], , drop = FALSE]
    sputum_mat <- sputum_coords[sputum_sub$sample_id[sputum_sub$pair_id %in% shared_pair_ids], , drop = FALSE]

    # make sure order is identical
    throat_mat <- throat_mat[order(rownames(throat_mat)), ]
    sputum_mat <- sputum_mat[order(rownames(sputum_mat)), ]

    proc <- protest(throat_mat, sputum_mat)
    return(data.frame(subject = subj, r = proc$t0, p = proc$signif))
  } else {
    return(NULL)
  }
})

throat_sputum_results <- do.call(rbind, results)
print(throat_sputum_results)
```
# compare sputum-throat to sputum-gut results
```{r}
gut_sputum_results$comparison <- "gut_sputum"
throat_sputum_results$comparison <- "throat_sputum"

combined <- rbind(gut_sputum_results, throat_sputum_results)

# Merge by subject to keep only shared subjects
paired_data <- merge(gut_sputum_results, throat_sputum_results, by = "subject", suffixes = c("_gut", "_throat"))

# Paired t-test (or Wilcoxon if non-normal)
wilcox.test(paired_data$r_throat, paired_data$r_gut, paired = TRUE)

library(ggplot2)

ggplot(paired_data, aes(x = r_gut, y = r_throat, color=subject)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(x = "Gut–Sputum Procrustes r", y = "Throat–Sputum Procrustes r",
       title = "Per-Subject Comparison of Procrustes Correlations")

```

# Conclusion:

Quick interpretation of Procrustes protest results:
Comparison	Sum of Squares	Correlation (r)	p-value	Interpretation
Throat vs Sputum	0.8172	0.4275	0.042	Significant alignment (p < 0.05) → evidence for similarity
Gut vs Sputum	0.8424	0.3970	0.103	Not significant (p > 0.05) → weaker or noisier alignment
Biological read:

→ The throat microbiome is significantly more similar to sputum than the gut microbiome is.
→ This is biologically intuitive: the throat and sputum (lung) are physically closer, and microbial exchange is plausible (aspiration, mucus flow).

Gut-lung interactions are more complex — maybe indirect, systemic (immune-mediated), or only affecting specific taxa.

But cave: P-values are decreasing the more axes I include in the rotations, but overall I have the feeling the results differ too little between stool-sputum and throat-sputum comparisons. 

# Mantel test:
```{r}
sputum_common  <- subset_samples(ps_Sputum_Stool_full_unique_2_relab, material=="Sputum")
gut_common<- subset_samples(ps_Sputum_Stool_full_unique_2_relab, material=="Stool")
# Create distance matrices (e.g., Bray-Curtis)
gut_dist <- phyloseq::distance(gut_common, method = "bray")
sputum_dist <- phyloseq::distance(sputum_common, method = "bray")

mantel(gut_dist, sputum_dist, permutations = 999)

sputum_common  <- subset_samples(ps_sputum_throat_full_unique_2_relab, material=="Sputum")
throat_common<- subset_samples(ps_sputum_throat_full_unique_2_relab, material=="Throat")

# Create distance matrices (e.g., Bray-Curtis)
throat_dist <- phyloseq::distance(throat_common, method = "bray")
sputum_dist <- phyloseq::distance(sputum_common, method = "bray")

mantel(throat_dist, sputum_dist, permutations = 999)
```
# Interpretaion of mantel test: 
→ Mantel r is a correlation between the structure of distance matrices — so:

    "Do samples that are similar in gut also tend to be similar in sputum?"

→ Here, both r values are very small (~0.07 - 0.10), meaning:

    There's only a very weak correspondence between the community structure of gut vs. sputum or throat vs. sputum.

→ p-values ~0.10 → not formally significant, but some weak trend.

# Try entenglement plot
```{r}
library(phyloseq)
library(dendextend)

sputum_common  <- subset_samples(ps_Sputum_Stool_full_unique_2_relab, material=="Sputum")
gut_common<- subset_samples(ps_Sputum_Stool_full_unique_2_relab, material=="Stool")
# Create distance matrices (e.g., Bray-Curtis)
gut_dist <- phyloseq::distance(gut_common, method = "bray")
sputum_dist <- phyloseq::distance(sputum_common, method = "bray")

# Cluster gut
gut_clust <- hclust(gut_dist)
gut_dend <- as.dendrogram(gut_clust)

# Cluster sputum
sputum_clust <- hclust(sputum_dist)
sputum_dend <- as.dendrogram(sputum_clust)

# Match labels
labels(gut_dend) <- sample_data(gut_common)$id_visit
labels(sputum_dend) <- sample_data(sputum_common)$id_visit

ggplot(gut_dend)
ggplot(sputum_dend)

# Plot tanglegram
dend_list <- dendlist(gut_dend, sputum_dend)
tanglegram(gut_dend, sputum_dend,
  highlight_distinct_edges = TRUE, # Turn-off dashed lines
  common_subtrees_color_lines = FALSE, # Turn-off line colors
  common_subtrees_color_branches = TRUE, # Color common branches 
  main = paste("entanglement =", round(entanglement(dend_list), 2)))

```

```{r}
sputum_common  <- subset_samples(ps_sputum_throat_full_unique_2_relab, material=="Sputum")
throat_common<- subset_samples(ps_sputum_throat_full_unique_2_relab, material=="Throat")
# Create distance matrices (e.g., Bray-Curtis)
throat_dist <- phyloseq::distance(throat_common, method = "bray")
sputum_dist <- phyloseq::distance(sputum_common, method = "bray")

# Cluster throat
throat_clust <- hclust(throat_dist)
throat_dend <- as.dendrogram(throat_clust)

# Cluster sputum
sputum_clust <- hclust(sputum_dist)
sputum_dend <- as.dendrogram(sputum_clust)

# Match labels
labels(throat_dend) <- sample_data(throat_common)$id_visit
labels(sputum_dend) <- sample_data(sputum_common)$id_visit

ggplot(throat_dend)
ggplot(sputum_dend)

# Plot tanglegram
dend_list <- dendlist(throat_dend, sputum_dend)
tanglegram(throat_dend, sputum_dend,
  highlight_distinct_edges = TRUE, # Turn-off dashed lines
  common_subtrees_color_lines = FALSE, # Turn-off line colors
  common_subtrees_color_branches = TRUE, # Color common branches 
  main = paste("entanglement =", round(entanglement(dend_list), 2)))
```

# Do co-correlation analysis
```{r}
sputum_common  <- subset_samples(ps_sputum_throat_full_unique_2_relab, material=="Sputum")
throat_common<- subset_samples(ps_sputum_throat_full_unique_2_relab, material=="Throat")

for(taxon in taxa) {
   cor(throat_common[taxon,], sputum_common[taxon,], method="spearman")
}

```
```{r}
# Extract OTU/ASV table

ps_sputum_throat_fil  <- phyloseq::filter_taxa(ps_sputum_throat_full_unique_2_relab, function(x) sum(x) > 0, prune = T)

otu_mat <- as(otu_table(ps_sputum_throat_fil), "matrix")

# Ensure taxa are in rows
if(!taxa_are_rows(ps_sputum_throat_fil)){
  otu_mat <- t(otu_mat)
}

# Get sample data
sample_meta <- sample_data(ps_sputum_throat_fil)

# Subset to sputum and throat samples
sputum_samples <- rownames(sample_meta[sample_meta$material == "Sputum", ])
throat_samples <- rownames(sample_meta[sample_meta$material == "Throat", ])

# Only keep taxa present in both sample sets
taxa_shared <- intersect(rownames(otu_mat), taxa)



```
```{r}
library(purrr)
sputum_common  <- subset_samples(ps_sputum_throat_fil, material=="Sputum")
throat_common<- subset_samples(ps_sputum_throat_fil, material=="Throat")

results <- map_dfr(taxa_names(ps_sputum_throat_fil), function(taxon) {
  
  if(taxa_are_rows(throat_common)) {
    throat_abund <- as.numeric(otu_table(throat_common)[taxon, ])
    sputum_abund <- as.numeric(otu_table(sputum_common)[taxon, ])
  } else {
    throat_abund <- as.numeric(otu_table(throat_common)[, taxon])
    sputum_abund <- as.numeric(otu_table(sputum_common)[, taxon])
  }
  
  valid_idx <- which(!is.na(throat_abund) & !is.na(sputum_abund))
  
  tibble(
    taxon = taxon,
    spearman_rho = if(length(valid_idx) >= 2) cor(throat_abund[valid_idx], sputum_abund[valid_idx], method = "spearman") else NA
  )
})

results_woNA <- results %>% 
  filter(!is.na(spearman_rho))

# Pivot to wide format
corr_mat <- results_woNA %>% 
  pivot_wider(
    names_from = taxon, 
    values_from = spearman_rho
  ) %>% 
  select(-taxon) %>% # if taxon is duplicated
  as.matrix()


# Now you can plot it
corrplot::corrplot(corr_mat, method = "color", 
                   tl.cex = 0.6, 
                   na.label = " ", 
                   col = colorRampPalette(c("blue", "white", "red"))(200))

ggplot(results_woNA, aes(x = reorder(taxon, spearman_rho), y = spearman_rho)) +
  geom_col() +
  coord_flip() +
  theme_minimal()

corr_mat <- results_woNA %>% 
  column_to_rownames("taxon") %>% 
  as.matrix()

corrplot::corrplot(t(corr_mat), is.corr = FALSE, method = "color")


df <- psmelt(ps_sputum_throat_fil)

df$OTU

df_c <- df %>% 
  left_join(results_woNA, by= c("OTU"="taxon"))


df_c %>% 
  group_by(material, OTU) %>% 
  filter(spearman_rho >= abs(0.25)) %>% 
ggplot(aes(y = OTU, x=material, size = Abundance, color = spearman_rho)) +
  geom_point(alpha = 0.7) +  # Adjust the transparency for better visibility
  scale_size_continuous(range = c(1, 5)) +  # Adjust bubble size range
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +  # Color scale for correlation
  facet_wrap(~ Genus, scales = "free") +  # Facet by material
  theme_minimal() +
  labs(
    title = "Bubble Plot of ASV Abundance and Correlation by Material",
    x = "ASV",
    y = "Material",
    size = "Abundance",
    color = "Spearman Correlation"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),  # Rotate taxon labels
    panel.grid.major = element_blank(),  # Remove gridlines for clarity
    panel.grid.minor = element_blank()
  )

df_c_sputum <- df_c %>% 
  filter(material=="Sputum") %>% 
  mutate(sputum_abundance = Abundance) %>% 
  select(OTU, Genus, id_visit, sputum_abundance, spearman_rho)%>% 
  mutate(unique_id = paste(OTU, id_visit, sep = "_"))

df_c_throat <- df_c %>% 
  filter(material=="Throat") %>% 
  mutate(throat_abundance= Abundance)%>% 
  select(OTU, Genus, id_visit, throat_abundance, spearman_rho) %>% 
  mutate(unique_id = paste(OTU, id_visit, sep = "_"))

df_thr_spu <- df_c_sputum %>% 
   left_join(df_c_throat, by= "unique_id" )

df_thr_spu$spearman_rho.x==df_thr_spu$spearman_rho.y

df_thr_spu %>% 
  filter(!is.na(spearman_rho.x)) %>% 
  filter(!is.na(spearman_rho.y)) %>% 
  group_by(OTU.x) %>% 
  ggplot(aes(x = sputum_abundance, y = throat_abundance, color = spearman_rho.x)) +
  geom_point(alpha = 1, size = 1) + 
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Abundance vs Correlation of Shared ASVs Between Body Sites",
    x = "Sputum Abundance",
    y = "Throat Abundance",
    color = "Spearman Correlation"
  ) +
  theme_minimal()+
  scale_x_log10()+
  scale_y_log10()

df_thr_spu %>% 
  filter(!is.na(spearman_rho.x)) %>% 
  filter(!is.na(spearman_rho.y)) %>% 
  filter(spearman_rho.y>=abs(0.2)) %>% 
  ggplot(aes(x = sputum_abundance, y = throat_abundance, color = Genus.x)) +
  geom_point(aes(size=spearman_rho.y), alpha = 1) + 
  #scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Abundance vs Correlation of Shared ASVs Between Body Sites",
    x = "Sputum Abundance",
    y = "Throat Abundance",
    size = "Spearman Correlation"
  ) +
  theme_minimal()+
  scale_x_log10()+
  scale_y_log10()+
  theme(legend.position = "none")
```


# 
```{r}
# Extract OTU/ASV table

ps_sputum_stool_fil  <- phyloseq::filter_taxa(ps_Sputum_Stool_full_unique_2_relab, function(x) sum(x) > 0, prune = T)

sputum_common  <- subset_samples(ps_sputum_stool_fil, material=="Sputum")
stool_common<- subset_samples(ps_sputum_stool_fil, material=="Stool")

results <- map_dfr(taxa_names(ps_sputum_stool_fil), function(taxon) {
  
  if(taxa_are_rows(stool_common)) {
    stool_abund <- as.numeric(otu_table(stool_common)[taxon, ])
    sputum_abund <- as.numeric(otu_table(sputum_common)[taxon, ])
  } else {
    stool_abund <- as.numeric(otu_table(stool_common)[, taxon])
    sputum_abund <- as.numeric(otu_table(sputum_common)[, taxon])
  }
  
  valid_idx <- which(!is.na(stool_abund) & !is.na(sputum_abund))
  
  tibble(
    taxon = taxon,
    spearman_rho = if(length(valid_idx) >= 2) cor(stool_abund[valid_idx], sputum_abund[valid_idx], method = "spearman") else NA
  )
})

results_woNA <- results %>% 
  filter(!is.na(spearman_rho))

# Pivot to wide format
corr_mat <- results_woNA %>% 
  pivot_wider(
    names_from = taxon, 
    values_from = spearman_rho
  ) %>% 
  select(-taxon) %>% # if taxon is duplicated
  as.matrix()


# Now you can plot it
corrplot::corrplot(corr_mat, method = "color", 
                   tl.cex = 0.6, 
                   na.label = " ", 
                   col = colorRampPalette(c("blue", "white", "red"))(200))

ggplot(results_woNA, aes(x = reorder(taxon, spearman_rho), y = spearman_rho)) +
  geom_col() +
  coord_flip() +
  theme_minimal()

corr_mat <- results_woNA %>% 
  column_to_rownames("taxon") %>% 
  as.matrix()

corrplot::corrplot(t(corr_mat), is.corr = FALSE, method = "color")


df <- psmelt(ps_sputum_stool_fil)

df$OTU

df_c <- df %>% 
  left_join(results_woNA, by= c("OTU"="taxon"))


df_c %>% 
  group_by(material, OTU) %>% 
  filter(spearman_rho >= abs(0.25)) %>% 
ggplot(aes(y = OTU, x=material, size = Abundance, color = spearman_rho)) +
  geom_point(alpha = 0.7) +  # Adjust the transparency for better visibility
  scale_size_continuous(range = c(1, 5)) +  # Adjust bubble size range
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +  # Color scale for correlation
  facet_wrap(~ Genus, scales = "free") +  # Facet by material
  theme_minimal() +
  labs(
    title = "Bubble Plot of ASV Abundance and Correlation by Material",
    x = "ASV",
    y = "Material",
    size = "Abundance",
    color = "Spearman Correlation"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),  # Rotate taxon labels
    panel.grid.major = element_blank(),  # Remove gridlines for clarity
    panel.grid.minor = element_blank()
  )

df_c_sputum <- df_c %>% 
  filter(material=="Sputum") %>% 
  mutate(sputum_abundance = Abundance) %>% 
  select(OTU, Genus, id_visit, sputum_abundance, spearman_rho)%>% 
  mutate(unique_id = paste(OTU, id_visit, sep = "_"))

df_c_stool <- df_c %>% 
  filter(material=="Stool") %>% 
  mutate(stool_abundance= Abundance)%>% 
  select(OTU, Genus, id_visit, stool_abundance, spearman_rho) %>% 
  mutate(unique_id = paste(OTU, id_visit, sep = "_"))

df_thr_spu <- df_c_sputum %>% 
   left_join(df_c_stool, by= "unique_id" )

df_thr_spu$spearman_rho.x==df_thr_spu$spearman_rho.y

df_thr_spu %>% 
  filter(!is.na(spearman_rho.x)) %>% 
  filter(!is.na(spearman_rho.y)) %>% 
  filter(spearman_rho.y>=abs(0.2)) %>% 
  ggplot(aes(x = sputum_abundance, y = stool_abundance, color = Genus.x)) +
  geom_point(aes(size=spearman_rho.y), alpha = 1) + 
  #scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Abundance vs Correlation of Shared ASVs Between Body Sites",
    x = "Sputum Abundance",
    y = "Stool Abundance",
    color = "Spearman Correlation"
  ) +
  theme_minimal()+
  scale_x_log10()+
  scale_y_log10()+
  theme(legend.position = "none")
```
