---
title: "starplot stool including Controls"
author: "Rebecca L. Knoll and Theda Bartholomaeus"
date: " last edit `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: show
    number_sections: yes
    smart: no
    toc: yes
    df_print: paged
    toc_float: yesn
---

```{r setup 2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache= FALSE, warning = FALSE)
```

```{r run import script, include=FALSE}
# run script to tidy the data and to load packages, create phyloseq object (ps_clean) and load libraries
ps_full <- readRDS("/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/rds_files/ps_IMProveCF_patientsAndControls_Run1-23_18102023.rds")

pacman::p_load(tidyverse, phyloseq, magrittr, janitor, microbiome, knitr, lubridate, naniar, readxl, ggplot2, ggpubr)

library(datarium)
library(dplyr)
#library(ggplot2)

visit_sum_palette <- c("black", "#C3BC3FFF", "#6388B4FF", "#BB7693FF", "#55AD89FF", "#8176AA")

# Create a named vector for shape labels
shape_labels <- c("0", "3", "6-12", "15-18", "21-24 months from ETI tt start")
```

```{r subset per material}

# subset per material
ps_stool <- subset_samples(ps_full, material== "Stool")

# remove zero abundances from each dataset
ps_stool <- microViz::tax_filter(ps_stool, min_prevalence = 1,
  prev_detection_threshold = 1, min_total_abundance = 0, min_sample_abundance = 0, tax_level = NA,
  names_only = FALSE, use_counts = TRUE, undetected = NULL,verbose = TRUE)

# calculate relative abundances
ps_stool_relab <- transform_sample_counts(ps_stool, function(x) x/sum(x))

```

```{r}
PCA_all_visit_stool <- plot_ordination(ps_stool_relab, ordinate(ps_stool_relab, "MDS"), color = "visit_sum")
```


# Starplot for PCoA 

## stool

```{r calculate distances and prepare data}
metadata_stool <- (sample_data(ps_stool_relab))
metadata_stool <- as(metadata_stool, "data.frame")

##### Create dissimilarity indices
set.seed(100)
BC_dist<- phyloseq::distance(ps_stool_relab,
                             method="bray", weighted=F)

bray <- as.matrix(BC_dist)
  
# Principal Coordinate Analysis (PCoA)
  beta <- as.data.frame(cmdscale(bray, k = 2))
  
  # Add metadata
  beta <- base::transform(merge(sample_data(ps_stool_relab), beta, by = 0, all = TRUE), row.names = Row.names, Row.names = NULL)
  
beta <- beta%>%
  dplyr::select(id.x, visit_cal_cor, sex, age_y, visit_sum, V1, V2)

# Ensure "Control" is a level in the factor
beta$visit_cal_cor <- factor(beta$visit_cal_cor, levels = c(levels(beta$visit_cal_cor), "Control"))

# Replace NA values with "Control"
beta$visit_cal_cor[is.na(beta$visit_cal_cor)] <- "Control"

```



```{r starplot stool Teil C}

time_a <- beta[beta$visit_sum == "1", ]
time_f <- beta[beta$visit_sum == "Control", ]

# Add a constant key column to both data frames for the Cartesian join
time_a$key <- 1
time_f$key <- 1

# Perform the Cartesian join
all <- merge(time_a, time_f, by = "key", suffixes = c("", ".f"))

# for each donor, center PCoA Dims to the Upper lobe samples
center.2 <- all
for (i in c("V1","V1.f")){
  center.2[,i] <- (all[, i] - all[, "V1"])
}
for (i in c("V2","V2.f")){
  center.2[,i] <- (all[, i] - all[, "V2"])
}
center.2 <- center.2[, c("V1","V1.f","V2","V2.f", "key")]

#center.v <- na.exclude(center.v) # nur Donoren die alle swabsite haben

# get mean of the to arrow categories (arrows to lower lobe, arrows to stool)
arrowCentroids.2 <- as.data.frame(matrix(c(mean(na.exclude(center.2$V1.f)),
                                           mean(na.exclude(center.2$V2.f)), 
                                        
                                           center.2$V1[1], center.2$V2[1]), ncol = 4))

# plot centered 
#plot_all.v <- ggplot(center.v, aes(x = V1, y = V2)) +

plot_all <- ggplot(center.2, aes(x = V1, y = V2)) +
  geom_point(data = center.2, aes(x = V1, y = V2), size = 4, alpha = 0.8) + 
  theme_classic() +
  #geom_point(data = center, aes(x = V1.b, y = V2.b),color = "#C3BC3FFF", pch=17, size = 3, alpha = 0.8) +
  #geom_point(data = center, aes(x = V1.c, y = V2.c),color = "#6388B4FF", pch=15, size = 3, alpha = 0.8) +
  scale_x_continuous(limits = c(-0.4,0.4))+
  #scale_y_continuous(breaks = c( -0.50, -0.25, 0.00,0.25, 0.5), limits = c(-0.50, 0.50))+
  geom_segment(data = center.2, color = "#8175AA",
          aes(x = V1,
         y = V2,
         xend = V1.f,
          yend = V2.f,
        color = "a to c", alpha = .1),
         arrow = arrow (length = unit(0.3,"cm"))) +
  geom_point(data = arrowCentroids.2, aes(x = V1, y = V2), pch=16, color = "black", size = 6) +
  geom_point(data = arrowCentroids.2, aes(x = V3, y = V4), pch=18, color = "black", size = 6) +
  geom_point(data = arrowCentroids.2, aes(x = V1, y = V2), pch=16, color = "#8175AA", size = 5) +
  geom_point(data = arrowCentroids.2, aes(x = V3, y = V4), pch=18, color = "#8175AA", size = 5) +
  geom_segment(data = arrowCentroids.2,
               aes(x = V3,
                   y = V4,
                   xend = V1,
                  yend = V2,
                  color = "mean"),
               arrow = arrow (length = unit(0.3,"cm")), color = "black") +
  xlab ("PCo 1") +
  ylab ("PCo 2") +
  guides(color = guide_legend(title = "trajectory from baseline to")) +
  theme (axis.title.x = element_text (size = 16), 
         axis.text.x = element_text (size = 16), 
         axis.text.y = element_text (size = 16), 
         axis.title.y = element_text (size = 16),
         legend.position = "bottom",
         title = element_text(size=18))+
  ggtitle("Stool")-> plot.controls

print(plot.controls)
```

```{r starplot stool Teil A}
#write_csv(beta, "~/Documents/Forschung/IMMProveCF/R_analysis/data/beta_starplot.csv")

time_a <- beta[beta$visit_sum == "1", ]
time_b <- beta[beta$visit_sum == "2", ]
time_c <- beta[beta$visit_sum == "3-5", ]

# we will toss a - since there not much information 
# we will put c to 0 so we can see the trajectory 
# join df zu einem df 
all <- merge(time_b, time_c, by = "id.x", all = TRUE, suffixes = c(".b",".c"))
all <- merge(all, time_a, by = "id.x", all = TRUE)

# for each donor, center PCoA Dims to the Upper lobe samples
center <- all
for (i in c("V1","V1.b","V1.c")){
  center[,i] <- (all[, i] - all[, "V1"])
}
for (i in c("V2","V2.b","V2.c")){
  center[,i] <- (all[, i] - all[, "V2"])
}
center <- center[, c("V1","V1.b","V1.c","V2","V2.b","V2.c", "id.x")]

pal.treat  <- c("#803300ff", "#822bc1ff")
# center <- na.exclude(center) # nur Donoren die alle swabsite haben

# get mean of the to arrow categories (arrows to lower lobe, arrows to stool)
arrowCentroids <- as.data.frame(matrix(c(mean(na.exclude(center$V1.b)),
                                         mean(na.exclude(center$V2.b)), 
                                         mean(na.exclude(center$V1.c)),
                                         mean(na.exclude(center$V2.c)),
                                         center$V1[1], center$V2[1]), ncol = 6))

# plot centered 
plot.controls +
  scale_x_continuous(limits = c(-0.4,0.4))+
  #scale_y_continuous(breaks = c( -0.50, -0.25, 0.00,0.25, 0.5), limits = c(-0.50, 0.50))+
  geom_segment(data = center,color ="#C3BC3FFF",
               aes(x = V1,
                  y = V2,
                   xend = V1.b,
                   yend = V2.b,
                   color = "a to b", alpha = .2),
               arrow = arrow (length = unit(0.3,"cm"))) +
  geom_segment(data = center, color = "#6388B4FF",
              aes(x = V1,
                y = V2,
                 xend = V1.c,
                 yend = V2.c,
                color = "a to c", alpha = .2),
               arrow = arrow (length = unit(0.3,"cm"))) +
  geom_point(data = arrowCentroids, aes(x = V1, y = V2),pch=17, color = "black", size = 6) +
  geom_point(data = arrowCentroids, aes(x = V3, y = V4), pch=15, color = "black", size = 6) +
  geom_point(data = arrowCentroids, aes(x = V5, y = V6), color = "black", size = 2) +
  geom_point(data = arrowCentroids, aes(x = V1, y = V2),pch=17, color = "#C3BC3FFF", size = 5) +
  geom_point(data = arrowCentroids, aes(x = V3, y = V4), pch=15, color = "#6388B4FF", size = 5) +
  geom_segment(data = arrowCentroids,
               aes(x = V5,
                   y = V6,
                   xend = V3,
                   yend = V4,
                   color = "mean"),
               arrow = arrow (length = unit(0.3,"cm")), color = "black") +
  geom_segment(data = arrowCentroids,
               aes(x = V5,
                   y = V6,
                   xend = V1,
                   yend = V2,
                   color = "mean"),
               arrow = arrow (length = unit(0.3,"cm")), color ="black") +
  xlab ("PCo 1") +
  ylab ("PCo 2") +
  guides(color = guide_legend(title = "trajectory")) +
  # scale_color_manual(values = c("#4DBBD5FF", "#00A087FF", "black"))+
  
  theme (axis.title.x = element_text (size = 13), 
         axis.text.x = element_text (size = 13), 
         axis.text.y = element_text (size = 13), 
         axis.title.y = element_text (size = 13))->plot.2

print(plot.2)
```

```{r starplot stool Teil B}

time_a <- beta[beta$visit_sum == "1", ]
time_d <- beta[beta$visit_sum == "6-7", ]
time_e <- beta[beta$visit_sum == "8-10", ]

# we will toss a - since there not much information 
# we will put c to 0 so we can see the trajectory 
# join df zu einem df 
all <- merge(time_d, time_e, by = "id.x", all = TRUE, suffixes = c(".d",".e"))
all <- merge(all, time_a, by = "id.x", all = TRUE)

# for each donor, center PCoA Dims to the Upper lobe samples
center.2 <- all
for (i in c("V1","V1.d","V1.e")){
  center.2[,i] <- (all[, i] - all[, "V1"])
}
for (i in c("V2","V2.d","V2.e")){
  center.2[,i] <- (all[, i] - all[, "V2"])
}
center.2 <- center.2[, c("V1","V1.d","V1.e","V2","V2.d","V2.e", "id.x")]



#center.v <- na.exclude(center.v) # nur Donoren die alle swabsite haben

# get mean of the to arrow categories (arrows to lower lobe, arrows to stool)
arrowCentroids.2 <- as.data.frame(matrix(c(mean(na.exclude(center.2$V1.d)),
                                           mean(na.exclude(center.2$V2.d)), 
                                           mean(na.exclude(center.2$V1.e)),
                                           mean(na.exclude(center.2$V2.e)),
                                           center.2$V1[1], center.2$V2[1]), ncol = 6))

# plot centered 
#plot_all.v <- ggplot(center.v, aes(x = V1, y = V2)) +


plot.2+
  geom_point(data = center.2, aes(x = V1, y = V2), size = 5, alpha = 0.8) + 
  theme_classic() +
  #geom_point(data = center.2, aes(x = V1.d, y = V2.d),pch=16, color = "#BB7693FF", size = 3, alpha = 0.8) +
  #geom_point(data = center.2, aes(x = V1.e, y = V2.e), pch=18,color = "#55AD89FF", size = 3, alpha = 0.8) +
  scale_x_continuous(limits = c(-0.4,0.4))+
  # scale_y_continuous(breaks = c( -0.50, -0.25, 0.00,0.25, 0.5), limits = c(-0.50, 0.50))+
  geom_segment(data = center.2, color = "#BB7693FF", 
             aes(x = V1,
             y = V2,
             xend = V1.d,
             yend = V2.d,
             color = "a to b", alpha = .15),
             arrow = arrow (length = unit(0.3,"cm"))) +
  geom_segment(data = center.2, color = "#55AD89FF",
          aes(x = V1,
         y = V2,
          xend = V1.e,
          yend = V2.e,
         color = "a to c", alpha = .15),
          arrow = arrow (length = unit(0.3,"cm"))) +
  geom_point(data = arrowCentroids.2, aes(x = V1, y = V2), pch=16, color = "black", size = 6) +
  geom_point(data = arrowCentroids.2, aes(x = V3, y = V4), pch=18, color = "black", size = 6) +
  geom_point(data = arrowCentroids.2, aes(x = V5, y = V6), color = "black", size = 2) +
  geom_point(data = arrowCentroids.2, aes(x = V1, y = V2), pch=16, color = "#BB7693FF", size = 5) +
  geom_point(data = arrowCentroids.2, aes(x = V3, y = V4), pch=18, color = "#55AD89FF", size = 5) +
  geom_segment(data = arrowCentroids.2,
               aes(x = V5,
                   y = V6,
                   xend = V3,
                   yend = V4,
                   color = "mean"),
               arrow = arrow (length = unit(0.3,"cm")), color = "black") +
  geom_segment(data = arrowCentroids.2,
               aes(x = V5,
                   y = V6,
                   xend = V1,
                   yend = V2,
                   color = "mean"),
               arrow = arrow (length = unit(0.3,"cm")), color= "black") +
  xlab ("PCo 1") +
  ylab ("PCo 2") +
  guides(color = guide_legend(title = "trajectory from baseline to")) +
  theme (axis.title.x = element_text (size = 16), 
         axis.text.x = element_text (size = 16), 
         axis.text.y = element_text (size = 16), 
         axis.title.y = element_text (size = 16),
         legend.position = "none",
         title = element_text(size=18))+
  ggtitle("Stool")-> plot.plot_stool

print(plot.plot_stool)
#saveRDS(plot.plot_stool, "/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/starplot_stool.rds")
```

# now inverse the plot so that Controls are in the center

```{r}
# Create subsets based on 'visit_cal_cor'
time_a <- beta[beta$visit_cal_cor == "Control", ]
time_b <- beta[beta$visit_cal_cor == "1", ]
time_c <- beta[beta$visit_cal_cor == "2", ]
time_d <- beta[beta$visit_cal_cor == "3", ]
time_e <- beta[beta$visit_cal_cor == "4", ]
time_f <- beta[beta$visit_cal_cor == "5", ]
time_g <- beta[beta$visit_cal_cor == "6", ]
time_h <- beta[beta$visit_cal_cor == "7", ]
time_i <- beta[beta$visit_cal_cor == "8", ]
time_j <- beta[beta$visit_cal_cor == "9", ]

# Add a constant key column for Cartesian joins
time_a$key <- 1
time_b$key <- 1
time_a$key.2 <- 2
time_c$key.2 <- 2
time_a$key.3 <- 3
time_d$key.3 <- 3
time_a$key.4 <- 4
time_e$key.4 <- 4
time_a$key.5 <- 5
time_f$key.5 <- 5
time_a$key.6 <- 6
time_g$key.6 <- 6
time_a$key.7 <- 7
time_h$key.7 <- 7
time_a$key.8 <- 8
time_i$key.8 <- 8
time_a$key.9 <- 9
time_j$key.9 <- 9

# Perform the Cartesian joins
ab <- merge(time_a, time_b, by = "key", all = TRUE, suffixes = c("", ".b")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))
ac <- merge(time_a, time_c, by = "key.2", all = TRUE, suffixes = c("", ".c")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))
ad <- merge(time_a, time_d, by = "key.3", all = TRUE, suffixes = c("", ".d")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))
ae <- merge(time_a, time_e, by = "key.4", all = TRUE, suffixes = c("", ".e")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))
af <- merge(time_a, time_f, by = "key.5", all = TRUE, suffixes = c("", ".f")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))
ag <- merge(time_a, time_g, by = "key.6", all = TRUE, suffixes = c("", ".g")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))
ah <- merge(time_a, time_h, by = "key.7", all = TRUE, suffixes = c("", ".h")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))
ai <- merge(time_a, time_i, by = "key.8", all = TRUE, suffixes = c("", ".i")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))
aj <- merge(time_a, time_j, by = "key.9", all = TRUE, suffixes = c("", ".j")) %>%
  dplyr::select(-c(key, key.2, key.3, key.4, key.5, key.6, key.7, key.8, key.9))

# Compute differences for V1 and V2 for each merged dataset and dplyr::select relevant columns

# For ab
ab_center <- ab
ab_center$V1_diff <- ab$V1.b - ab$V1
ab_center$V2_diff <- ab$V2.b - ab$V2
ab_center <- ab_center[, c("V1", "V1.b", "V1_diff", "V2", "V2.b", "V2_diff")]

# For ac
ac_center <- ac
ac_center$V1_diff <- ac$V1.c - ac$V1
ac_center$V2_diff <- ac$V2.c - ac$V2
ac_center <- ac_center[, c("V1", "V1.c", "V1_diff", "V2", "V2.c", "V2_diff")]

# For ad
ad_center <- ad
ad_center$V1_diff <- ad$V1.d - ad$V1
ad_center$V2_diff <- ad$V2.d - ad$V2
ad_center <- ad_center[, c("V1", "V1.d", "V1_diff", "V2", "V2.d", "V2_diff")]

# For ae
ae_center <- ae
ae_center$V1_diff <- ae$V1.e - ae$V1
ae_center$V2_diff <- ae$V2.e - ae$V2
ae_center <- ae_center[, c("V1", "V1.e", "V1_diff", "V2", "V2.e", "V2_diff")]

# For af
af_center <- af
af_center$V1_diff <- af$V1.f - af$V1
af_center$V2_diff <- af$V2.f - af$V2
af_center <- af_center[, c("V1", "V1.f", "V1_diff", "V2", "V2.f", "V2_diff")]

# For ag
ag_center <- ag
ag_center$V1_diff <- ag$V1.g - ag$V1
ag_center$V2_diff <- ag$V2.g - ag$V2
ag_center <- ag_center[, c("V1", "V1.g", "V1_diff", "V2", "V2.g", "V2_diff")]

# For ah
ah_center <- ah
ah_center$V1_diff <- ah$V1.h - ah$V1
ah_center$V2_diff <- ah$V2.h - ah$V2
ah_center <- ah_center[, c("V1", "V1.h", "V1_diff", "V2", "V2.h", "V2_diff")]

# For ai
ai_center <- ai
ai_center$V1_diff <- ai$V1.i - ai$V1
ai_center$V2_diff <- ai$V2.i - ai$V2
ai_center <- ai_center[, c("V1", "V1_diff", "V2", "V2_diff", "visit_sum.i")]
ai_center$visit_sum <- ai_center$visit_sum.i
ai_center <- ai_center[, c("V1", "V1_diff", "V2", "V2_diff", "visit_sum")]
ai_center$comparison_visit <- "8"

# For aj
aj_center <- aj
aj_center$V1_diff <- aj$V1.j - aj$V1
aj_center$V2_diff <- aj$V2.j - aj$V2
aj_center <- aj_center[, c("V1", "V1_diff", "V2", "V2_diff")]
aj_center$comparison_visit <- "9"
```
```{r}
# Create subsets based on 'visit_cal_cor'
time_a <- beta[beta$visit_cal_cor == "Control", ]
time_b <- beta[beta$visit_cal_cor == "1", ]
time_c <- beta[beta$visit_cal_cor == "2", ]
time_d <- beta[beta$visit_cal_cor == "3", ]
time_e <- beta[beta$visit_cal_cor == "4", ]
time_f <- beta[beta$visit_cal_cor == "5", ]
time_g <- beta[beta$visit_cal_cor == "6", ]
time_h <- beta[beta$visit_cal_cor == "7", ]
time_i <- beta[beta$visit_cal_cor == "8", ]
time_j <- beta[beta$visit_cal_cor == "9", ]

# Add a constant key column for Cartesian joins
time_a$key <- 1
time_b$key <- 1
time_a$key.2 <- 2
time_c$key.2 <- 2
time_a$key.3 <- 3
time_d$key.3 <- 3
time_a$key.4 <- 4
time_e$key.4 <- 4
time_a$key.5 <- 5
time_f$key.5 <- 5
time_a$key.6 <- 6
time_g$key.6 <- 6
time_a$key.7 <- 7
time_h$key.7 <- 7
time_a$key.8 <- 8
time_i$key.8 <- 8
time_a$key.9 <- 9
time_j$key.9 <- 9

# Perform the Cartesian joins and compute differences
calculate_center <- function(time_a, time_other, suffix, key_col, comparison_visit) {
  merged <- merge(time_a, time_other, by = key_col, all = TRUE, suffixes = c("", suffix)) %>%
    dplyr::select(-contains("key"))
  center <- merged
  center$V1_diff <- merged[[paste0("V1", suffix)]] - merged$V1
  center$V2_diff <- merged[[paste0("V2", suffix)]] - merged$V2
  center$visit_sum <- merged[[paste0("visit_sum", suffix)]]
  center$comparison_visit <- comparison_visit
  center <- center[, c("V1", "V1_diff", "V2", "V2_diff", "visit_sum", "comparison_visit")]
  return(center)
}

ab_center <- calculate_center(time_a, time_b, ".b", "key", "1")
ac_center <- calculate_center(time_a, time_c, ".c", "key.2", "2")
ad_center <- calculate_center(time_a, time_d, ".d", "key.3", "3")
ae_center <- calculate_center(time_a, time_e, ".e", "key.4", "4")
af_center <- calculate_center(time_a, time_f, ".f", "key.5", "5")
ag_center <- calculate_center(time_a, time_g, ".g", "key.6", "6")
ah_center <- calculate_center(time_a, time_h, ".h", "key.7", "7")
ai_center <- calculate_center(time_a, time_i, ".i", "key.8", "8")
aj_center <- calculate_center(time_a, time_j, ".j", "key.9", "9")

# Combine all the center dataframes
all_centers <- rbind(
  ab_center,
  ac_center,
  ad_center,
  ae_center,
  af_center,
  ag_center,
  ah_center,
  ai_center,
  aj_center
)

# Set V1 and V2 from time_a (control) to 0
all_centers$V1 <- 0
all_centers$V2 <- 0

# View the final combined dataframe
print(all_centers)

```
```{r}
# calculate the means of bc distances per visit_sum
means <- all_centers %>% 
  group_by(visit_sum) %>% 
  summarise(Mean_V1=median(V1_diff), Mean_V2=median(V2_diff))

means$center_v1 <- 0
means$center_v2 <- 0
```

# plot arrows
```{r, fig.height=5, fig.width=5}
visit_sum_palette <- c("black", "#C3BC3FFF", "#6388B4FF", "#BB7693FF", "#55AD89FF", "#8176AA")

# Create a named vector for shape labels
shape_labels <- c("0", "3", "6-12", "15-18", "21-24 months from ETI tt start")

p1 <-ggplot(all_centers, aes(x = V1, y = V2)) +
  geom_point(data = all_centers, aes(x = V1, y = V2), size = 6,color="#8176AA", shape=8) + 
  geom_segment(aes(x = V1,
                  y = V2,
                   xend = V1_diff,
                   yend = V2_diff,
                   color = visit_sum), alpha = 0.045,
               arrow = arrow (length = unit(0.3,"cm")))+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  scale_fill_manual(values = visit_sum_palette, labels=shape_labels)+
  scale_shape_manual(values = c(21:25),labels=shape_labels)+
  geom_point(data = means, aes(x = Mean_V1, y = Mean_V2,  fill= visit_sum, shape = visit_sum), color="black", size = 4)+
  # add arrow centroids (mean per visit sum)
  geom_segment(data = means,
               aes(x = center_v1,
                   y = center_v2,
                   xend = Mean_V1,
                   yend = Mean_V2,
                   color = visit_sum),
               arrow = arrow (length = unit(0.4,"cm")), linetype = 2)+
  xlim(-0.6,0.15)+
  ylim(-0.45, 0.4)+
  guides(color="none",shape = guide_legend(title = "Median BC dissimilarity to controls by"),fill = guide_legend(title = "Median BC dissimilarity to controls by"))+
  theme(legend.position = "bottom", legend.title.position = "top",text=element_text(size=20))+
  xlab ("PCo 1") +
  ylab ("PCo 2")
p1
saveRDS(p1,"/Users/rebecca/Documents/Forschung/IMMProveCF/R_analysis/figures/starplot_stool_controls.rds")
```

# plot median arrows only
```{r}
ggplot(all_centers, aes(x = V1, y = V2)) +
  geom_point(data = all_centers, aes(x = V1, y = V2), size = 6, color="#8176AA") + 
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  scale_fill_manual(values = visit_sum_palette)+
  scale_shape_manual(values = c(21:25))+
  geom_point(data = means, aes(x = Mean_V1, y = Mean_V2, color=visit_sum, fill= visit_sum, shape = visit_sum), size = 6)+
  # add arrow centroids (mean per visit sum)
  geom_segment(data = means,
               aes(x = center_v1,
                   y = center_v2,
                   xend = Mean_V1,
                   yend = Mean_V2,
                   color = visit_sum),
               arrow = arrow (length = unit(0.4,"cm")), linetype =                2)+
  xlab ("PCo 1") +
  ylab ("PCo 2") 
```

# create 3 Axes plots
```{r}
##### Create dissimilarity indices
set.seed(100)

# Bray-Curtis distance axis comparisons
BC_dist <- phyloseq::distance(ps_stool_relab, method="bray", weighted=F)
# Perform Principal Coordinate Analysis (PCoA)
pcoa_result <- ape::pcoa(BC_dist)
# Extract eigenvalues
eigenvalues <- pcoa_result$values$Eigenvalues
# Create a scree plot
plot(eigenvalues, type = "b", xlab = "Principal Coordinates", ylab = "Eigenvalue", 
     main = "Scree Plot", pch = 19)

# Calculate total variance
total_variance <- sum(eigenvalues)

# Calculate variance explained by each component
variance_explained <- eigenvalues / total_variance

# Convert to percentage
percentage_explained <- variance_explained * 100
# Optionally, create a scree plot showing the percentage of variance explained
plot(percentage_explained, type = "b", xlab = "Principal Coordinates", ylab = "Percentage of Variance Explained",
     main = "Percentage of Variance Explained", pch = 19)
percentage_explained[1]

bray <- as.matrix(BC_dist)

# MDS comparison
pcoa_result <- ordinate(ps_stool_relab, method = "MDS")
# Extract eigenvalues
eigenvalues <- pcoa_result$values$Eigenvalues
# Create a scree plot
plot(eigenvalues, type = "b", xlab = "Principal Coordinates", ylab = "Eigenvalue", 
     main = "Scree Plot", pch = 19)
# the eigenvalues are exactly the same, it does not matter whether calculated on MDS or bray

# Extract the coordinates from the ordination result
ordination_df <- as.data.frame(pcoa_result$vectors[, 1:3])

# Add metadata from the phyloseq object
ordination_df <- cbind(ordination_df, sample_data(ps_stool_relab))

# PCo1 vs. PCo2
plot1 <- ggplot(ordination_df, aes(x = Axis.1, y = Axis.2, color = visit_sum)) +
  geom_point(size = 4) +
  theme_classic() +
  scale_color_manual(values = visit_sum_palette)+
  xlab("PCo1") +
  ylab("PCo2") +
  ggtitle("PCo1 vs. PCo2")+
  stat_ellipse()

# PCo1 vs. PCo3
plot2 <- ggplot(ordination_df, aes(x = Axis.1, y = Axis.3, color = visit_sum)) +
  geom_point(size = 4) +
  theme_classic() +
  scale_color_manual(values = visit_sum_palette)+
  xlab("PCo1") +
  ylab("PCo3") +
  ggtitle("PCo1 vs. PCo3")+
  stat_ellipse()

# PCo2 vs. PCo3
plot3 <- ggplot(ordination_df, aes(x = Axis.2, y = Axis.3, color = visit_sum)) +
  geom_point(size = 4) +
  theme_classic() +
  scale_color_manual(values = visit_sum_palette)+
  #xlab("PCo2") +
  #ylab("PCo3") +
  ggtitle("PCo2 vs. PCo3")+
  stat_ellipse()

# Display the plots
print(plot1)
print(plot2)
print(plot3)
```
My problem now is that my data has 6 timepoints, and I relate my ordination analysis to timepoint 1, which I put in the center of my ordination plot, does the variance explained remains the same?

Versa
When you perform ordination analysis (such as PCA, PCoA, or NMDS), the variance explained by each axis is determined by the eigenvalues or stress values derived from the overall dataset. This variance explanation is a property of the data structure and its configuration in the chosen ordination method.

If you relate your ordination results to a specific timepoint (e.g., timepoint 1) and place this timepoint in the center of your ordination plot, you are essentially re-centering or possibly rotating the ordination plot. However, this does not change the underlying eigenvalues, stress values, or the total variance explained by the ordination axes. Re-centering can change the visual interpretation or emphasize certain relationships, but the total variance explained by the axes remains constant because it is a property of the dataset as a whole.

So, in summary:

The variance explained by each axis remains the same regardless of re-centering the plot around a specific timepoint.
The visual representation may change, but this does not affect the quantitative measures of variance explained.
# try to create 3 axes plot for my centered analysis
```{r}
# Principal Coordinate Analysis (PCoA)
beta <- as.data.frame(cmdscale(bray, k = 3))  # Extract the first 3 principal coordinates

# Add metadata
beta <- base::transform(merge(sample_data(ps_stool_relab), beta, by = 0, all = TRUE), row.names = Row.names, Row.names = NULL)

beta <- beta%>%
  dplyr::select(id.x, visit_cal_9, sex, age_y, visit_sum, V1, V2, V3)

# Create subsets based on 'visit_cal_9'
time_a <- beta[beta$visit_cal_9 == "Control", ]
time_b <- beta[beta$visit_cal_9 == "1", ]
time_c <- beta[beta$visit_cal_9 == "2", ]
time_d <- beta[beta$visit_cal_9 == "3", ]
time_e <- beta[beta$visit_cal_9 == "4", ]
time_f <- beta[beta$visit_cal_9 == "5", ]
time_g <- beta[beta$visit_cal_9 == "6", ]
time_h <- beta[beta$visit_cal_9 == "7", ]
time_i <- beta[beta$visit_cal_9 == "8", ]
time_j <- beta[beta$visit_cal_9 == "9", ]

# Add a constant key column for Cartesian joins
time_a$key <- 1
time_b$key <- 1
time_a$key.2 <- 2
time_c$key.2 <- 2
time_a$key.3 <- 3
time_d$key.3 <- 3
time_a$key.4 <- 4
time_e$key.4 <- 4
time_a$key.5 <- 5
time_f$key.5 <- 5
time_a$key.6 <- 6
time_g$key.6 <- 6
time_a$key.7 <- 7
time_h$key.7 <- 7
time_a$key.8 <- 8
time_i$key.8 <- 8
time_a$key.9 <- 9
time_j$key.9 <- 9

# Perform the Cartesian joins and compute differences
calculate_center <- function(time_a, time_other, suffix, key_col, comparison_visit) {
  merged <- merge(time_a, time_other, by = key_col, all = TRUE, suffixes = c("", suffix)) %>%
    dplyr::select(-contains("key"))
  center <- merged
  center$V1_diff <- merged[[paste0("V1", suffix)]] - merged$V1
  center$V2_diff <- merged[[paste0("V2", suffix)]] - merged$V2
  center$V3_diff <- merged[[paste0("V3", suffix)]] - merged$V3
  center$visit_sum <- merged[[paste0("visit_sum", suffix)]]
  center$comparison_visit <- comparison_visit
  center <- center[, c("V1", "V1_diff", "V2", "V2_diff", "V3", "V3_diff", "visit_sum", "comparison_visit")]
  return(center)
}

ab_center <- calculate_center(time_a, time_b, ".b", "key", "1")
ac_center <- calculate_center(time_a, time_c, ".c", "key.2", "2")
ad_center <- calculate_center(time_a, time_d, ".d", "key.3", "3")
ae_center <- calculate_center(time_a, time_e, ".e", "key.4", "4")
af_center <- calculate_center(time_a, time_f, ".f", "key.5", "5")
ag_center <- calculate_center(time_a, time_g, ".g", "key.6", "6")
ah_center <- calculate_center(time_a, time_h, ".h", "key.7", "7")
ai_center <- calculate_center(time_a, time_i, ".i", "key.8", "8")
aj_center <- calculate_center(time_a, time_j, ".j", "key.9", "9")

# Combine all the center dataframes
all_centers <- rbind(
  ab_center,
  ac_center,
  ad_center,
  ae_center,
  af_center,
  ag_center,
  ah_center,
  ai_center,
  aj_center
)

# Set V1, V2, and V3 from time_a (control) to 0
all_centers$V1 <- 0
all_centers$V2 <- 0
all_centers$V3 <- 0

# Calculate the means of BC distances per visit_sum
means <- all_centers %>%
  group_by(visit_sum) %>%
  summarise(Mean_V1 = median(V1_diff), Mean_V2 = median(V2_diff), Mean_V3 = median(V3_diff))

means$center_v1 <- 0
means$center_v2 <- 0
means$center_v3 <- 0

# Plot arrows for each pair of coordinates V1 and V2
p1 <- ggplot(all_centers, aes(x = V1, y = V2)) +
  geom_point(aes(x = V1, y = V2), size = 6,color="#8176AA", shape=8) + 
  geom_segment(aes(x = V1,
                  y = V2,
                   xend = V1_diff,
                   yend = V2_diff,
                   color = visit_sum), alpha = 0.045,
               arrow = arrow (length = unit(0.3,"cm")))+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  scale_fill_manual(values = visit_sum_palette, labels=shape_labels)+
  scale_shape_manual(values = c(21:25),labels=shape_labels)+
  geom_point(data = means, aes(x = Mean_V1, y = Mean_V2,  fill= visit_sum, shape = visit_sum), color="black", size = 4)+
  # add arrow centroids (mean per visit sum)
  geom_segment(data = means,
               aes(x = center_v1,
                   y = center_v2,
                   xend = Mean_V1,
                   yend = Mean_V2,
                   color = visit_sum),
               arrow = arrow (length = unit(0.4,"cm")), linetype = 2)+
  xlim(-0.6,0.15)+
  ylim(-0.45, 0.4)+
  guides(color="none",shape = guide_legend(title = "Median BC dissimilarity to controls by"),fill = guide_legend(title = "Median BC dissimilarity to controls by"))+
  theme(legend.position = "bottom", legend.title.position = "top",text=element_text(size=20))+
  xlab(paste0("PCo1 (", round(percentage_explained[1], digits = 1), "%)")) +
  ylab(paste0("PCo2 (", round(percentage_explained[2], digits = 1), "%)"))

# plot "V1", "V3"
p2 <- ggplot(all_centers, aes(x = V1, y = V3)) +
  geom_point(aes(x = V1, y = V3), size = 6,color="#8176AA", shape=8) + 
  geom_segment(aes(x = V1,
                  y = V3,
                   xend = V1_diff,
                   yend = V3_diff,
                   color = visit_sum), alpha = 0.045,
               arrow = arrow (length = unit(0.3,"cm")))+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  scale_fill_manual(values = visit_sum_palette, labels=shape_labels)+
  scale_shape_manual(values = c(21:25),labels=shape_labels)+
  geom_point(data = means, aes(x = Mean_V1, y = Mean_V3,  fill= visit_sum, shape = visit_sum), color="black", size = 4)+
  # add arrow centroids (mean per visit sum)
  geom_segment(data = means,
               aes(x = center_v1,
                   y = center_v3,
                   xend = Mean_V1,
                   yend = Mean_V3,
                   color = visit_sum),
               arrow = arrow (length = unit(0.4,"cm")), linetype = 2)+
  xlim(-0.6,0.15)+
  ylim(-0.45, 0.4)+
  guides(color="none",shape = guide_legend(title = "Median BC dissimilarity to controls by"),fill = guide_legend(title = "Median BC dissimilarity to controls by"))+
  theme(legend.position = "bottom", legend.title.position = "top",text=element_text(size=20))+
  xlab(paste0("PCo1 (", round(percentage_explained[1], digits = 1), "%)"))+
  ylab(paste0("PCo3 (", round(percentage_explained[3], digits = 1), "%)"))

#plot "V2", "V3"
p3 <- ggplot(all_centers, aes(x = V2, y = V3)) +
  geom_point(aes(x = V2, y = V3), size = 6,color="#8176AA", shape=8) + 
  geom_segment(aes(x = V2,
                  y = V3,
                   xend = V2_diff,
                   yend = V3_diff,
                   color = visit_sum), alpha = 0.045,
               arrow = arrow (length = unit(0.3,"cm")))+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  scale_fill_manual(values = visit_sum_palette, labels=shape_labels)+
  scale_shape_manual(values = c(21:25),labels=shape_labels)+
  geom_point(data = means, aes(x = Mean_V2, y = Mean_V3,  fill= visit_sum, shape = visit_sum), color="black", size = 4)+
  # add arrow centroids (mean per visit sum)
  geom_segment(data = means,
               aes(x = center_v2,
                   y = center_v3,
                   xend = Mean_V2,
                   yend = Mean_V3,
                   color = visit_sum),
               arrow = arrow (length = unit(0.4,"cm")), linetype = 2)+
  xlim(-0.45,0.45)+
  ylim(-0.35, 0.45)+
  guides(color="none",shape = guide_legend(title = "Median BC dissimilarity to controls by"),fill = guide_legend(title = "Median BC dissimilarity to controls by"))+
  theme(legend.position = "bottom", legend.title.position = "top",text=element_text(size=20))+
  xlab(paste0("PCo2 (", round(percentage_explained[2], digits = 1), "%)"))+
  ylab(paste0("PCo3 (", round(percentage_explained[3], digits = 1), "%)"))
```


```{r, fig.width=16, fig.height=5}
# Display plots
ggarrange(p1,p2,p3, nrow = 1, common.legend = T, legend = "bottom")

```
```{r}
p1 <- ggplot(all_centers, aes(x = V1, y = V2)) +
  geom_point(aes(x = V1, y = V2), size = 6,color="#8176AA", shape=8) + 
  geom_point(aes(x = V1_diff,
                  y = V2_diff,
                   color = visit_sum), alpha = 0.05)+
  stat_ellipse(aes(x = V1_diff,
                  y = V2_diff,
                   color = visit_sum))+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  scale_fill_manual(values = visit_sum_palette, labels=shape_labels)+
  scale_shape_manual(values = c(21:25),labels=shape_labels)+
  geom_point(data = means, aes(x = Mean_V1, y = Mean_V2,  fill= visit_sum, shape = visit_sum), color="black", size = 4)+
  # add arrow centroids (mean per visit sum)
  geom_segment(data = means,
               aes(x = center_v1,
                   y = center_v2,
                   xend = Mean_V1,
                   yend = Mean_V2,
                   color = visit_sum),
               arrow = arrow (length = unit(0.4,"cm")), linetype = 2)+
  #xlim(-0.45,0.45)+
  #ylim(-0.35, 0.45)+
  guides(color="none",shape = guide_legend(title = "Median BC dissimilarity to controls by"),fill = guide_legend(title = "Median BC dissimilarity to controls by"))+
  theme(legend.position = "bottom", legend.title.position = "top",text=element_text(size=20))+
  xlab(paste0("PCo1 (", round(percentage_explained[1], digits = 1), "%)"))+
  ylab(paste0("PCo2 (", round(percentage_explained[2], digits = 1), "%)"))

p2 <- ggplot(all_centers, aes(x = V1, y = V3)) +
  geom_point(aes(x = V1, y = V3), size = 6,color="#8176AA", shape=8) + 
  geom_point(aes(x = V1_diff,
                  y = V3_diff,
                   color = visit_sum), alpha = 0.05)+
  stat_ellipse(aes(x = V1_diff,
                  y = V3_diff,
                   color = visit_sum))+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  scale_fill_manual(values = visit_sum_palette, labels=shape_labels)+
  scale_shape_manual(values = c(21:25),labels=shape_labels)+
  geom_point(data = means, aes(x = Mean_V1, y = Mean_V3,  fill= visit_sum, shape = visit_sum), color="black", size = 4)+
  # add arrow centroids (mean per visit sum)
  geom_segment(data = means,
               aes(x = center_v1,
                   y = center_v3,
                   xend = Mean_V1,
                   yend = Mean_V3,
                   color = visit_sum),
               arrow = arrow (length = unit(0.4,"cm")), linetype = 2)+
  #xlim(-0.45,0.45)+
  #ylim(-0.35, 0.45)+
  guides(color="none",shape = guide_legend(title = "Median BC dissimilarity to controls by"),fill = guide_legend(title = "Median BC dissimilarity to controls by"))+
  theme(legend.position = "bottom", legend.title.position = "top",text=element_text(size=20))+
  xlab(paste0("PCo1 (", round(percentage_explained[1], digits = 1), "%)"))+
  ylab(paste0("PCo3 (", round(percentage_explained[3], digits = 1), "%)"))

p3 <- ggplot(all_centers, aes(x = V2, y = V3)) +
  geom_point(aes(x = V2, y = V3), size = 6,color="#8176AA", shape=8) + 
  geom_point(aes(x = V2_diff,
                  y = V3_diff,
                   color = visit_sum), alpha = 0.05)+
  stat_ellipse(aes(x = V2_diff,
                  y = V3_diff,
                   color = visit_sum))+
  theme_classic()+
  scale_color_manual(values = visit_sum_palette)+
  scale_fill_manual(values = visit_sum_palette, labels=shape_labels)+
  scale_shape_manual(values = c(21:25),labels=shape_labels)+
  geom_point(data = means, aes(x = Mean_V2, y = Mean_V3,  fill= visit_sum, shape = visit_sum), color="black", size = 4)+
  # add arrow centroids (mean per visit sum)
  geom_segment(data = means,
               aes(x = center_v2,
                   y = center_v3,
                   xend = Mean_V2,
                   yend = Mean_V3,
                   color = visit_sum),
               arrow = arrow (length = unit(0.4,"cm")), linetype = 2)+
  xlim(-0.45,0.45)+
  ylim(-0.35, 0.5)+
  guides(color="none",shape = guide_legend(title = "Median BC dissimilarity to controls by"),fill = guide_legend(title = "Median BC dissimilarity to controls by"))+
  theme(legend.position = "bottom", legend.title.position = "top",text=element_text(size=20))+
  xlab(paste0("PCo2 (", round(percentage_explained[2], digits = 1), "%)"))+
  ylab(paste0("PCo3 (", round(percentage_explained[3], digits = 1), "%)"))
```

```{r, fig.width=16, fig.height=5}
# Display plots
ggarrange(p1,p2,p3, nrow = 1, common.legend = T, legend = "bottom")

```

